---
title: "Beyond the Manual"
lang: en
output:
  rmarkdown::html_vignette:
    css: tda-manual.css
    highlight: null
    toc: true
    toc_depth: 2
    includes:
      in_header: theme-toggle.html
pkgdown:
  as_is: true
vignette: >
  %\VignetteIndexEntry{Beyond the Manual}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
source("tda-hooks.R")
library(tdaR)
# the North Carolina shapefile belongs to the sf package: the shapefile
# boxes run when sf is installed and are skipped otherwise
has_sf <- requireNamespace("sf", quietly = TRUE)
```

The *TDA User's Manual* stops at chapter 8, and development did not. The
spatial commands, the file readers for formats that postdate the manual,
and TDA's online help are all in the program and in none of its
documentation. This is what is there.

## What happens inside

There is no reimplementation. TDA's C, context-threaded and compiled
into the package, runs unchanged; a wrapper writes the command file
TDA would have been given, hands it over in memory, and reads the
answer back. Every object therefore carries the run it came from.

```{r inside, box = "One fit, and the run underneath it"}
d <- tda_rrdat()
fit <- tda_rate(Surv(TFP, DES) ~ EDU, d)
cat(fit$run$commands, sep = "\n")
```

The answer arrives on two channels. The console is what TDA printed --
`tda_output()` shows all of it, and `tda_diagnostics()` the lines
that are errors, warnings or notes, which TDA otherwise leaves in the
middle of its output for the reader to find. The exports are the doubles
behind the numbers TDA printed: coefficients, covariance matrices,
tables, and a tap on every number the console formatter was handed. A
wrapper takes its numbers from the exports and uses the text only to
find its way around, so `coef()` holds the double TDA computed, not a
number re-read from a print format.

```{r inside2, box = "Printed, and computed"}
grep("EDU", fit$run$output, value = TRUE)
sprintf("%.17g", coef(fit)[["EDU"]])
names(fit$run$exports)
```

`tda_output_audit()` is the check that this holds: it classifies every
line TDA printed as boilerplate, an echo of the input, a setting, a
diagnostic, or a number that is in the object or in the exports, and
reports whatever is none of those. The package's tests run it over
every example in this manual and over the 634 command files of TDA's
own regression suite.

```{r inside3, box = "Where every line went"}
attr(tda_output_audit(fit), "summary")
```

## Spatial data

TDA holds points, lines and polygons in a spatial structure of its own,
and has about twenty commands over it. None of them appears in the
manual.

`tda_spatial()` builds the structure from a data frame. `type = 1` is
points; `type = 3` groups consecutive rows sharing an `id` into one
polygon each.

```{r sd1, box = "A spatial structure, and what is in it"}
set.seed(7)
pts <- data.frame(id = 1:12, x = round(runif(12), 3), y = round(runif(12), 3))
s <- tda_spatial(pts)
tda_sd_info(s)
```

`tda_sd_voronoi()` is the Voronoi diagram. What `sdvd` writes is itself a
spatial data file -- points for the vertices, lines for the edges -- so
the result plots with the same machinery the input does.

```{r sd2, box = "The Voronoi diagram of those points", fig.height = 4.2}
v <- tda_sd_voronoi(s, opt = 2)
plot(v, xlim = c(-0.2, 1.2), ylim = c(-0.2, 1.2))
```

`tda_sd_select()` keeps what falls inside a rectangle and
`tda_sd_clip()` cuts to it. Both want the rectangle given; TDA will not
guess one.

```{r sd3, box = "Selecting and clipping to a region"}
tda_sd_select(s, rec = "0,0,0.5,0.5")$table
tda_sd_clip(s, c(0, 0, 0.5, 0.5))$table
```

Polygons are the same structure with `type = 3`. `tda_sd_data()` gives
the vertices back as an ordinary data frame, which is the form most
things in R want.

```{r sd4, box = "Two square polygons"}
poly <- data.frame(id = c(1, 1, 1, 1, 2, 2, 2, 2),
                   x  = c(0, 2, 2, 0, 3, 5, 5, 3),
                   y  = c(0, 0, 2, 2, 0, 0, 2, 2))
sp <- tda_spatial(poly, type = 3)
tda_sd_data(sp)$table
```

## Reading a shapefile

`tda_read_shapefile()` reads geometry and attaches the `.dbf` attribute
table as `$attributes`. The manual documents SPSS and Stata files and
nothing else; shapefiles, E00, DCW, GSHHS, `.dbf` and BIFF `.xls` are all
readable and all undocumented. The file is the North Carolina SIDS
shapefile that the sf package ships, read here by TDA itself; these
boxes run when sf is installed.

```{r shp, box = "North Carolina counties, read through TDA", eval = has_sf}
nc <- tda_read_shapefile(system.file("shape/nc.shp", package = "sf"))
tda_sd_info(nc)
head(nc$attributes[, c("AREA", "PERIMETER", "NAME", "FIPS")])
```

The geometry is TDA's, so TDA's plotting draws it.

```{r shpplot, box = "The same counties, drawn by TDA", eval = has_sf, fig.height = 3}
plot(nc)
```

`tda_sd_enclosing()` gives the bounding box or the convex hull of
everything in the structure, and `tda_sd_data()` the vertices.

```{r shp2, box = "The bounding box, and the vertices", eval = has_sf}
tda_sd_enclosing(nc)$table
head(tda_sd_data(nc)$table)
```

Some of these commands compare one object type against another in the
same structure. A set of polygons on its own has no lines or points for
them to find, and saying so is the right answer rather than an error.

```{r shp3, box = "Nothing to report, correctly", eval = has_sf}
tda_sd_neighbours(nc)$text
```

## TDA's help

The program carries its help text for every command, and
`tda_help()` reads it. This is the fastest way to find what a command's
options are actually called, since the R wrappers deliberately do not
expose all of them by name.

```{r help, box = "tda_help"}
cat(head(tda_help("ploth"), 20), sep = "\n")
```

## Running TDA directly

`tda_run()` takes command-file text and runs it, which is the escape
hatch for anything without a wrapper. `tda_block()` builds one command
from named arguments so the syntax does not have to be written by hand.

```{r run, box = "A command file, from R"}
d <- data.frame(X = c(1, 2, 3, 4), Y = c(2, 4, 6, 9))
res <- tda_run(c(tda_nvar(d), "dstat;"), data = d)
cat(grep("^ *(Variable|X|Y)", res$output, value = TRUE), sep = "\n")
```

Every wrapper returns its `$run`, so the command file a call
generated can always be read back. This is worth knowing when a result
looks wrong: the question is usually what TDA was asked, not what it
computed.

```{r run2, box = "What a wrapper actually sent"}
fit <- tda_lsreg(Y ~ X, d)
cat(fit$run$commands, sep = "\n")
```

## The expression language

TDA's operators are available without a data set through
`tda_evalf()`, and its distribution functions are the ones the manual
lists in 5.2.5 -- `nd`, `ndf`, `td`, `chd`, `fd`, `mr`, `poisson`,
`negbin`. They agree with R's, which is checked by a differential fuzzer
on every run of the test suite, so either side gives the same numbers.

```{r expr, box = "The same quantity, twice"}
sprintf("%.17g", c(tda_evalf("nd(x)", options = list(x = 1.96))$value,
                   pnorm(1.96)))
sprintf("%.17g", c(tda_evalf("mr(x)", options = list(x = 0.5))$value,
                   dnorm(0.5) / (1 - pnorm(0.5))))
```

## Character data

TDA holds fixed-width string variables, and a character or factor column
reaches it: `tda_nvar()` declares it as `str(n,m)` and `tda_strings()`
brings it back. The four string operators -- `strlen`, `strsp`, `strv`,
`strvp` -- work on it.

```{r str, box = "Strings through TDA"}
d2 <- data.frame(ID = 1:3, Name = c("Ashe", "Alleghany", "Surry"),
                 Area = c(0.114, 0.061, 0.143))
res <- tda_run(c(tda_nvar(d2), "pdata() = out.txt;"), data = d2)
tda_strings(res)
```

## Graphs

Chapter 7 of the manual covers relational data, and the 31 `tda_g_*`
commands implement it: components, cliques, cut points, spanning trees,
cycles, flows, shortest paths, and the social-network measures of 7.6.
They are documented but easy to miss, so one example, on a small graph.

```{r graph, box = "A graph, and some of what TDA asks of it"}
e <- data.frame(from = c(1, 1, 2, 3, 4, 5, 5),
                to   = c(2, 3, 3, 4, 5, 6, 7))
g <- tda_graph(e, directed = FALSE)
tda_g_degrees(g)
tda_g_components(g)
tda_g_cutpoints(g)
```

```{r graphplot, box = "The same graph, drawn by TDA", fig.height = 3.4}
plot(g)
```
