---
title: "Techniques of Event History Modeling, in R"
lang: en
output:
  rmarkdown::html_vignette:
    css: tda-manual.css
    highlight: null
    toc: true
    toc_depth: 2
    includes:
      in_header: theme-toggle.html
pkgdown:
  as_is: true
vignette: >
  %\VignetteIndexEntry{Techniques of Event History Modeling, in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
source("tda-hooks.R")
library(tdaR)
```

The command files distributed with Blossfeld and Rohwer, *Techniques of
Event History Modeling*, are the second body of TDA examples, and the one
most people met the program through. They live in `examples/ehhnew`,
about ninety of them, in eight families:

| prefix | book chapter | what it covers |
|---|---|---|
| `ehb` | 2 | episode data |
| `ehc` | 3 | life tables and product-limit estimation |
| `ehd` | 4 | exponential models |
| `ehe` | 5 | piecewise constant models |
| `ehf` | 6 | episode splitting and time-varying covariates |
| `ehg` | 7 | parametric models |
| `ehh` | 8 | graphical tests of distributional assumptions |
| `ehi` | 9 | Cox models |

The suffixes are worth knowing, since they are not what they look like:
a file ending in `a` plots the fitted rate, while `s` and `p` are the
second and third steps of the pseudoresidual check of the book's section
8.2 -- `s` estimates a product-limit survivor function for a model's
residuals, `p` plots `log S` against them. They are separate command
files in TDA only because each step writes its result to disk for the
next to read; here they follow the fit directly.

This vignette walks the families in order, naming the command file each
box corresponds to and, in brackets, the page and box number in the book
where the same output is printed, so the two can be read side by side.
`tests/vignette-qa/vignette-qa-book.csv` in the sources is the other direction: one row
per box and figure in the book, with the vignette box that reproduces it. The data are the ones of the manual's section 3.3.3,
and the covariate set is the book's own.

```{r data, box = "The data, and the book's covariates"}
d  <- tda_rrdat()            # rrdat.1, with the derived variables
d4 <- tda_rrdat(states = 4)  # job changes split by prestige

XVARS <- c("EDU", "COHO2", "COHO3", "LFX", "PNOJ", "PRES")
with_x <- function(lhs) stats::reformulate(XVARS, response = lhs)
```

Most plots below share one coordinate system, taken from the command
files' own `psetup` rather than from the range of the data, so that a
picture can be laid beside the book.

## 2. Event History Data Structures

*Book chapter 2, pp. 41-55. Command files `ehb1` to `ehb3`.*

**ehb1** is the data before any episode definition, and **ehb2** the
episode data itself: one transition, from 0 to 1.

```{r ehb, box = "ehb1, ehb2: the data and the episodes [book pp. 43-52, Boxes 2.2.1-2.2.5]"}
# the twelve variables of rrdat.1; DES and TFP are derived by ehb1.cf
str(d[, 1:12])
head(d[, 1:12], 9)
head(d[, c("ID", "NOJ", "TStart", "TFin", "SEX", "TFP", "DES")], 4)
tda_rate(Surv(TFP, DES) ~ 1, d)$episodes
```

**ehb3** is the multi-episode form of the same data (Box 2.2.6): the
spells of a person are chained, the origin state of spell k is
`1 + 2(k - 1)`, the destination the next number up if the job ended,
and time runs from the person's first job. `id=` and `spell=` are
`edef(id=, sn=)`; the episode table is the book's Box 2.2.8.

```{r ehb3, box = "ehb3: multi-episode data [book pp. 53-55, Boxes 2.2.6-2.2.8]"}
dm <- d
dm$ORG <- 1 + 2 * (dm$NOJ - 1)
dm$DESM <- ifelse(dm$TFin < dm$TI, dm$ORG + 1, dm$ORG)
dm$TFC <- ave(dm$TStart, dm$ID, FUN = function(x) x[1])
dm$TSP <- dm$TStart - dm$TFC
dm$TFPM <- dm$TFin - dm$TFC + 1
# the episodes of the first three individuals, as edef() will see them
head(dm[dm$ID <= 3, c("ID", "NOJ", "TStart", "TFin", "SEX", "ORG", "DESM",
                      "TSP", "TFPM")], 9)
eb3 <- suppressWarnings(
    tda_rate(Surv(TSP, TFPM, ORG, DESM) ~ 1, dm, id = "ID", spell = "NOJ"))
eb3$episodes
attr(eb3$episodes, "sums")   # the totals under each spell; sn = NA is all spells
```

The ninth spell holds a single, censored episode -- no transition to
estimate -- and TDA marks it with `*` in the `Excl` column, which is
`excluded = TRUE` here; the warning that says so is the one suppressed
above.

## 3. Descriptive Methods

*Book chapter 3, pp. 57-85. Life tables and the product-limit estimator; command files `ehc1` to `ehc10`.*

**ehc1** is the life table over 30-month intervals, **ehc3** the same by
sex.

```{r ehc1, box = "ehc1, ehc3: life tables [book pp. 60-66, Boxes 3.1.1-3.1.4]"}
lt <- tda_ltb(Surv(TFP, DES) ~ 1, d, tp = seq(0, 500, by = 30))
summary(lt, n = nrow(lt$tables[[1]]))

lt2 <- tda_ltb(Surv(TFP, DES) ~ as.factor(SEX), d, tp = seq(0, 500, by = 30))
lt2$episodes   # the episodes behind both groups
lt2$summary
```

**ehc2** plots the survivor function the table implies, and **ehc2a** and
**ehc2b** the density and the rate, each on its own scale.

```{r ehc2, box = "ehc2: the life table survivor function [book p. 65, Fig. 3.1.1]", class.source = "folded"}
sv <- tda_survivor(lt)
p <- tda_ps(sv, width = 80, height = 40, xlim = c(0, 300), ylim = c(0, 1))
p <- tda_pl(p, "plxa", sc = 60, ic = 5)
p <- tda_pl(p, "plya", sc = 1, ic = 10)
p <- tda_pl_lines(tda_pl(p, "plframe"), "time", "survivor")
plot(tda_pl_text(p, "Life Table Survivor Function", at = c(80, 0.8)))
```

```{r ehc2a, box = "ehc2a, ehc2b: the density and the rate [book p. 65, Fig. 3.1.1]", class.source = "folded"}
p <- tda_ps(sv, width = 80, height = 40, xlim = c(0, 300), ylim = c(0, 0.02))
p <- tda_pl(p, "plxa", sc = 60, ic = 5)
p <- tda_pl(p, "plya", sc = 0.01, ic = 10)
p <- tda_pl_lines(tda_pl(p, "plframe"), "time", "density")
plot(tda_pl_text(p, "Life Table Density Function", at = c(80, 0.017)))

p <- tda_ps(sv, width = 80, height = 40, xlim = c(0, 300), ylim = c(0, 0.02))
p <- tda_pl(p, "plxa", sc = 60, ic = 5)
p <- tda_pl(p, "plya", sc = 0.01, ic = 10)
p <- tda_pl_lines(tda_pl(p, "plframe"), "time", "rate")
plot(tda_pl_text(p, "Life Table Transition Rate", at = c(80, 0.017)))
```

**ehc4** draws one curve per sex.

```{r ehc4, box = "ehc4: survivor functions by sex [book p. 67, Fig. 3.1.2]", class.source = "folded"}
sv2 <- tda_survivor(lt2)
p <- tda_ps(sv2, width = 80, height = 40, xlim = c(0, 300), ylim = c(0, 1))
p <- tda_pl(p, "plxa", sc = 60, ic = 5)
p <- tda_pl(p, "plya", sc = 1, ic = 10)
p <- tda_pl_lines(p, "time", "survivor", by = "group")
p <- tda_pl_text(p, "Men",   at = c(240, 0.25))
p <- tda_pl_text(p, "Women", at = c(240, 0.09))
plot(tda_pl_text(p, "Life Table Survivor Function", at = c(80, 0.8)))
```

**ehc5** is the product-limit estimate, **ehc7** the comparison of the two
curves, and **ehc9** the four destination states.

```{r ehc5, box = "ehc5, ehc7, ehc9: product-limit estimation [book pp. 74-83, Boxes 3.2.1-3.2.2, 3.3.1-3.3.4a, 3.3.4b]"}
pl <- tda_ple(Surv(TFP, DES) ~ 1, d)
pl$episodes          # the episode table TDA prints above the estimate
print(pl, n = 10)

pl2 <- tda_ple(Surv(TFP, DES) ~ as.factor(SEX), d, compare = TRUE)
pl2$comparison

# one product-limit table per destination: upward, then lateral moves
pl4 <- tda_ple(Surv(TFP, DES) ~ 1, d4)
pl4$episodes
pl4$summary
for (b in pl4$blocks[1:2]) print(utils::head(b, 10), row.names = FALSE)
```

**ehc6** plots the estimate, **ehc10** one curve per destination state,
and **ehc8** adds the confidence bands, cut at 290 months where they
widen past usefulness.

```{r ehc6, box = "ehc6, ehc10: the estimated survivor functions [book pp. 76-85, Box 3.2.3; Fig. 3.2.1, Fig. 3.2.3]", class.source = "folded"}
p <- tda_ps(tda_survivor(pl), width = 80, height = 40, xlim = c(0, 300), ylim = c(0, 1))
p <- tda_pl(p, "plxa", sc = 60, ic = 5)
p <- tda_pl(p, "plya", sc = 1, ic = 10)
p <- tda_pl_lines(p, "time", "survivor")
plot(tda_pl_text(p, "Product-Limit Survivor Function", at = c(60, 0.8)))

s4 <- tda_survivor(pl4)
p <- tda_ps(s4, width = 80, height = 40, xlim = c(0, 300), ylim = c(0, 1))
p <- tda_pl(p, "plxa", sc = 60, ic = 5)
p <- tda_pl(p, "plya", sc = 1, ic = 10)
for (i in seq_along(g4 <- c("0,1", "0,2", "0,3")))
    p <- tda_pl_lines(p, "time", "survivor", by = "group",
                      select = g4[i], lty = c(1, 5, 8)[i])
p <- tda_pl_text(p, "upward moves",   at = c(200, 0.75))
p <- tda_pl_text(p, "lateral moves",  at = c(200, 0.3))
plot(tda_pl_text(p, "downward moves", at = c(200, 0.55)))
```

A band is two fills, the upper curve to the axis in grey and the lower
curve to the axis in white on top of it; `ehc8.cf` draws the axes
first, so the white fill lies over half the width of the x axis line
and the axis looks thinner under the bands than beyond them. The axes
go last here.

```{r ehc8, box = "ehc8: with confidence bands [book p. 78, Fig. 3.3.1]", class.source = "folded"}
sb <- tda_survivor(pl2)
sb <- sb[sb$time < 290, ]
p <- tda_ps(sb, width = 80, height = 40, xlim = c(0, 300), ylim = c(0, 1))
p <- tda_pl_lines(p, "time", "survivor", by = "group", select = "1",
                  band = c("upper", "lower"), lty = 1)
p <- tda_pl_lines(p, "time", "survivor", by = "group", select = "2",
                  band = c("upper", "lower"), lty = 5)
p <- tda_pl(p, "plxa", sc = 60, ic = 5)
p <- tda_pl(p, "plya", sc = 1, ic = 10)
p <- tda_pl_text(p, "Men", at = c(240, 0.30))
p <- tda_pl_text(p, "Women", at = c(50, 0.10))
plot(tda_pl_text(p, "Survivor Functions for Men and Women", at = c(60, 0.8)))
```

## 4. Exponential Transition Rate Models

*Book chapter 4, pp. 87-118. Command files `ehd1` to `ehd7`.*

**ehd1** is the null model and **ehd2** the same with the book's
covariates.

```{r ehd12, box = "ehd1, ehd2: exponential models [book pp. 91-98, Boxes 4.1.1-4.1.5]"}
ehd1 <- tda_rate(Surv(TFP, DES) ~ 1, d, prate = "0(10)300")
ehd1
ehd1$episodes   # the episodes TDA built, before the estimates
tda_rates(ehd1)

summary(tda_rate(with_x("Surv(TFP, DES)"), d))
```

The rate this model implies is a constant, `exp(-4.4891)` = 0.0112, and
the book prints it as the table `ehd1.prs` above rather than drawing it.
Everything that follows is an attempt to let that constant vary.

**ehd3** estimates the same specification over three destination states.

```{r ehd3, box = "ehd3: three destinations [book pp. 102-104, Boxes 4.2.1-4.2.2]"}
ehd3 <- tda_rate(with_x("Surv(TFP, DES)"), d4)
ehd3$episodes
ehd3
```

**ehd4** and **ehd5** take random subsamples so that the destinations are
comparably sized. `tda_runif()` is TDA's generator, a multiplicative
congruential one with a fixed seed, so the episodes drawn here are the
episodes TDA drew; `runif()` would give a different subsample and so
different estimates.

```{r ehd45, box = "ehd4, ehd5: subsamples, one transition at a time [book pp. 106-110, Boxes 4.2.3-4.2.6]"}
sub <- d4[tda_runif(nrow(d4)) <= 84 / 219, ]
sub$DES <- ifelse(sub$DES == 2, 2L, 0L)
cat(sprintf("%d of %d episodes selected\n", nrow(sub), nrow(d4)))
ehd4 <- tda_rate(with_x("Surv(TFP, DES)"), sub)
ehd4$episodes
ehd4

d5 <- d4[tda_runif(nrow(d4)) <= 84 / 155, ]
d5$DES <- ifelse(d5$DES == 3, 3L, 0L)
cat(sprintf("%d of %d episodes selected\n", nrow(d5), nrow(d4)))
ehd5 <- tda_rate(with_x("Surv(TFP, DES)"), d5)
ehd5$episodes
ehd5
```

Box 4.2.7 of the book lays the two sets of estimates side by side, which
is the point of the exercise: with the events standardized, `COHO3` and
`LFX` lose their significance for lateral moves and `COHO3` for downward
ones, so the same data can support different substantive conclusions
depending on how many events each transition happens to have. Upward
moves are the baseline and appear in both columns unchanged.

```{r ehd427, box = "Box 4.2.7: standardized against unstandardized estimates [book p. 111, Box 4.2.7]"}
unstd <- function(des) {
    e <- ehd3$estimates
    e[e$Des == des, c("Variable", "Coeff", "C/Error")]
}
std <- list(`1` = unstd(1),
            `2` = ehd4$estimates[, c("Variable", "Coeff", "C/Error")],
            `3` = ehd5$estimates[, c("Variable", "Coeff", "C/Error")])

for (des in 1:3) {
    a <- unstd(des)
    b <- std[[as.character(des)]]
    cmp <- data.frame(Variable = a$Variable,
                      Coeff    = a$Coeff,   `C/Error`   = a$`C/Error`,
                      Coeff.s  = b$Coeff,   `C/Error.s` = b$`C/Error`,
                      check.names = FALSE)
    cat(sprintf("\n%s moves. Events: %d (unstandardized) %d (standardized)\n",
                c("Upward", "Lateral", "Downward")[des],
                ehd3$episodes$episodes[ehd3$episodes$des == des],
                if (des == 1) ehd3$episodes$episodes[ehd3$episodes$des == 1]
                else list(NULL, ehd4, ehd5)[[des]]$episodes$episodes[2]))
    print(format(cmp, digits = 4), row.names = FALSE)
}
```

**ehd6** holds the cohort effects equal across the three transitions,
which is what a constraint is for.

```{r ehd6, box = "ehd6: cohort effects constrained equal [book pp. 112-113, Boxes 4.2.8-4.2.9]"}
ehd6 <- tda_rate(with_x("Surv(TFP, DES)"), d4,
                 constraints = c("b3 - b10 = 0", "b3 - b17 = 0",
                                 "b4 - b11 = 0", "b4 - b18 = 0"))
ehd6
ehd6$episodes   # the episodes TDA built, as it reports them
```

**ehd7** keeps the first four spells and gives each its coefficients:
the effect of education on leaving a first job need not be its effect on
leaving a fourth. `id=` and `spell=` are TDA's `edef(id=, sn=)`, and
`spell_terms=` gives one spell a covariate set of its own -- the first
job has no labour force experience to speak of, so `LFX` is left out
of it, as `ehd7.cf` does.

```{r ehd7, box = "ehd7: a multi-episode model [book pp. 115-118, Boxes 4.3.1-4.3.2a, 4.3.2b]"}
d7 <- d[d$NOJ <= 4, ]
d7 <- d7[order(d7$ID, d7$NOJ), ]

# ehd7.cf measures time from the start of the FIRST job, not from the
# start of each spell: TFC is the first spell's starting time
# carried forward, and the episode runs from TStart - TFC to
# TFin - TFC + 1. Using each spell's duration instead would start
# every spell at 0 and give a different episode structure.
d7$TFC <- ave(ifelse(d7$NOJ == 1, d7$TStart, NA), d7$ID,
              FUN = function(v) v[!is.na(v)][1])
d7$TSP <- d7$TStart - d7$TFC
d7$TFPC <- d7$TFin - d7$TFC + 1

ehd7 <- tda_rate(Surv(TSP, TFPC, DES) ~ EDU + COHO2 + COHO3 + LFX + PRES, d7,
                 id = "ID", spell = "NOJ",
                 spell_terms = list("1" = c("EDU", "COHO2", "COHO3", "PRES")))
ehd7$episodes
ehd7
```

## 5. Piecewise Constant Exponential Models

*Book chapter 5, pp. 119-130. Command files `ehe1` to `ehe3`.*

The rate is constant within periods and free between them, which buys
duration dependence without committing to a shape.

```{r pc, box = "The piecewise constant helper"}
pc <- function(f, data, model = "exponential_periods", tp = "0 (12) 96", ...)
    tda_rate(f, data, model = model, options = list(tp = tp), ...)
```

**ehe1** is the null model, **ehe2** the same with covariates.

```{r ehe12, box = "ehe1, ehe2: piecewise constant exponential [book pp. 122-127, Boxes 5.2.1-5.2.2, 5.3.1-5.3.2]"}
ehe1 <- pc(Surv(TFP, DES) ~ 1, d, prate = "0(1)300")
ehe1
ehe1$episodes      # the episodes TDA built
ehe1$periods       # the period boundaries and the events in each
ehe2 <- pc(with_x("Surv(TFP, DES)"), d)
ehe2
ehe2$periods
```

**ehe1a** plots the rate. The steps are the periods. (`ehe2p` is not a
second rate plot but the pseudoresidual check of `ehe2`, which comes
below with the rest of section 8.2.)

```{r ehe1a, box = "ehe1a: the estimated rate [book pp. 122-124, Boxes 5.2.1-5.2.2; Fig. 5.2.1]", class.source = "folded"}
er <- tda_rates(ehe1)
p <- tda_ps(er, width = 80, height = 40, xlim = c(0, 300), ylim = c(0, 0.03))
p <- tda_pl(p, "plxa", sc = 60, ic = 5)
p <- tda_pl(p, "plya", sc = 0.01, ic = 2)
p <- tda_pl_lines(p, "time", "rate")
plot(tda_pl_text(p, "Piecewise Constant Exponential Rate", at = c(60, 0.025)))
```

**ehe3** uses three periods chosen by hand rather than a regular grid,
which is model 16 rather than model 3.

```{r ehe3, box = "ehe3: three periods [book pp. 129-130, Boxes 5.4.1-5.4.2]"}
ehe3 <- pc(with_x("Surv(TFP, DES)"), d,
           model = "exponential_periods2", tp = "0,24,60")
ehe3

# the two tables TDA prints before the estimates: the episodes it built,
# and how they fall across the periods
ehe3$episodes
ehe3$periods
```

## 6. Exponential Models with Time-dependent Covariates

*Book chapter 6, pp. 131-157. Episode splitting; command files `ehf1` to `ehf6`.*

**ehf1** splits every episode at the date of marriage and writes the
pieces out (`rrdat.s1`, Box 6.4.2); **ehf5** splits on a fixed grid of
60 months (`rrdat.d60`, Box 6.5.2). `tda_split()` is both, `at=` for a
per-episode split time and `grid=` for the fixed one.

```{r ehf15, box = "ehf1, ehf5: the split episode data [book pp. 143-154, Boxes 6.4.1-6.4.2, 6.5.1-6.5.2]"}
ds <- d
ds$MarrDate <- ifelse(ds$TMAR <= 0, 10000, ds$TMAR - ds$TStart)
sp <- tda_split(stats::reformulate(c(XVARS, "SEX"),
                                   response = "Surv(TFP, DES)"),
                ds, at = "MarrDate")
cat(sprintf("%d episodes split into %d\n", nrow(ds), nrow(sp)))
attr(sp, "episodes")   # the episodes epdat reports before writing them
head(sp, 6)
sp6 <- tda_split(with_x("Surv(TFP, DES)"), d, grid = seq(60, 420, by = 60))
cat(sprintf("%d episodes split into %d on the 60-month grid\n",
            nrow(d), nrow(sp6)))
attr(sp6, "episodes")
head(sp6, 6)
```

A covariate that changes during an episode is handled by cutting the
episode where it changes. Marriage is the example: `MarrDate` is the
month it happens, or past the end of the episode if it never does.
**ehf2**, **ehf3**, **ehf4** and **ehf6** are the models on the pieces.

```{r ehfsplit, box = "ehf2, ehf3: episode splitting at marriage [book pp. 145-149, Boxes 6.4.3-6.4.6]"}
sp$MARR <- as.integer(sp$MarrDate <= sp$ts)
cat(sprintf("%d episodes split into %d pieces\n", nrow(ds), nrow(sp)))

# Surv(ts, tf, des), not Surv(tf - ts, des): a piece that begins at ts is
# left truncated, and a model that is told otherwise puts it into risk
# sets it was never in. An exponential rate only ever sees the difference
# and so is unharmed, but the Cox models of ehi3 to ehi5 below are not.
ehf2 <- tda_rate(Surv(ts, tf, des) ~ MARR, sp)
ehf2$episodes
ehf2
ehf3 <- tda_rate(stats::reformulate(c("MARR", XVARS),
                                 response = "Surv(ts, tf, des)"), sp)
ehf3$episodes
ehf3
```

**ehf4** distinguishes married men from married women.

```{r ehf4, box = "ehf4: marriage, by sex [book pp. 150-151, Boxes 6.4.7-6.4.8]"}
sp$MarrMen <- as.integer(sp$SEX == 1 & sp$MarrDate <= sp$ts)
ehf4 <- tda_rate(stats::reformulate(c(XVARS, "MARR", "MarrMen"),
                                    response = "Surv(ts, tf, des)"), sp)
ehf4
ehf4$episodes
```

**ehf6** is the model on the episodes **ehf5** split on a fixed grid -- at
60, 120, ... 420 months -- so that a covariate measured in calendar time
can change within a job.
`grid=` takes those times; `at=` is for a split time that differs per
episode, as marriage does. `LFX60` is the sub-episode's starting
time, which is labour force experience accumulated within the job.

```{r ehf6, box = "ehf6: splitting on a fixed grid [book pp. 155-156, Boxes 6.5.3-6.5.4]"}
sp6$LFX60 <- sp6$ts
cat(sprintf("%d episodes split into %d pieces\n", nrow(d), nrow(sp6)))
ehf6 <- tda_rate(stats::reformulate(c(XVARS, "LFX60"),
                                    response = "Surv(ts, tf, des)"), sp6)
ehf6
ehf6$episodes   # the split episodes, as TDA reports them
```

Box 6.5.5 of the book does the same with finer grids -- 24 and 12
months -- and shows how little the estimates move with the grid.

```{r ehf6b, box = "Box 6.5.5: the grid at 60, 24 and 12 months [book p. 157, Box 6.5.5]"}
grid_fit <- function(step) {
    spg <- tda_split(with_x("Surv(TFP, DES)"), d, grid = seq(step, 420, by = step))
    spg$LFXG <- spg$ts
    fit <- tda_rate(stats::reformulate(c(XVARS, "LFXG"),
                                       response = "Surv(ts, tf, des)"), spg)
    e <- fit$estimates
    stats::setNames(data.frame(e$Coeff, e$`C/Error`), c("Coeff", "T-Stat"))
}
# the three grids side by side, coefficient and t statistic for each
g <- lapply(c(60, 24, 12), grid_fit)
out <- do.call(cbind, g)
names(out) <- paste(rep(c("60", "24", "12"), each = 2), names(out))
rownames(out) <- c("Const", XVARS, "LFXG")
print(format(out, digits = 5))
```

## 7. Parametric Models of Time-dependence

*Book chapter 7, pp. 159-212. Gompertz, Weibull, log-logistic, log-normal and sickle; command files `ehg1` to `ehg13`.*

Five families, each fitted without covariates and then with them. `on=`
selects which parameter of the distribution the covariates attach to —
`xa`, `xb` or `xc` — which is the difference between the `ehg2`/`ehg3`/
`ehg4` group of files.

```{r ehgfit, box = "The two fits each family gets"}
fit1 <- function(model, ...)
    tda_rate(Surv(TFP, DES) ~ 1, d, model = model, prate = "0(1)300", ...)
fitx <- function(model, on = "xa", ...)
    tda_rate(with_x("Surv(TFP, DES)"), d, model = model, on = on, ...)
```

**ehg1**, **ehg5**, **ehg8**, **ehg10** and **ehg12** are the null
models: Gompertz, Weibull, log-logistic, log-normal and sickle.

```{r ehgnull, box = "ehg1, ehg5, ehg8, ehg10, ehg12: five families [book pp. 183-210, Boxes 7.2.1-7.2.2, 7.3.1-7.3.2, 7.4.1-7.4.2, 7.5.1-7.5.2, 7.6.1-7.6.2]"}
models <- c("gompertz", "weibull", "loglogistic", "lognormal", "sickle")
fits <- lapply(models, fit1)
names(fits) <- models
for (m in models) print(fits[[m]])

# all five are fitted to the same episodes, which TDA reports before
# each set of estimates
fits$gompertz$episodes
```

**ehg1a**, **ehg5a** and the rest plot each fitted rate, which is where
the families differ from each other and from the exponential. All five
command files set the same axes -- 0 to 300 months against a rate of 0
to 0.02 -- so the shapes can be compared directly. The book prints them
one to a section, beside the model they belong to: Figures 7.2.2, 7.3.2,
7.4.2, 7.5.2 and 7.6.2, in that order. The exponential's constant
`r sprintf("%.4f", exp(coef(ehd1)[[1L]]))` would be a horizontal line
across the middle of each.

```{r ehgplots, box = "ehg1a, ehg5a, ehg8a, ehg10a, ehg12a: the five fitted rates [book pp. 185-211, Fig. 7.2.2, Fig. 7.3.2, Fig. 7.4.2, Fig. 7.5.2, Fig. 7.6.2]", class.source = "folded"}
p <- tda_ps(tda_rates(fits$gompertz), file = "ehg1a.ps", width = 80, height = 40,
            xlim = c(0, 300), ylim = c(0, 0.02))
p <- tda_pl(p, "plxa", sc = 60, ic = 5)
p <- tda_pl(p, "plya", sc = 0.01, ic = 5)
plot(tda_pl_lines(p, "time", "rate"))

p <- tda_ps(tda_rates(fits$weibull), file = "ehg5a.ps", width = 80, height = 40,
            xlim = c(0, 300), ylim = c(0, 0.02))
p <- tda_pl(p, "plxa", sc = 60, ic = 5)
p <- tda_pl(p, "plya", sc = 0.01, ic = 5)
plot(tda_pl_lines(p, "time", "rate"))

p <- tda_ps(tda_rates(fits$loglogistic), file = "ehg8a.ps", width = 80, height = 40,
            xlim = c(0, 300), ylim = c(0, 0.02))
p <- tda_pl(p, "plxa", sc = 60, ic = 5)
p <- tda_pl(p, "plya", sc = 0.01, ic = 5)
plot(tda_pl_lines(p, "time", "rate"))

p <- tda_ps(tda_rates(fits$lognormal), file = "ehg10a.ps", width = 80, height = 40,
            xlim = c(0, 300), ylim = c(0, 0.02))
p <- tda_pl(p, "plxa", sc = 60, ic = 5)
p <- tda_pl(p, "plya", sc = 0.01, ic = 5)
plot(tda_pl_lines(p, "time", "rate"))

p <- tda_ps(tda_rates(fits$sickle), file = "ehg12a.ps", width = 80, height = 40,
            xlim = c(0, 300), ylim = c(0, 0.02))
p <- tda_pl(p, "plxa", sc = 60, ic = 5)
p <- tda_pl(p, "plya", sc = 0.01, ic = 5)
plot(tda_pl_lines(p, "time", "rate"))
```

**ehg2**, **ehg6**, **ehg9**, **ehg11** and **ehg13** add the covariates.
In the Gompertz-Makeham model of `ehg2` they go on `xb`, the term the
rate's growth depends on, not on the Makeham constant `xa`; the other
four families put them on `xa`.

```{r ehgcov, box = "ehg2, ehg6, ehg9, ehg11, ehg13: with covariates [book pp. 186-212, Boxes 7.2.3-7.2.4, 7.3.3-7.3.4, 7.4.3-7.4.4, 7.5.3-7.5.4, 7.6.3-7.6.4]"}
sickle13 <- fitx("sickle")
sickle13$episodes   # the same episodes for all five, as TDA reports

fitx("gompertz", on = "xb", prate = "0(10)300")
fitx("weibull", prate = "0(10)300")
fitx("loglogistic")
fitx("lognormal")
fitx("sickle")
```

**ehg3** and **ehg4** put the same covariates on the other parameters as
well: `xb` and `xc`, then all three. **ehg7** does the same for the
Weibull, on `xa` and `xb`.

```{r ehgon, box = "ehg3, ehg4, ehg7: covariates on more than one term [book pp. 188-197, Boxes 7.2.5-7.2.8, 7.3.5-7.3.6]"}
ehg4 <- fitx("gompertz", on = c("xa", "xb", "xc"))
ehg4$episodes   # the same episodes for every model in this group

fitx("gompertz", on = c("xb", "xc"))
fitx("gompertz", on = c("xa", "xb", "xc"))
fitx("weibull", on = c("xa", "xb"))
```

## 8. Methods to Check Parametric Assumptions

*Book chapter 8, pp. 213-228. The transformed survivor function and the pseudoresiduals; command files `ehh1` to `ehh4` and the `*s`/`*p` pairs.*

Each of these transforms the survivor function so that one distribution
appears as a straight line, and regresses it on time. The closer to a
line, the better that distribution fits.

The regression runs on the survivor function at five decimals, the
precision `ehc5.ple` carries: the command files read that file rather than
the estimate in memory, and the transformations are sensitive enough for
it to show in the t statistics.

The axes are each command file's own.

```{r ehhsetup, box = "ehh1 to ehh4: the four transformations [book pp. 215-218, Boxes 8.1.1-8.1.3]"}
s <- tda_survivor(pl, conf.int = NULL)
s <- s[s$survivor > 0 & s$survivor < 1 & s$time > 0, ]
s$survivor <- round(s$survivor, 5)   # the precision ehc5.ple carries

checks <- list(
    Exponential    = list(y = log(s$survivor),                    x = s$time,
                          xlim = c(0, 300), ylim = c(-3, 0), xsc = 60, ysc = 1),
    Weibull        = list(y = log(-log(s$survivor)),              x = log(s$time),
                          xlim = c(0, 6),   ylim = c(-6, 2), xsc = 1,  ysc = 1),
    `Log-Logistic` = list(y = log(1 / s$survivor - 1), x = log(s$time),
                          xlim = c(0, 6),   ylim = c(-6, 2), xsc = 1,  ysc = 1),
    `Log-Normal`   = list(y = stats::qnorm(1 - s$survivor),       x = log(s$time),
                          xlim = c(0, 6),   ylim = c(-3, 2), xsc = 1,  ysc = 1))

ehh <- lapply(checks, function(z) {
    keep <- is.finite(z$y) & is.finite(z$x)
    tda_lsreg(y ~ x, data.frame(x = z$x[keep], y = z$y[keep]))
})
# each regression in full: standard errors and t statistics with the
# two coefficients
for (nm in names(ehh)) {
    cat("\n", nm, "\n", sep = "")
    print(ehh[[nm]])
}
```

The slope of the exponential line is the rate the transformation implies,
`r sprintf("%.4f", -coef(ehh$Exponential)[[2L]])`, against the maximum
likelihood estimate of `ehd1`. Each command file ends by drawing its own
fitted line over the curve, the endpoints written into `plotp` by hand
(`plotp = 0, -0.3527, 300, -2.6027` in `ehh1.cf`); here the line comes
from the fit above, and the four panels are laid out as the book's
Figure 8.1.1.

```{r ehhplots, box = "ehh1, ehh2, ehh3, ehh4: the transformed survivor functions [book pp. 215-217, Boxes 8.1.1-8.1.2; Fig. 8.1.1]", fig.height = 3.6, class.source = "folded"}
z <- checks$`Exponential`
b <- coef(ehh$`Exponential`)
p <- tda_ps(data.frame(x = z$x, y = z$y), file = "ehh1.ps",
            width = 80, height = 40, xlim = c(0, 300), ylim = c(-3, 0))
p <- tda_pl(p, "plxa", sc = 60, ic = 5)
p <- tda_pl(p, "plya", sc = 1, ic = 10)
p <- tda_pl(p, "plframe")
p <- tda_pl_lines(p, "x", "y")
p <- tda_pl_polyline(p, c(0, 300), b[[1L]] + b[[2L]] * c(0, 300))
ehh1 <- tda_pl_text(p, "Graphical Test (Exponential)", at = c(150.0, -0.30000000000000004))

z <- checks$`Weibull`
b <- coef(ehh$`Weibull`)
p <- tda_ps(data.frame(x = z$x, y = z$y), file = "ehh2.ps",
            width = 80, height = 40, xlim = c(0, 6), ylim = c(-6, 2))
p <- tda_pl(p, "plxa", sc = 1, ic = 5)
p <- tda_pl(p, "plya", sc = 1, ic = 2)
p <- tda_pl(p, "plframe")
p <- tda_pl_lines(p, "x", "y")
p <- tda_pl_polyline(p, c(0, 6), b[[1L]] + b[[2L]] * c(0, 6))
ehh2 <- tda_pl_text(p, "Graphical Test (Weibull)", at = c(3.0, 1.2))

z <- checks$`Log-Logistic`
b <- coef(ehh$`Log-Logistic`)
p <- tda_ps(data.frame(x = z$x, y = z$y), file = "ehh3.ps",
            width = 80, height = 40, xlim = c(0, 6), ylim = c(-6, 2))
p <- tda_pl(p, "plxa", sc = 1, ic = 5)
p <- tda_pl(p, "plya", sc = 1, ic = 2)
p <- tda_pl(p, "plframe")
p <- tda_pl_lines(p, "x", "y")
p <- tda_pl_polyline(p, c(0, 6), b[[1L]] + b[[2L]] * c(0, 6))
ehh3 <- tda_pl_text(p, "Graphical Test (Log-Logistic)", at = c(3.0, 1.2))

z <- checks$`Log-Normal`
b <- coef(ehh$`Log-Normal`)
p <- tda_ps(data.frame(x = z$x, y = z$y), file = "ehh4.ps",
            width = 80, height = 40, xlim = c(0, 6), ylim = c(-3, 2))
p <- tda_pl(p, "plxa", sc = 1, ic = 5)
p <- tda_pl(p, "plya", sc = 1, ic = 2)
p <- tda_pl(p, "plframe")
p <- tda_pl_lines(p, "x", "y")
p <- tda_pl_polyline(p, c(0, 6), b[[1L]] + b[[2L]] * c(0, 6))
ehh4 <- tda_pl_text(p, "Graphical Test (Log-Normal)", at = c(3.0, 1.5))

r <- tda_combine_ps(list(list(ehh1, ehh2), list(ehh3, ehh4)), width = 170, height = 90)
tda_plot_ps(tda_read_ps(file.path(r$dir, "combined.ps")))
```

The book's section 8.2 checks a model through its pseudoresiduals, in
three steps: `residuals = TRUE` writes them (`ehd2.res`, Box 8.2.3); a
product-limit estimate treats them as durations (`ehd2s.cf`, Box 8.2.5);
and `ehd2p.cf` plots `log S` against the residual. If the model is right
the residuals are standard exponential, so that curve should lie on a
straight line of slope -1 through the origin. `ehd2p.cf` draws that line
beside it, `plotp = 0,0,3,-3`; the plot is read as the distance between
the two.

```{r ehd2res, box = "ehd2.res, ehd2s: the pseudoresiduals and their survivor function [book pp. 97-225, Boxes 4.1.4-4.1.5, 8.2.1, 8.2.3-8.2.5]"}
ehd2r <- tda_rate(with_x("Surv(TFP, DES)"), d, residuals = TRUE)
head(ehd2r$residuals, 10)
tail(ehd2r$residuals, 6)
rp <- tda_ple(Surv(Residual, Des) ~ 1, ehd2r$residuals)
rp$episodes   # the residuals treated as durations
print(rp, n = 10)
```

```{r ehd2pplot, box = "ehd2p: log S against the residual, with its reference line [book p. 226, Box 8.2.6]", class.source = "folded"}
resid_panel <- function(fit, label, file) {
    rp <- tda_ple(Surv(Residual, Des) ~ 1, fit$residuals)
    rs <- tda_survivor(rp)
    rs <- rs[rs$time > 0 & rs$survivor > 0 & rs$survivor < 1, ]
    rs$LogSurv <- log(rs$survivor)
    p <- tda_ps(rs, file = file, width = 80, height = 40,
                xlim = c(0, 4), ylim = c(-3, 0))
    p <- tda_pl(p, "plxa", sc = 1, ic = 0)
    p <- tda_pl(p, "plya", sc = 1, ic = 0)
    p <- tda_pl_lines(tda_pl(p, "plframe"), "time", "LogSurv")
    p <- tda_pl_polyline(p, c(0, 3), c(0, -3))   # the standard exponential
    tda_pl_text(p, label, at = c(1.5, -0.7))
}
plot(resid_panel(ehd2r, "Exponential Model", "ehd2p.ps"))
```

The exponential curve sits above the reference at the start, crosses it
around 1.7 and falls away below it after that -- the same departure the
transformed survivor function showed in `ehh1`.

Box 8.2.6's footnote lists the command files for the other models, and
the book's Figure 8.2.1 puts eight of them side by side. Each is the
same three steps on a different fit.

```{r f821, box = "ehd2p to ehg13p: the pseudoresidual check [book pp. 226-227, Box 8.2.6; Fig. 8.2.1]", fig.height = 7.2, class.source = "folded"}
resid_fits <- list(
    "Exponential Model" = function()
        tda_rate(with_x("Surv(TFP, DES)"), d, residuals = TRUE),
    "Piecewise Constant Exponential (I)" = function()
        pc(with_x("Surv(TFP, DES)"), d, residuals = TRUE),
    "Piecewise Constant Exponential (II)" = function()
        pc(with_x("Surv(TFP, DES)"), d, model = "exponential_periods2",
           tp = "0,24,60", residuals = TRUE),
    "Gompertz Model" = function()
        fitx("gompertz", on = "xb", residuals = TRUE),
    "Weibull Model"      = function() fitx("weibull", residuals = TRUE),
    "Log-logistic Model" = function() fitx("loglogistic", residuals = TRUE),
    "Log-normal Model"   = function() fitx("lognormal", residuals = TRUE),
    "Sickle Model"       = function() fitx("sickle", residuals = TRUE))

# the arrangement of Figure 8.2.1: exponential, Gompertz, Weibull and
# sickle down the left column, the two piecewise constant models,
# log-logistic and log-normal down the right
i <- 0L
panels <- lapply(names(resid_fits), function(nm) {
    i <<- i + 1L
    resid_panel(resid_fits[[nm]](), nm, sprintf("res%02d.ps", i))
})
r <- tda_combine_ps(split(panels[c(1, 2, 4, 3, 5, 6, 8, 7)], rep(1:4, each = 2)), width = 170, height = 180)
tda_plot_ps(tda_read_ps(file.path(r$dir, "combined.ps")))
```

The book reads these as the exponential and the Weibull departing most
from the reference, the Gompertz, log-logistic and log-normal tracking
it closely, and the two piecewise constant models closest of all. They
are hints for model selection rather than a goodness-of-fit test: the
log-logistic, log-normal and piecewise constant models resemble each
other so closely here that the choice between them has to rest on
something other than these plots.

## 9. Semi-parametric Transition Rate Models

*Book chapter 9, pp. 229-251. Cox models; command files `ehi1` to `ehi13`.*

**ehi1** is the Cox model with the book's covariates. `tda_coxph()` is
`tda_rate(model = 1)` under a name that says what it is.

```{r ehi1, box = "ehi1: a Cox model [book pp. 232-243, Boxes 9.1.2, 9.3.3]"}
ehi1 <- tda_coxph(with_x("Surv(TFP, DES)"), d)
summary(ehi1)
ehi1$episodes   # the episode table TDA reports first
```

**ehi2** adds marriage as a time-varying covariate. `time` is TDA's
position within the episode, so the definition has to be evaluated inside
TDA rather than computed here; `helpers=` writes `MDate` for it to refer
to, without `MDate` entering the model.

```{r ehi2, box = "ehi2: marriage as a time-varying covariate [book pp. 233-235, Boxes 9.1.3, 9.2.1-9.2.2]"}
dm <- d
dm$MDate <- ifelse(dm$TMAR <= 0, 10000, dm$TMAR - dm$TStart)
ehi2 <- tda_coxph(stats::reformulate(c(XVARS, "MARR"),
                                     response = "Surv(TFP, DES)"), dm,
                  helpers = "MDate",
                  define = list(MARR = "gt(time,MDate)"))
ehi2
ehi2$episodes   # the episode table TDA reports first
```

**ehi3**, **ehi4** and **ehi5** are Cox models on the split episodes of
`ehf` -- covariates alone, then with marriage, then distinguishing
married men.

```{r ehi345, box = "ehi3, ehi4, ehi5: Cox models on the split episodes [book pp. 237-239, Boxes 9.2.3-9.2.6]"}
ehi3 <- tda_coxph(stats::reformulate(XVARS,
                                     response = "Surv(ts, tf, des)"), sp)
ehi3
ehi3$episodes   # all three are fitted to these same split episodes

tda_coxph(stats::reformulate(c(XVARS, "MARR"),
                                  response = "Surv(ts, tf, des)"), sp)

tda_coxph(stats::reformulate(c(XVARS, "MARR", "MarrMen"),
                                  response = "Surv(ts, tf, des)"), sp)
```

**ehi7** and **ehi8** ask the proportionality question twice: once as an
interaction with log time, once by letting the effect vary between
periods. Either way, a proportional effect is one that does not move.

```{r ehi78, box = "ehi7, ehi8: proportional hazards [book pp. 243-245, Boxes 9.3.2, 9.3.4-9.3.5]"}
dw <- d
dw$Women <- as.integer(dw$SEX == 2)
ehi7 <- tda_coxph(stats::reformulate(c(XVARS, "Women", "WTest"),
                                     response = "Surv(TFP, DES)"), dw,
                  define = list(WTest = "Women * (log(time) - 4.73)"))
ehi7
ehi7$episodes

ehi8 <- tda_coxph(stats::reformulate(c(XVARS, "Women"),
                                    response = "Surv(TFP, DES)"), dw,
                  options = list(tp = "0 (12) 96"))
summary(ehi8)   # the global goodness-of-fit test comes with the summary
```

**ehi6** asks it graphically instead: `log(-log(S))` per sex, parallel if
the hazards are proportional. The axes, tick spacing and the two labels
are `ehi6.cf`'s own.

```{r ehi6, box = "ehi6: the graphical check [book pp. 241-242, Box 9.3.1; Fig. 9.3.1]", class.source = "folded"}
lls <- tda_survivor(pl2, conf.int = NULL)
lls <- lls[lls$survivor > 0 & lls$survivor < 1, ]
lls$LLS <- log(-log(lls$survivor))
p <- tda_ps(lls, width = 80, height = 40, xlim = c(0, 240), ylim = c(-5, 1))
p <- tda_pl(p, "plxa", sc = 24, ic = 0)
p <- tda_pl(p, "plya", sc = 1, ic = 0)
p <- tda_pl_lines(p, "time", "LLS", by = "group", lty = 1)  # ehi6.cf sets no lt=
p <- tda_pl_text(p, "Women", at = c(192, 0.6))
plot(tda_pl_text(p, "Men", at = c(216, 0.1)))
```

**ehi9** gives each sex its baseline rate instead of assuming one.

```{r ehi9, box = "ehi9: a stratified Cox model [book pp. 246-247, Boxes 9.3.6-9.3.7]"}
ehi9 <- tda_coxph(stats::reformulate(c(XVARS, "strata(SEX)"),
                                     response = "Surv(TFP, DES)"), d)
ehi9
ehi9$episodes   # one block per stratum, as TDA reports them
```

**ehi10** evaluates the baseline rate and its cumulation, **ehi11** draws
it, **ehi12** smooths it with a spline and **ehi13** draws the smoothed
rate.

```{r ehi1011, box = "ehi10: the baseline rate, two cohorts [book pp. 249-250, Boxes 9.4.1-9.4.2]"}
# ehi10.cf asks for two constellations, cohort 1 and cohort 3, both
# written to ehi10.bl; they come back as sub-tables of $rates, told
# apart by ID
ehi10 <- tda_coxph(with_x("Surv(TFP, DES)"), d,
                   prate = list(
                       list(EDU = 13, COHO2 = 0, COHO3 = 0,
                            LFX = 5, PNOJ = 1, PRES = 30),
                       list(EDU = 13, COHO2 = 0, COHO3 = 1,
                            LFX = 5, PNOJ = 1, PRES = 30)))
ehi10
ehi10$episodes   # the episodes behind the baseline

bl <- ehi10$rates
cols <- c("ID", "Time", "RiskSet", "CumRate", "BaselineRate")
head(bl[, cols], 10)
tail(bl[bl$ID == 0, cols], 8)
```

`ehi11.cf` draws the two cumulated rates against time (the book's
Figure 9.4.1).

```{r ehi11plot, box = "ehi11: the two cohorts' cumulated rates [book p. 250, Fig. 9.4.1]", class.source = "folded"}
p <- tda_ps(bl, width = 80, height = 40, xlim = c(0, 300), ylim = c(0, 5))
p <- tda_pl(p, "plxa", sc = 60, ic = 5)
p <- tda_pl(p, "plya", sc = 1, ic = 0)
p <- tda_pl_lines(p, "Time", "CumRate", by = "ID", select = "0", lty = 1)
p <- tda_pl_lines(p, "Time", "CumRate", by = "ID", select = "1", lty = 5)
p <- tda_pl_text(p, "COHO1", at = c(240, 2.3))
plot(tda_pl_text(p, "COHO3", at = c(240, 4.2)))
```

The rate itself jumps at every event time, so the book does not smooth it
directly. `ehi12` smooths the *cumulative* rate -- column 7 of `ehi10.bl`,
the one for cohort 1 -- with a cubic spline over a half-month grid, and
`ehi13` plots that spline's first derivative, which is the rate.

```{r ehi12, box = "ehi12: the cumulated rate smoothed [book p. 251]"}
# ehi12.cf selects the first sub-table, cohort 1
b1 <- bl[bl$ID == 0, ]
sm <- tda_spl(b1$Time, b1$CumRate, sig = 0.2, deg = 3,
              rx = seq(0, 300, by = 0.5))$table
head(sm[, c("x", "fitted", "d1")], 4)
```

`ehi13.cf` draws the smoother's first derivative, which is the rate
itself (the book's Figure 9.4.2).

```{r ehi13, box = "ehi13: the smoothed baseline rate [book p. 251, Fig. 9.4.2]", class.source = "folded"}
p <- tda_ps(sm, width = 80, height = 40, xlim = c(0, 240), ylim = c(0, 0.03))
p <- tda_pl(p, "plxa", sc = 60, ic = 5)
p <- tda_pl(p, "plya", sc = 0.01, ic = 0)
plot(tda_pl_lines(p, "x", "d1"))
```

## 10. Models with a Mixture Distribution

*Book chapter 10, pp. 261-271. A gamma-distributed error term on the
exponential and the Weibull; no command file ships for either, so these
two fits are checked against the book's boxes rather than against
TDA's reference output.*

The book's chapter 10 adds unobserved heterogeneity: `mix = 1` is a
gamma mixture on the exponential model (Box 10.2.1) and on the Weibull
(Box 10.2.2); `D Constant` is the log of the mixture variance.

```{r mix, box = "Boxes 10.2.1, 10.2.2: gamma mixtures [book pp. 268-271, Boxes 10.2.1-10.2.2]"}
tda_rate(with_x("Surv(TFP, DES)"), d, options = list(mix = 1))
tda_rate(with_x("Surv(TFP, DES)"), d, model = "weibull",
         options = list(mix = 1))
```


Finally, the product-limit survivor function on its own, the picture the
whole exercise starts from. This one answers to no command file: the
book has no such figure, and `ple5p.cf` -- which the name would suggest
-- draws something else, the two groups with their confidence bands, as
`ehc8` already does above.

```{r closing, box = "The product-limit survivor function (no book box)", class.source = "folded"}
p <- tda_ps(tda_survivor(pl), width = 80, height = 40, xlim = c(0, 300), ylim = c(0, 1))
p <- tda_pl(p, "plxa", sc = 60, ic = 5)
p <- tda_pl(p, "plya", sc = 1, ic = 10)
plot(tda_pl_lines(p, "time", "survivor"))
```
