---
title: "Following the TDA User's Manual"
lang: en
output:
  rmarkdown::html_vignette:
    css: tda-manual.css
    highlight: null
    toc: true
    toc_depth: 2
    includes:
      in_header: theme-toggle.html
pkgdown:
  as_is: true
vignette: >
  %\VignetteIndexEntry{Following the TDA User's Manual}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
source("tda-hooks.R")
library(tdaR)
ex <- function(f) system.file("extdata", "exam", f, package = "tdaR")
```

<p class="dedication">In memoriam Götz Rohwer</p>
<p class="dedication-quote">„So, wir fangen jetzt an.“</p>

## TDA, and this package

TDA -- *Transition Data Analysis* -- is Götz Rohwer's program for the
statistical analysis of event histories: transition rate models,
parametric and semi-parametric, with episode splitting, competing risks,
multi-episode data and time-dependent covariates, and around them the
rest of a working statistician's toolkit -- descriptive statistics, life
tables and product-limit estimation, regression of every ordinary kind,
sequence analysis, loglinear models, graphs, spatial data, and its own
PostScript plotting. It is a command-file program: a text file of
commands, a data file, and a listing of what it computed.

Its history, as the manual's preface and TDA's `history.txt` give
it: begun in 1989 with early support from the Hamburger Institut für
Sozialforschung; grown inside Hans-Peter Blossfeld's project *Household
Dynamics and Social Inequality* at the European University Institute in
Florence, with European Commission support, and then at the University
of Bremen. Version 5.7 (June 1994) was the first the authors called
mature, and it is the program behind Blossfeld and Rohwer's *Techniques
of Event History Modeling* (1995; 2nd ed. 2002). The 6.x line was a
redesign, and from version 6.2 the program and the manual have two
authors, Rohwer and Ulrich Pötter. The 6.x releases, 6.1 in 1997 to
6.4p in March 2009, fall in Rohwer's years as professor at
Ruhr-Universität Bochum (1997–2013), where TDA's homepage still is.
TDA has always been a non-commercial project under the GNU
General Public License; its source is C, and the *User's Manual* of some
800 pages is its documentation.

tdaR is that program, compiled into an R package. The C is TDA's,
made re-entrant so that it can run inside R and be called more than
once; nothing in it has been rewritten, and where it has been changed
the change is a repair of a demonstrated fault, listed in the package's
`doc/changes-from-tda.md`. Each R function writes the command file TDA would have
been given, runs it, and returns what TDA produced -- the printed
listing, the numbers behind it, and the tables -- as an R object with
the usual methods. The results are TDA's, at TDA's precision; the
vignette *Beyond the Manual* shows how that is checked.

This vignette follows the *TDA User's Manual* of Götz Rohwer and Ulrich
Pötter and shows the same examples written against this package. The
section numbers are the manual's own, so that a page of the manual and
the corresponding part of this document can be read side by side, and the
data are the manual's files, shipped with the package under
`extdata/exam`. Boxes hold code and output, as they do there.

Chapter 3 is the data structures, episode, sequence and graph data;
chapter 4 TDA's PostScript plotting; chapter 5 the matrix language and
its expression, smoothing and minimisation commands; chapter 6,
statistical procedures, is the bulk of the document; chapter 7 is
relational data, graphs and clustering, and chapter 8, set-valued data,
closes it.

## 3 Data Structures

### 3.3.3 An Example Data Set

The manual's main example data set is taken from the German Life History
Study (GLHS), collected at the Max-Planck-Institut für Bildungsforschung
in Berlin in 1981--1983 (Mayer and Brückner 1989); the manual thanks
Karl Ulrich Mayer and Hans-Peter Blossfeld for providing it, and it is
the data set used in Blossfeld and Rohwer, *Techniques of Event History
Modeling*. It holds 600 job episodes from 201 respondents. Each
record is one employment episode, and the consecutive jobs of a
respondent's career follow each other in the file.

`tda_rrdat()` reads `rrdat.1`, names its twelve columns, and derives the
variables the manual's `nvar` block derives.

```{r data, box = "The data file rrdat.1, read and named"}
d <- tda_rrdat()
head(d[, 1:12], 9)
```

All dates are century months: 1 is January 1900, 13 is January 1901, and
so on. `TStart` and `TFin` are the first and last month in the job, `TI`
the interview date, `TB` the date of birth, `TE` the date of entry into
the labour market, and `TMAR` the date of marriage, zero for the still
unmarried. `PRES` is the prestige score of the current job and `PRESN`
that of the next one, `-1` when there is no next job.

Three derived variables carry the episode: `TFP` is the duration,
`TFin - TStart + 1`; `DES` is the destination state, 1 for a job change
and 0 for an episode still running at the interview; `LFX` and `PNOJ` are
labour force experience and the number of previous jobs.

**Example 1** Where TDA needs an `edef` command to turn columns into
episode data, the episode is named in R by the left hand side of a
formula: `Surv(TFP, DES)` is `ts = 0, tf = TFP, org = 0, des = DES`.
Every command that takes episode data reports what it was given.

```{r edef, box = "Episode data, single destination state"}
tda_rate(Surv(TFP, DES) ~ 1, d, model = 2)$episodes
```

458 episodes end in an event, a transition from 0 to 1, with a mean
duration of 49 months; 142 are right censored, and their mean of 128
months is no estimate of anything but still says something about the data.

**Example 2** The manual's `ed2.cf` changes nothing but the destination
state, splitting the job changes by the prestige of the next job.
`tda_rrdat(states = 4)` is that data set.

```{r edef4, box = "Episode data, alternative destination states"}
d4 <- tda_rrdat(states = 4)
tda_rate(Surv(TFP, DES) ~ 1, d4, model = 2)$episodes
```

84 episodes end in an upward move, 155 in a downward move, and 219 do not
change the prestige level. The number of censored episodes has not
changed.

`ed3.cf` defines the same data as multi-episode instead: one episode per
job rather than one per person, told apart by the spell number. TDA
prints a table per spell, each closed by its own total (the manual's
Box 8).

```{r ed3, box = "ed3.cf: multi-episode data"}
eb <- suppressWarnings(
    tda_rate(Surv(TFP, DES) ~ 1, d, id = "ID", spell = "NOJ"))
eb$episodes
attr(eb$episodes, "sums")   # the totals under each spell; sn = NA is all spells
```

### 3.6.2 Graph Data

A graph is given to TDA as an edge list, `gdd = I, J, V`: two columns
of node numbers and one value per edge, a negative value standing for
no edge at all, which is how an isolated node is declared. The manual's
Example 1 has five nodes and four edges, node 9 the isolated one, and
its Figure 1 is `gd2.cf`'s drawing of it as a directed, valued graph.

```{r gd2, box = "gd2.cf", fig.width = 2.36, fig.height = 1.48, out.width = "227px", class.source = "folded"}
p <- tda_ps(xlim = c(0, 6), ylim = c(0, 4), width = 60, height = 30)
nodes <- data.frame(id = c(1, 7, 3, 4, 9), x = c(1, 3, 5, 2, 4),
                    y = c(3, 3, 3, 1, 1), grey = 0.8)
edges <- data.frame(from = c(1, 3, 1, 4), to = c(7, 7, 4, 7),
                    label = c(10, 0, 5, 13), arrow = "1.5,1.0")
p <- tda_pl_graph(p, nodes, edges)

plot(p)
```

### 3.6.5 Drawing Graphs

TDA draws a graph with `plg`, nodes placed by hand and edges with
optional curvature, arrowheads and values; `tda_pl_graph()` takes the
nodes and the edges as data frames. This is the manual's Figure
3.6.2-2, the graph of `gd1.dat` drawn by `gd3.cf`: dashed edges carry
the second value variable.

```{r plg1, box = "plg: the graph of gd1.dat (gd3.cf)", fig.width = 3.15, fig.height = 1.87, out.width = "302px", class.source = "folded"}
p <- tda_ps(xlim = c(0, 9), ylim = c(0, 4), width = 80, height = 40)
nodes <- data.frame(id = c(1, 7, 8, 5, 11, 12, 9),
                    x = c(3, 6, 8, 1, 3, 6, 8),
                    y = c(3, 3, 3, 1, 1, 1, 1), grey = 0.8)
edges <- data.frame(from  = c(1, 1, 7, 8, 11, 5, 7, 8, 12),
                    to    = c(5, 7, 1, 8, 12, 1, 1, 7, 11),
                    curve = c(3, 0, 3, 0, 0, 0, -3, 0, 3),
                    label = c(3, 5, 6, 2, 13, 4, 15, 0, 14),
                    lty   = c(1, 1, 1, 1, 1, 5, 5, 5, 5),
                    arrow = "1.5,1.0")
plot(tda_pl_graph(p, nodes, edges))
```

### 3.6.6 Trees

The same tree as `gd15.cf`, laid out radially: `pl = 1` rather than
`gd15`'s `pl = 3`.
```{r gd15a, box = "gd15a.cf", class.source = "folded"}
edges <- data.frame(
    from = c(6, 6, 7, 7, 8, 8, 9, 9, 1, 1, 1, 7),
    to   = c(2, 4, 3, 5, 1, 6, 7, 8, 10, 11, 12, 13),
    v    = c(1.0, 1.5, 2.0, 2.5, 1.5, 1.0, 3.0, 1.2, 1.6, 2.0, 3.0, 4.0))
g <- tda_graph(edges, directed = FALSE)

plot(g, width = 50, height = 70, layout = "tree", rt = 9, pl = 1, nc = 100)
```

The same tree again, hierarchically: `pl = 2`.
```{r gd15b, box = "gd15b.cf", class.source = "folded"}
edges <- data.frame(
    from = c(6, 6, 7, 7, 8, 8, 9, 9, 1, 1, 1, 7),
    to   = c(2, 4, 3, 5, 1, 6, 7, 8, 10, 11, 12, 13),
    v    = c(1.0, 1.5, 2.0, 2.5, 1.5, 1.0, 3.0, 1.2, 1.6, 2.0, 3.0, 4.0))
g <- tda_graph(edges, directed = FALSE)

plot(g, width = 90, height = 50, layout = "tree", rt = 9, pl = 2, nc = 100)
```

`pltree` lays a tree out by itself from a root; `plot(g, layout =
"tree")` is that command, and `gd15.cf`'s tree of `gd1.dat` is the
manual's figure.

```{r pltree, box = "pltree (gd15.cf)", fig.height = 3.6, class.source = "folded"}
tree <- data.frame(from = c(6, 6, 7, 7, 8, 8, 9, 9, 1, 1, 1, 7),
                   to   = c(2, 4, 3, 5, 1, 6, 7, 8, 10, 11, 12, 13),
                   v    = c(1.0, 1.5, 2.0, 2.5, 1.5, 1.0, 3.0, 1.2, 1.6, 2.0, 3.0, 4.0))
plot(tda_graph(tree, directed = FALSE), width = 90, height = 50,
     layout = "tree", rt = 9, pl = 3, nc = 100)
```

### 3.4.2 Sequence Data Types

TDA reads sequence data two ways. Type 1 is one column per time point.
Type 2 reads the columns as state/time pairs, one pair per spell, and
builds the time axis from the times rather than from the column count --
so the same six columns become three variables over 0 to 8 instead of
six over 0 to 5. `type = 2` is TDA's `seqdef(m = 2)`.

```{r seqtype, box = "seq.d2 read both ways (seq3.cf)"}
seqd2 <- read.table(ex("seq.d2"),
                    col.names = c("ID", "Y1", "T1", "Y2", "T2", "Y3", "T3"))
tda_seq_info(seqd2[, -1])

# seq3.cf declares both structures at once, the second as type 2; the
# manual's box shows them in one table, with their own time axes
seqd3i <- read.table(ex("seq.d3"),
                     col.names = c("ID", paste0("Y", 0:7),
                                   "S1", "T1", "S2", "T2", "S3", "T3"))
tda_seq_info(list(paste0("Y", 0:7), c("S1", "T1", "S2", "T2", "S3", "T3")),
             data = seqd3i, type = c(1, 2))

tda_seqpd(seqd2, type = 2, id = "ID")
```

### 3.4.3 Printing Sequence Data

`seqpd` writes the sequence data back out in one of four layouts, TDA's
own `m=`: every structure side by side on one row per case, one row per
case and structure, one row per case and position, or one row per spell
with its origin and destination state and the times it spans. The
manual's example defines two sequence structures and so prints two sets
of columns; this one has a single structure.

```{r seqpd, box = "seqpd (seq5.cf)"}
# seq5.cf declares two structures over seq.d3: Y0..Y7 one column per
# time point, and three state/time pairs read as a type 2 sequence. The
# tables then carry a column per structure, and the time axis runs to 8
# rather than to 7.
seqd3 <- read.table(ex("seq.d3"),
                    col.names = c("ID", paste0("Y", 0:7),
                                  "S1", "T1", "S2", "T2", "S3", "T3"))
pairs <- c("S1", "T1", "S2", "T2", "S3", "T3")
tda_seqpd(seqd3, id = "ID", variables = paste0("Y", 0:7), second = pairs)
tda_seqpd(seqd3, layout = 3, id = "ID", variables = paste0("Y", 0:7),
          second = pairs)
tda_seqpd(seqd3, layout = 4, id = "ID", variables = paste0("Y", 0:7),
          second = pairs)
```

### 3.4.5 State Indicators

`seqsi` turns a sequence into an indicator matrix: for each case and
each time point asked for, one column per state, 1 where the case is in
that state. `m = 2` counts how often each state has been visited up to
that point instead.

```{r seqsi, box = "seqsi (seq6.cf)"}
seqd1i <- read.table(ex("seq.d1"))
names(seqd1i) <- c("ID", paste0("Y", seq_len(ncol(seqd1i) - 1L)))
tda_seqsi(seqd1i, tp = 1:5, id = "ID")
tda_seqsi(seqd1i, tp = 1:5, id = "ID", options = list(m = 2))
```

### 3.3.5 Joining Episode Data

`ejoin` merges two episode files onto one time axis: every point where
either file changes state starts a new sub-episode, and each side's
state is carried into it (`-3` where that side has none). The manual's
own example is two files, six episodes and two, joined into sixteen.

```{r ejoin, box = "ejoin (ej1.cf)"}
ej1 <- read.table(ex("ej1.dat"),
                  col.names = c("ID", "NS", "SN", "TS", "TF", "S", "X1", "X2"))
ej2 <- read.table(ex("ej2.dat"),
                  col.names = c("ID", "NS", "SN", "TS", "TF", "S", "Y1"))
# every column past the structural six is carried through, filled -3
# where that side has no spell
tda_ejoin(ej1[, c("ID", "TS", "TF", "S", "X1", "X2")],
          id = "ID", start = "TS", end = "TF", state = "S",
          with = ej2[, c("ID", "TS", "TF", "S", "Y1")],
          with_id = "ID", with_start = "TS", with_end = "TF",
          with_state = "S",
          options = list(nw = 3, fmt0 = 4, fmt1 = "8.4"))
```

## 4 PostScript Plots

TDA draws its PostScript. The command files of the manual's chapter 4
are reproduced here through the `tda_pl_*` family; each box names the
command file it follows, and the package's `inst/examples/gallery/` carries the same code as a
standalone script beside a PNG of TDA's output for comparison.

### 4.1 Preparing the Plot Environment

The manual's first plot: axes, a frame, the sine function and, by
`plotf1`, its first derivative -- `deriv = 1` in `tda_pl_function()`
draws the derivative of the expression it is given, so the second call
repeats `sin(x1)` as the command file does.

```{r plot1, box = "plot1.cf"}
p <- tda_ps(xlim = c(0, 6), ylim = c(-1, 1), width = 90, height = 50)
p <- tda_pl_axes(p, sc = c(1, 1), ic = c(10, 0))
p <- tda_pl_frame(p)
p <- tda_pl_function(p, "sin(x1)", range = c(0, 6), step = 0.1)
p <- tda_pl_function(p, "sin(x1)", range = c(0, 6), step = 0.1, deriv = 1)
p <- tda_pl_text(p, "sine", at = c(2.7, 0.6))
p <- tda_pl_text(p, "cosine", at = c(4.6, 0.6))

plot(p)
```

```{r plot2, box = "plot2.cf", fig.width = 3.94, fig.height = 6.21, out.width = "378px"}
p <- tda_ps(xlim = c(0, 595), ylim = c(0, 842), width = 100, height = 150)
p <- tda_pl_axes(p, sc = 72)
p <- tda_pl_frame(p, gray = 0.95)
p <- tda_pl_grid(p, at_x = seq(72, 576, 72), at_y = seq(72, 792, 72))

plot(p)
```

### 4.2 Plotting Coordinate Systems

```{r plot3, box = "plot3.cf", fig.width = 3.54, fig.height = 2.27, out.width = "340px"}
p <- tda_ps(xlim = c(-2, 2), ylim = c(0, 1), width = 90, height = 50)
p <- tda_pl_axes(p, sc = c(1, 0.5), ic = c(2, 0))
p <- tda_pl_frame(p)
p <- tda_pl_grid(p, at_x = c(-1, 0, 1), at_y = 0.5)
p <- tda_pl_labels(p, "Linear Coordinate System", which = "title")
p <- tda_pl_labels(p, "X axis", which = "x")
p <- tda_pl_labels(p, "Y axis", which = "y", sc = 3)

plot(p)
```

```{r plot4, box = "plot4.cf", fig.width = 3.54, fig.height = 2.27, out.width = "340px"}
p <- tda_ps(xlim = c(1, 1000), ylim = c(2, 500), width = 90, height = 50,
           log = "xy")
p <- tda_pl_axes(p, sc = c(10, 5), ic = c(10, 0))
p <- tda_pl_frame(p)
p <- tda_pl_grid(p, at_x = c(10, 100, 1000), at_y = c(10, 50, 250))
p <- tda_pl_labels(p, "Logarithmic Coordinate System", which = "title")
p <- tda_pl_labels(p, "X axis", which = "x")
p <- tda_pl_labels(p, "Y axis", which = "y", sc = 3)

plot(p)
```

A second pair of axes, drawn at a position given in user coordinates
rather than on the boundary of the plot: `at =` and `dir =` in
`tda_pl_axis()` are `plx`'s and `ply`'s parameters.

```{r plot5, box = "plot5.cf"}
p <- tda_ps(xlim = c(-2, 2), ylim = c(0, 1), width = 90, height = 50)
p <- tda_pl_axes(p, sc = c(1, 0.5), ic = c(2, 0))
p <- tda_pl_axis(p, "x", sc = 1, ic = 5, dir = 1, at = c(-1, 1, 1, 1))
p <- tda_pl_axis(p, "y", sc = 0.5, ic = 5, dir = 1, at = c(2, 0, 2, 1))

plot(p)
```

A second panel on the same page, placed with `tda_pl_panel()`.

```{r plot6, box = "plot6.cf"}
# plot6.cf places the first plot with psorg = 100,600, in PostScript
# points; origin = is in mm like width = and height =
p <- tda_ps(xlim = c(0, 6), ylim = c(-1, 1), width = 50, height = 40,
            origin = c(100, 600) * 25.4 / 72)
p <- tda_pl_axes(p, sc = 1)
p <- tda_pl_function(p, "sin(x1)", range = c(0, 6), step = 0.1)
p <- tda_pl_text(p, "Plot 1", at = c(4, 0.2))

# the second plot has psorg = 270,600: 170 points to the right of the
# first, and a panel's origin is relative to the previous one's
p <- tda_pl_panel(p, origin = c(170, 0) * 25.4 / 72, width = 50, height = 30)
p <- tda_pl_axes(p, sc = 1)
p <- tda_pl_function(p, "sin(x1)", range = c(0, 6), step = 0.1)
p <- tda_pl_text(p, "Plot 2", at = c(4, 0.2))

plot(p)
```

### 4.3 Combining Plot Files

`dplot` puts PostScript files that already exist into a grid. The two
coordinate systems of 4.2 are drawn again as files, and
`tda_combine_ps()` places each of them twice on one page.

```{r plot7, box = "plot7.cf", fig.width = 3.54, fig.height = 2.27, out.width = "340px"}
p3 <- tda_ps(xlim = c(-2, 2), ylim = c(0, 1), width = 90, height = 50)
p3 <- tda_pl_axes(p3, sc = c(1, 0.5), ic = c(2, 0))
p3 <- tda_pl_frame(p3)
p3 <- tda_pl_grid(p3, at_x = c(-1, 0, 1), at_y = 0.5)
p3 <- tda_pl_labels(p3, "Linear Coordinate System", which = "title")
p3 <- tda_pl_labels(p3, "X axis", which = "x")
p3 <- tda_pl_labels(p3, "Y axis", which = "y", sc = 3)
f3 <- tda_ps_file(p3)

p4 <- tda_ps(xlim = c(1, 1000), ylim = c(2, 500), width = 90, height = 50,
             log = "xy")
p4 <- tda_pl_axes(p4, sc = c(10, 5), ic = c(10, 0))
p4 <- tda_pl_frame(p4)
p4 <- tda_pl_grid(p4, at_x = c(10, 100, 1000), at_y = c(10, 50, 250))
p4 <- tda_pl_labels(p4, "Logarithmic Coordinate System", which = "title")
p4 <- tda_pl_labels(p4, "X axis", which = "x")
p4 <- tda_pl_labels(p4, "Y axis", which = "y", sc = 3)
f4 <- tda_ps_file(p4)

r <- tda_combine_ps(list(c(f3, f3), c(f4, f4)), width = 100, height = 60,
                    file = "plot7.ps")

tda_plot_ps(file.path(r$dir, "plot7.ps"))
```

### 4.4.1 Polygons and Scatterplots

`plot8.cf` draws twenty points with `rd(-2, 2)`, sorts them, and takes
`nd()` of each -- TDA's standard normal distribution function. Both
are used here, so the points are the manual's.

```{r plot8, box = "plot8.cf", fig.width = 3.54, fig.height = 2.27, out.width = "340px"}
pdat <- tda_derive(data.frame(RDS = sort(tda_rd(20, -2, 2))),
                   ND = "nd(RDS)")

p <- tda_ps(pdat, xlim = c(-2, 2), ylim = c(0, 1), width = 90, height = 50)
p <- tda_pl_axes(p, sc = c(0.5, 1), ic = c(0, 10), fmt = c("4.1", NA))
p <- tda_pl_frame(p)
p <- tda_pl_points(p, "RDS", "ND", symbol = 4, size = 1.5, lty = 1)

plot(p)
```

The nine line types, `lty = 1` to `9`.

```{r plot9, box = "plot9.cf", fig.width = 3.94, fig.height = 2.66, out.width = "378px"}
p <- tda_ps(xlim = c(-2, 2), ylim = c(0, 1), width = 100, height = 60)
p <- tda_pl_frame(p)
for (lt in 1:9)
    p <- tda_pl_lines(p, x = c(-1.5, 0.5), y = c(lt, lt) / 10, lty = lt)
for (lt in 1:9)
    p <- tda_pl_text(p, sprintf("line type %d", lt), at = c(0.75, lt / 10))

plot(p)
```

The seventeen marker symbols: `symbol =` in `tda_pl_text()` is
`pltext`'s `s=`, and puts the marker at the text position.

```{r plot10, box = "plot10.cf", fig.width = 3.94, fig.height = 2.66, out.width = "378px"}
p <- tda_ps(xlim = c(-2, 2), ylim = c(0, 1), width = 100, height = 60)
p <- tda_pl_frame(p)

for (s in 9:1)
    p <- tda_pl_text(p, sprintf("marker %d", s), at = c(-1.5, (10 - s) / 10),
                     symbol = s)
for (s in 17:10)
    p <- tda_pl_text(p, sprintf("marker %d", s), at = c(0, (19 - s) / 10),
                     symbol = s)

plot(p)
```

### 4.4.3 Plotting Text

TDA writes special characters as `@nnn` (the Symbol font) or `\\nnn`
(Times-Roman, through its own re-encoding table); both round-trip, a
literal UTF-8 character does not. `plots.cf` shows each escape and,
beside it, the escape written out literally: `\\100` is the `@` sign
and `\\\\` a backslash, so `\\100141` prints as `@141`, and `\\324`
is the quotation mark around each. The R strings below carry one more
level of backslashes than the command file, R's own.

```{r plots, box = "plots.cf", fig.width = 4.72, fig.height = 2.66, out.width = "453px"}
p <- tda_ps(xlim = c(0, 10), ylim = c(0, 5), width = 120, height = 60)
p <- tda_pl_axes(p, sc = 1, ic = c(0, 2))
p <- tda_pl_frame(p)
p <- tda_pl_text(p, "@141 = 2.5 \\245 @142", at = c(1, 4))
p <- tda_pl_text(p, "pltext (xy=1,4) = \\324\\100141 = 2.5 \\\\245 \\100142\\324",
                 at = c(4, 4))
p <- tda_pl_text(p, "@123 = @173 @161 @174 @161 @316 @121 @175", at = c(1, 2))
p <- tda_pl_text(p, paste("pltext (xy=1,2) = \\324\\100123 = \\100173 \\100161",
                          "\\100174 \\100161 \\100316 \\100121 \\100175\\324"),
                 at = c(2.2, 0.9))
p <- tda_pl_text(p, "Sm\\277rebr\\277d", at = c(1, 3))
p <- tda_pl_text(p, "pltext (xy=1,3) = \\324Sm\\\\277rebr\\\\277d\\324", at = c(4, 3))

p <- tda_pl_lines(p, x = c(3.5, 2.7), y = c(4.05, 4.05), a = "1,1")
p <- tda_pl_lines(p, x = c(3.5, 2.7), y = c(3.05, 3.05), a = "1,1")
p <- tda_pl_arc(p, x = c(1.5, 3.2, 3.2), y = c(1.5, 1.5, 2.1), curvature = -90)
p <- tda_pl_lines(p, x = c(3.2, 2.9), y = c(2.1, 2.1), a = "1,1")
p <- tda_pl_arc(p, x = c(1.5, 1.5), y = c(1.5, 0.95), curvature = -90)
p <- tda_pl_lines(p, x = c(1.5, 2), y = c(0.95, 0.95))

plot(p)
```

### 4.4.5 Step Functions

`plot11.cf` is `plot8.cf` with `dir=1` on the `plot` command: the pen
moves along the x axis first and then along y, and the twenty points on
the normal distribution function become a step function. `dir` keeps
its name in `tda_pl_lines()`.

```{r plot11, box = "plot11.cf", fig.width = 3.54, fig.height = 2.27, out.width = "340px"}
pdat <- tda_derive(data.frame(RDS = sort(tda_rd(20, -2, 2))),
                   ND = "nd(RDS)")

p <- tda_ps(pdat, xlim = c(-2, 2), ylim = c(0, 1), width = 90, height = 50)
p <- tda_pl_axes(p, sc = c(0.5, 1), ic = c(0, 10), fmt = c("4.1", NA))
p <- tda_pl_frame(p)
p <- tda_pl_lines(p, "RDS", "ND", dir = 1, gray = 0.9)

plot(p)
```

### 4.4.6 General Functions

Two normal densities, one of them filled, and a line at each mean. The
expression string goes to TDA's parser, not R's, so it is written with
TDA's functions -- here `ndf()`, the standard normal density.

```{r plot12, box = "plot12.cf", fig.width = 3.54, fig.height = 2.27, out.width = "340px"}
p <- tda_ps(xlim = c(-3, 3), ylim = c(0, 1), width = 90, height = 50)
p <- tda_pl_axes(p, sc = c(0.5, 1), ic = c(0, 10), fmt = c("4.1", NA))
p <- tda_pl_frame(p)
p <- tda_pl_function(p, "ndf((x1 - 1.5) / 0.5) / 0.5", range = c(0, 3),
                     step = 0.05)
p <- tda_pl_lines(p, x = c(1.5, 1.5), y = c(0, 0.798))
p <- tda_pl_function(p, "ndf(x1)", range = c(-3, 3), step = 0.1, gray = 0.8)
p <- tda_pl_lines(p, x = c(0, 0), y = c(0, 0.399))

plot(p)
```

### 4.4.7 Arrows

A polygon can end in an arrow: `a = xs, ys` gives the length and width
of the head in millimetres. `plot13.cf` draws eight of them from one
point, in two line types and with heads of growing size; `a` keeps its
name in `tda_pl_lines()`.

```{r plot13, box = "plot13.cf", fig.width = 3.54, fig.height = 2.27, out.width = "340px"}
p <- tda_ps(xlim = c(0, 10), ylim = c(0, 6), width = 90, height = 50)
p <- tda_pl_axes(p, sc = 1)
p <- tda_pl_frame(p)

p <- tda_pl_lines(p, x = c(5, 8),   y = c(3, 3),   a = "1,1")
p <- tda_pl_lines(p, x = c(5, 7),   y = c(3, 5),   a = "1.5,1.5")
p <- tda_pl_lines(p, x = c(5, 5),   y = c(3, 5.5), a = "2,2")
p <- tda_pl_lines(p, x = c(5, 3),   y = c(3, 5),   a = "2.5,2.5")
p <- tda_pl_lines(p, x = c(5, 2),   y = c(3, 3),   a = "1,1",   lty = 5)
p <- tda_pl_lines(p, x = c(5, 3),   y = c(3, 1),   a = "1.5,1.5", lty = 5)
p <- tda_pl_lines(p, x = c(5, 5),   y = c(3, 0.5), a = "2,2",   lty = 5)
p <- tda_pl_lines(p, x = c(5, 7),   y = c(3, 1),   a = "2.5,2.5", lty = 5)

p <- tda_pl_text(p, "[1,1]", at = c(8.3, 3))
p <- tda_pl_text(p, "[1.5,1.5]", at = c(7.3, 5))
p <- tda_pl_text(p, "[2,2]", at = c(5.3, 5.6))
p <- tda_pl_text(p, "[2.5,2.5]", at = c(1.6, 5))

plot(p)
```

### 4.4.8 Circles, Arcs, Ellipses

`ploto`, which is `tda_pl_circle()`, draws a circle or, with `angles =`,
an arc of one; `plotk`, which is `tda_pl_arc()`, connects points by
circular arcs that can carry an arrowhead. The manual's Figure 3 labels
one of them with the Symbol font's alpha, which TDA writes as `@141`;
the same escape is used here (see 4.4.3).

```{r plot14, box = "plot14.cf", fig.width = 3.54, fig.height = 2.27, out.width = "340px"}
p <- tda_ps(xlim = c(0, 10), ylim = c(0, 6), width = 90, height = 50)
p <- tda_pl_axes(p, sc = 1)
p <- tda_pl_frame(p)

p <- tda_pl_lines(p, x = c(1, 4), y = c(2, 5), lty = 5)
p <- tda_pl_arc(p, x = c(1, 4), y = c(2, 5), curvature = 45, arrow = c(1, 1))
p <- tda_pl_lines(p, x = c(1, 1), y = c(2, 4), lty = 5)
p <- tda_pl_circle(p, at = c(1, 2), r = 10, angles = c(45, 90))
p <- tda_pl_text(p, "plotk(a=1,1,sc=45) = a,b,c,d;", at = c(0.5, 1))
p <- tda_pl_text(p, "(a,b)", at = c(0.7, 1.6))
p <- tda_pl_text(p, "(c,d)", at = c(4.2, 5.1))
p <- tda_pl_circle(p, at = c(1, 2), r = 0.8, angles = c(45, 90), lty = 2)
p <- tda_pl_text(p, "@141", at = c(1.15, 2.4))

p <- tda_pl_arc(p, x = c(6, 6, 7), y = c(2, 4, 5), curvature = 45)
p <- tda_pl_arc(p, x = c(9, 9, 8), y = c(2, 4, 5), curvature = -45)
p <- tda_pl_text(p, "plotk(sc=45) = 6,2,6,4,7,5;", at = c(5.5, 1.3))
p <- tda_pl_text(p, "plotk(sc=-45) = 9,2,9,4,8,5;", at = c(5.5, 0.8))
p <- tda_pl_lines(p, x = c(5.3, 5.3, 5.7), y = c(1.4, 2.2, 2.5), a = "1,1")
p <- tda_pl_lines(p, x = c(9.6, 9.6, 9.2), y = c(1.2, 2.2, 2.5), a = "1,1")

plot(p)
```

A shaded circle and ellipse, dimension lines, and labelled arrows.

```{r plot15, box = "plot15.cf", fig.width = 3.54, fig.height = 2.27, out.width = "340px"}
p <- tda_ps(xlim = c(0, 10), ylim = c(0, 2), width = 90, height = 50)
p <- tda_pl_axes(p, sc = 1)
p <- tda_pl_frame(p)

p <- tda_pl_circle(p, at = c(2, 1), r = 1, gray = 0.9)
p <- tda_pl_ellipse(p, at = c(7, 1), axes = c(2, 0.5), gray = 0.9)
p <- tda_pl_lines(p, x = c(0, 7), y = c(0.5, 0.5), lty = 5)
p <- tda_pl_lines(p, x = c(0, 7), y = c(1.5, 1.5), lty = 5)
p <- tda_pl_lines(p, x = c(1, 1), y = c(0, 1), lty = 5)
p <- tda_pl_lines(p, x = c(3, 3), y = c(0, 1), lty = 5)
p <- tda_pl_lines(p, x = c(5, 5), y = c(0, 1), lty = 5)
p <- tda_pl_lines(p, x = c(9, 9), y = c(0, 1), lty = 5)
p <- tda_pl_text(p, "ploto (xy=2,1,gs=0.9) = 1;", at = c(1.1, 1.7))
p <- tda_pl_text(p, "plote (xy=7,1,gs=0.9) = 2,0.5;", at = c(5.1, 1.7))
p <- tda_pl_lines(p, x = c(2, 3), y = c(1, 1), a = "1,1")
p <- tda_pl_lines(p, x = c(7, 9), y = c(1, 1), a = "1,1")
p <- tda_pl_lines(p, x = c(7, 7), y = c(1, 1.5), a = "1,1")
p <- tda_pl_text(p, "1", at = c(2.3, 0.85))
p <- tda_pl_text(p, "2", at = c(7.8, 0.85))
p <- tda_pl_text(p, "0.5", at = c(7.2, 1.2))

plot(p)
```

Three ellipses at one centre, two of them rotated: `rotate =` in
`tda_pl_ellipse()` is the third value on `plote`'s right-hand side, the
angle in degrees.

```{r plot16, box = "plot16.cf", fig.width = 3.54, fig.height = 2.27, out.width = "340px"}
p <- tda_ps(xlim = c(0, 10), ylim = c(0, 2), width = 90, height = 50)
p <- tda_pl_axes(p, sc = 1)

p <- tda_pl_ellipse(p, at = c(4, 1), axes = c(3, 0.3), gray = 0.9)
p <- tda_pl_ellipse(p, at = c(4, 1), axes = c(3, 0.3), rotate = 45)
p <- tda_pl_ellipse(p, at = c(4, 1), axes = c(3, 0.3), rotate = -45)

p <- tda_pl_text(p, "plote(xy=4,1,gs=0.9) = 3,0.3;", at = c(7.3, 1.0))
p <- tda_pl_text(p, "plote(xy=4,1) = 3,0.3,-45;", at = c(7.3, 0.5))
p <- tda_pl_text(p, "plote(xy=4,1) = 3,0.3,45;", at = c(7.3, 1.5))
p <- tda_pl_rect(p, x = c(11.5, 11.5), y = c(0, 0))

plot(p)
```

### 4.4.9 Convex Hull

`plot17.cf` draws one hundred points from `X = rd, Y = rd` and puts a
convex hull around them. The draws come from TDA's generator at its
default seed, so these are the manual's points: the PostScript this
writes carries the same drawing operations, in the same order, as TDA's
own run of the command file.

```{r plot17, box = "plot17.cf", fig.width = 3.54, fig.height = 2.27, out.width = "340px", class.source = "folded"}
gen <- tda_rng()
pdat <- data.frame(X = numeric(100), Y = numeric(100))
for (i in 1:100) {
    pdat$X[i] <- gen$rd()
    pdat$Y[i] <- gen$rd()
}
p <- tda_ps(pdat, xlim = c(-0.5, 1.5), ylim = c(-0.5, 1.5),
            width = 90, height = 50)
p <- tda_pl_axes(p, sc = 0.5)
p <- tda_pl_frame(p)
plot(tda_pl_hull(p, "X", "Y", gray = 0.95, symbol = 5, size = 0.7))
```

`plot18.cf` smooths the same hull. `smooth` is `plotch`'s `ns=`, the
number of intervals for its Akima smoothing, and `expand` is `ic=`, a
margin in millimetres: each point of the smoothed outline is pushed away
from the hull's centroid in proportion to its distance from it, so the
outline grows without changing shape.

```{r plot18, box = "plot18.cf", fig.width = 3.54, fig.height = 2.27, out.width = "340px", class.source = "folded"}
plot(tda_pl_hull(p, "X", "Y", gray = 0.95, symbol = 5, size = 0.7,
                 smooth = 10, expand = 2))
```

### 4.4.10 Smoothing Polygons

`plots` draws an Akima spline through the points of two variables, and
`plotsp` the same through literal coordinates; in the C both reach the
same function, as `plot` and `plotp` do. `tda_pl_smooth()` is `plots`,
and the points are given as a data frame.

```{r plot19, box = "plot19.cf", fig.width = 3.54, fig.height = 2.27, out.width = "340px"}
pdat <- data.frame(x = c(1, 2, 3, 4, 5, 6), y = c(1, 2.5, 2, 2.5, 2, 1))

p <- tda_ps(pdat, xlim = c(0, 7), ylim = c(0, 3), width = 90, height = 50)
p <- tda_pl_axes(p, sc = 1)
p <- tda_pl_frame(p)
p <- tda_pl_smooth(p, "x", "y", ns = 10, gray = 0.9)
p <- tda_pl_points(p, "x", "y", symbol = 5)

plot(p)
```

`plot20.cf` closes the same curve: `sc = 1`, which keeps its name.

```{r plot20, box = "plot20.cf", fig.width = 3.54, fig.height = 2.27, out.width = "340px"}
pdat <- data.frame(x = c(1, 2, 3, 4, 5, 6), y = c(1, 2.5, 2, 2.5, 2, 1))

p <- tda_ps(pdat, xlim = c(0, 7), ylim = c(0, 3), width = 90, height = 50)
p <- tda_pl_axes(p, sc = 1)
p <- tda_pl_frame(p)
p <- tda_pl_smooth(p, "x", "y", ns = 10, gray = 0.9, sc = 1)
p <- tda_pl_points(p, "x", "y", symbol = 5)

plot(p)
```

### 4.4.11 Contour Plots

`plotc` draws contour lines of a function of two variables, `x1` and
`x2` in TDA's expression syntax, over a grid of the given resolution;
`plot21.cf`'s function is the sum of two normal densities.

```{r plot21, box = "plot21.cf", fig.width = 2.76, fig.height = 3.06, out.width = "265px"}
p <- tda_ps(xlim = c(-2, 2), ylim = c(-2, 2), width = 70, height = 70)
p <- tda_pl_axes(p, sc = 1)
p <- tda_pl_frame(p)
p <- tda_pl_contour(p,
    "exp(-((x1+1)^2+(x2+1)^2)/2)/(2*pi)+exp(-((x1-1)^2+(x2-1)^2)/2)/(2*pi)",
    levels = c(0.01, 0.05, 0.1, 0.125, 0.15), resolution = c(50, 50),
    gray = 0.9)

plot(p)
```

## 5 Mathematical Procedures

### 5.1.2 Matrices

TDA has its matrix language, and a fair part of it exists because
TDA has no other way to get a matrix into a session. R does, so those
commands have no wrapper and need none:

| TDA | R |
|---|---|
| `mdef` (a literal matrix) | `matrix()` |
| `mdefc` (constant) | `matrix(d, m, n)` |
| `mdefi` (rectangular identity) | `diag(1, m, n)` |
| `mdeff` (from a file) | `as.matrix(read.table(...))` |
| `mpr` (print an expression) | printing the return value |
| `mplog`, `mppar`, `mpcov`, `mpgrad` | `logLik()`, `$estimates`, `vcov()` |

Two do compute something and do have wrappers: `mdefg`, a graph's
adjacency matrix, is `tda_mdefg()`, and `mbrr` is `tda_brr()`.
`mdefb` extracts a block of the data matrix and is reached through the
`commands=` argument of `tda_dblock()`. Everything else in 5.1.4 takes
and returns a plain R matrix.

The manual's matrix chapter prints with `mfmt = 13.10`; these boxes
print with ten digits too, so a number can be read off against it.

### 5.1.4.1 Elementary Matrix Commands

```{r m1, box = "Shape, product, transpose, trace"}
op <- options(digits = 10)
A <- matrix(1:6, 3, 2)
c(tda_mnrow(A), tda_mncol(A))
tda_mmul(A, tda_mtransp(A))
tda_mtrace(tda_mcross(A))
```

`mcross` is `t(A) %*% A`, and `mdiag`, `mdrow`, `mdcol` and `mdiagd`
make or take diagonals.

```{r m2, box = "Diagonals"}
tda_mdiag(1:4)
tda_mdrow(A)
tda_mdcol(A)
tda_mdiagd(A)
```

`msqrtd` and `msqrti` are the elementwise square root and its inverse.

```{r m3, box = "Elementwise square roots"}
tda_msqrtd(matrix(1:4, 2, 2))
tda_msqrti(matrix(1:4, 2, 2))
```

Concatenation, row and column sums, selection and trimming.

```{r m4, box = "Concatenating, summing, selecting, trimming"}
tda_mcath(A, A)
tda_mcatv(A, A)
tda_mrsum(A)
tda_mcsum(A)
tda_msrow(A, 1:2)
tda_mscol(A, 1)
tda_mtrim(A, 1)
tda_mtrim(A, leading_rows = 1)
```

Vectorisation, the Kronecker product, and the permutation commands.
`mpinv` inverts a permutation rather than a matrix.

```{r m5, box = "Vectorising, Kronecker product, permutations"}
tda_mcvec(A)
tda_mrvec(A)
tda_mkp(A, A)
tda_mprow(A, c(2, 1, 3))
tda_mpcol(A, c(2, 1))
tda_mpsym(matrix(1:4, 2, 2), c(2, 1))
tda_mpinv(c(4, 2, 1, 3))
```

`mag` solves a system by Gaussian elimination, and reports what it found
when there is no solution or more than one.

```{r m6, box = "Gaussian elimination"}
G <- matrix(c(1, 2, 3, 5, 6, 7, 1, 2, 1), 3, 3, byrow = TRUE)
tda_mag(G, c(1, 0, 1), c(1, 2, 3))
tda_mag(G, c(1, 0, 1), c(5, 4, 4))
```

### 5.1.4.2 Sorting and Ranking

`msort` sorts the rows of a matrix by a column, `msort1` a single
column, and `mrank` ranks.

```{r m7, box = "msort, msort1, mrank"}
tda_msort(matrix(rev(A), 3, 2), 1)
tda_msort1(matrix(c(1, 2, 1, 2, 1, 3, 1, 4), ncol = 2), 2)
tda_mrank(matrix(c(1, 2, 1, 2, 1, 3, 1, 4), 2), 1)
```

### 5.1.4.3 Standardization and Scaling

```{r m8, box = "mcent, mstand, mscal1"}
tda_mcent(A)
tda_mstand(A)
tda_mscal1(A)
```

### 5.1.4.4 Matrix Norms

```{r m9, box = "The three norms"}
c(tda_mnorm(A), tda_mnorm1(A), tda_mnorm2(A))
```

### 5.1.4.5 Inverse Matrices

`minvd` and `minvs` invert a general and a symmetric matrix. `mginv` is
the generalized inverse, and what makes it one is that `A B A` is `A`
again -- which is the manual's check.

```{r m10, box = "minvd, minvs"}
A
tda_minvd(A)
diag(1:2)
tda_minvs(diag(1:2))
```

```{r m11, box = "A generalized inverse, and the identity that defines it"}
B0 <- matrix(c(1, 5, 0, 2, 6, 0, 3, 7, 0, 4, 8, 0), 4, 3, byrow = TRUE)
Bg <- tda_mginv(B0)
Bg
tda_mmul(tda_mmul(B0, Bg), B0)
```

### 5.1.4.6 Cholesky Decomposition

```{r m12, box = "mchol"}
tda_mchol(diag(1:4))
```

### 5.1.4.7 Least Squares Regression

`mls` takes the design and the response in one matrix, the response
last.

```{r m13, box = "mls"}
mlsA <- matrix(c(1, 2, 3, 4, 7, 11, -1, 1, 0, 5, 6, 11), 4, 3, byrow = TRUE)
mlsA   # the design and the response in one matrix, as the manual shows
tda_mls(mlsA)
```

### 5.1.4.8 Linear Equations

```{r m14, box = "mlse"}
mlseA <- matrix(c(-0.4744, -0.4993, -0.7250, -0.8615, 0.5840, -0.7947,
                  0.1652, 0.0, -0.6587, -0.3450, 0.6687, 0.0), 3, 4,
                byrow = TRUE)
mlseA   # the manual's box prints the input too
tda_mlse(mlseA)
```

### 5.1.4.9 Linear Inequalities

```{r m15, box = "mlsi"}
mlsiA <- matrix(c(1, 2, 3, 5, 6, 11, -1, -2, -3, 0, 1, 1), 4, 3, byrow = TRUE)
mlsiA   # the manual's box prints the input too
tda_mlsi(mlsiA)
```

### 5.1.4.11 Eigenvalues and Eigenvectors

`mev` is the general case, which may have complex eigenvalues, and
`mevs` the symmetric one. The manual's `mev` example is a defective
problem -- 5.236 and 0.764 are each a double eigenvalue -- so the digits
beyond the seventh, and which two vectors span each repeated
eigenvalue's space, depend on the machine's arithmetic; the manual's
box was computed on a different one and differs there, not in the
eigenvalues themselves.

```{r m16, box = "mev and mevs"}
tda_mev(matrix(c(6, -3, 4, 1, 4, 2, 4, 0, 4, -2, 3, 1, 4, 2, 3, 1),
               4, 4, byrow = TRUE))
tda_mevs(matrix(c(1, 1, 1, 1, 2, 0, 1, 0, 10), 3, 3, byrow = TRUE))
```

### 5.1.4.12 Singular Value Decomposition

On a cross-product matrix the singular values are the eigenvalues, which
is the manual's way of showing the two agree. `msvd1` prints as in the
manual's box; `mevs` returns its eigenvalues in descending order where
the manual's box (2002) shows them ascending -- TDA 6.4p's `evectf()`
sorts descending, so that is the program's answer, not the manual's.

```{r m17, box = "msvd1 against mevs"}
A <- matrix(c(1, 0, 0, 5, 7, 0, 1, 0, 1), 3, 3, byrow = TRUE)
S <- tda_mcross(A)
S            # the cross-product matrix the manual's box prints
tda_msvd1(S)
tda_mevs(S)
```

### 5.1.4.13 Linear Programming

The manual's example is `mlp(T,X,Y,1)`: maximise x1 subject to x1 <= 1
and one equality, x2 = 1, which is the last row of T. That is
`tda_mlp1()`, with the equality given as its row.

```{r m18, box = "mlp with one equality row"}
# the manual writes the whole problem as one matrix T: the objective in
# the first row, then the inequality, then the equality
Tlp <- matrix(c(1, 0, 0,
                1, 0, 1,
                0, 1, 1), 3, 3, byrow = TRUE)
Tlp
tda_mlp1(objective = c(1, 0), constraints = matrix(c(1, 0), 1), bounds = 1,
         equalities = matrix(c(0, 1), 1), equalities_bounds = 1)
```

### 5.1.4.14 Some Further Commands

`mproc` is a Procrustes rotation: the rotation, scaling and translation
that brings one configuration onto another. The two squares below differ
by a factor of two, and that is what it recovers.

```{r m19, box = "mproc, mnum"}
X <- matrix(c(0, 0, 1, 0, 1, 1, 0, 1), 4, 2, byrow = TRUE)
Y <- matrix(c(0, 0, 2, 0, 2, 2, 0, 2), 4, 2, byrow = TRUE)
tda_mproc(X, Y)
tda_mnum(3.14159, 6, 3)
```

```{r m20, include = FALSE}
options(op)
```

### 5.2.2 Evaluating Expressions

`tda_evalf()` evaluates an expression in TDA's language, with the
variables given as options.

```{r e1, box = "evalf"}
tda_evalf("sin(x) + cos(x)", options = list(x = 1))$value
sin(1) + cos(1)
```

### 5.2.4 Random Numbers

`tda_rng()` is TDA's generator -- multiplicative congruential, with a
fixed seed -- so a sequence drawn from it is the sequence TDA drew.
`rd()` is uniform, `rdn()` normal. `tda_runif()` is the same generator
as a plain vector.

```{r e2, box = "TDA's random numbers"}
g <- tda_rng()
c(g$rd(), g$rd(a = 0, b = 10), g$rdn())
tda_runif(5)
```

### 5.2.6.6 Counting Changes

The operators of the expression language that have no R counterpart
are reached with `tda_derive()`: it runs `nvar` on a data frame with new
variables given as TDA expressions and returns the frame with them
added. The manual's example for `change` and `cntch` is a small
sequence of states per person, in record mode and then in block mode
(`block = "ID"`, TDA's `dblock=`), where the counts are per person.

```{r change, box = "change and cntch, record mode and block mode"}
seqs <- data.frame(ID = c(1, 1, 1, 2, 2, 3, 3, 3, 3, 3, 3),
                   S  = c(1, 3, 2, 2, 3, 3, 1, 2, 2, 2, 1),
                   T  = c(10, 15, 20, 12, 18, 8, 12, 15, 17, 18, 19))
# the manual's five: whether the state changed, how many changes in
# all, the running count, how far back the last change was, and the time
# at that change
tda_derive(seqs, CH = "change(S)", CNT = "cntch(S)", CCNT = "ccntch(S)",
           LAGCH = "lagch(S)", LAGT = "lag(T,-lagch(S))")
tda_derive(seqs, CH = "change(S)", CNT = "cntch(S)", CCNT = "ccntch(S)",
           LAGCH = "lagch(S)", LAGT = "lag(T,-lagch(S))", block = "ID")
```

### 5.2.6.7 Aggregates

The `g` operators aggregate within blocks: `gcnt` counts the block,
`grec` numbers the record within it, `gmean` and `gstd` average a
variable over it.

```{r aggregates, box = "gcnt, grec, gmean, gstd"}
agg <- data.frame(V = c(1, 4, 5, 3, 6, 2), A = c(1, 1, 1, 2, 7, 7))
# the manual's five: the block's size, the record's number in it,
# the block's number, and whether the record is its first or last
tda_derive(agg, GCNT = "gcnt(A)", GREC = "grec(A)", GSN = "gsn(A)",
           GFIRST = "gfirst(A)", GLAST = "glast(A)")

# gmean and gstd average a variable over the block
tda_derive(agg, GM = "gmean(V,A)", GS = "gstd(V,A)", block = "A")
```

### 5.3.3 Evaluating Functions

`tda_evalf()` evaluates a function; `derivatives = 2` is `evalf2`, the
value with the gradient and the Hessian, which TDA computes
analytically -- the manual's `deriv1.cf`.

```{r deriv, box = "evalf2: a function, its gradient and Hessian (deriv1.cf)"}
tda_evalf("x * y^2", options = list(x = 2, y = 5), derivatives = 2)
```

### 5.4.1 Commands for Numerical Integration

`int1.cf` integrates sqrt(x) over 0..1 with the QNG method, which
should give 2/3; the manual's box shows the approximation and the
number of function calls.

```{r e3, box = "integrate: int1.cf"}
int1 <- tda_integrate("sqrt(x)", from = 0, to = 1)
int1$value
# the rest of what TDA reports: the rule, the relative error it worked
# to, and how many times it called the function
c(method = int1$method, rel_error = int1$rel_error, calls = int1$calls)
```

### 5.4.2 Operators for Numerical Integration

`int(a, b, f)` is the integral as an operator inside an expression, so
its limits may be expressions themselves; `int2.cf` and `int3.cf`.

```{r intop, box = "int() inside an expression (int2.cf, int3.cf)"}
tda_evalf("int(0, 1, sqrt(t))")$value
tda_evalf("int(x, y * z, 2 * z^2)", options = list(x = 0, y = 2, z = 1.5))$value
```

### 5.5.1 Moving Averages

The manual's `ds3.cf` smooths Y = 1, 3, 7, 6, 6 three times with the
weights `gss = 0.5, 0.25`: with the copy-on end-value rule, with
`opt = 2` (the end values repeated), and with `opt = 2` applied twice
(the repeat factor `r=2`). The three output files are the manual's Box 3.

```{r s1, box = "sma: the three smoothings of ds3.cf"}
y <- c(1, 3, 7, 6, 6)
tda_sma(y, weights = c(0.5, 0.25))$table
tda_sma(y, weights = c(0.5, 0.25), options = list(opt = 2))$table
tda_sma(y, weights = c(0.5, 0.25), options = list(opt = 2, r = 2))$table
```

### 5.5.2 Running Median Smoothers

A running median is resistant to the single wild value a moving average
would spread over its whole window.

The manual's `ds4.cf` draws 20 uniform numbers with TDA's generator and
smooths them three ways with the `smd` operator inside `nvar` --
`3R`, `4253Ht` and `3RSSHt` -- then plots the three; `tda_derive()`
is the operator form, `tda_smd()` the command.

```{r s2, box = "smd: the three smoothers of ds4.cf"}
gen <- tda_rng()
ds4 <- data.frame(C = 1:20, A = sapply(1:20, function(i) gen$rd()))
ds4 <- tda_derive(ds4, M1 = "smd[sm=[3R]](A)", M2 = "smd[sm=[4253Ht]](A)",
                  M3 = "smd[sm=[3RSSHt]](A)")
head(ds4)
```

```{r s2plot, box = "smd: ds4.cf, drawn", fig.width = 3.54, fig.height = 2.27, out.width = "340px", class.source = "folded"}
p <- tda_ps(ds4, width = 90, height = 50, xlim = c(0, 21), ylim = c(0, 1))
p <- tda_pl(p, "plxa", sc = 1)
p <- tda_pl(p, "plya", sc = 1, ic = 10)
p <- tda_pl_points(p, "C", "M1", symbol = 5, size = 1)  # ds4.cf marks M1
p <- tda_pl_lines(p, "C", "M1")
p <- tda_pl_lines(p, "C", "M2", lty = 5)
plot(tda_pl_lines(p, "C", "M3", lty = 8))
```

### 5.5.3 Smoothing Splines

`sig=` is the smoothing parameter and `deg=` the degree; the derivatives
come back alongside the fit.

The manual's `ds5.cf` is a sine curve with uniform noise, 100 points,
smoothed with `sig = 1`, and plotted against the curve.

```{r s3, box = "spl: ds5.cf"}
gen <- tda_rng()
ds5 <- data.frame(X = (1:100) / 10)
ds5$Z <- sin(ds5$X)
ds5$Y <- ds5$Z + sapply(1:100, function(i) gen$rd(a = -0.2, b = 0.2))
sp5 <- tda_spl(ds5$X, ds5$Y, sig = 1)
head(sp5$table)
```

`ds5.cf` then draws the data, the true sine and the smoother's fitted
values.

```{r s3plot, box = "spl: ds5.cf, drawn", fig.width = 3.54, fig.height = 2.27, out.width = "340px", class.source = "folded"}
p <- tda_ps(ds5, width = 90, height = 50, xlim = c(0, 10), ylim = c(-1.5, 1.5))
p <- tda_pl(p, "plxa", sc = 1)
p <- tda_pl(p, "plya", sc = 0.5, ic = 5)
p <- tda_pl_points(p, "X", "Y", symbol = 5, size = 0.5)
p <- tda_pl_lines(p, "X", "Z", lty = 5)          # the sine the noise hides
plot(tda_pl_lines(p, x = sp5$table$x, y = sp5$table$fitted))
```

### 5.6.1 The fmin Command

The manual's `fmin1.cf` minimises (x - 5)^2 + (y + 6)^2 from the
starting values 0, 0; its box shows convergence in two iterations to
x = 5, y = -6.

```{r s4, box = "fmin: fmin1.cf"}
tda_minimize("(x - 5)^2 + (y + 6)^2")
```

`fmin2.cf` minimises a function of the data, the sum of (x - A)^2 over
A = 1, ..., 101 -- `fmin` with `v = A`. With data, `fmin` is
`tda_freg()`'s nonlinear least squares; the minimum is at x = 51.

```{r s5, box = "fmin with data: fmin2.cf"}
tda_freg({ fn = (x - A)^2 }, data = data.frame(A = 1:101))
```

## 6 Statistical Procedures

### 6.2.1 Mean and Range of Variables

`tda_dstat()` is the `dstat` command: mean, standard deviation and range
for every variable it is given. The data files of chapter 6 are read
straight from the package.

```{r dstat, box = "dstat on ds1.dat"}
ds1 <- read.table(ex("ds1.dat"), col.names = c("X1", "X2", "X3", "X4"))
tda_dstat(ds1)
```

### 6.2.2 Empirical Distribution Functions

`tda_gdf()` estimates the distribution function, and `what=` asks for the
survivor function or the expected values instead. The manual's
`gdf1.dat` holds three small examples told apart by its first column
(`SEL`, TDA's `tsel`): exact observations (Box 3), right-censored ones
with `cen=` (Boxes 4 and 5), and a two-dimensional one with `grp=`.

```{r gdf, box = "gdf: exact data (the manual's Box 3)"}
gdf1 <- read.table(ex("gdf1.dat"))
names(gdf1) <- c("SEL", "ID", "L1", "YL", "DELTA")
ex0 <- gdf1[gdf1$SEL == 0, ]
tda_gdf(~YL, data = ex0)$table
tda_gdf(~YL, data = ex0, what = "survivor")$table
tda_gdf(~YL, data = ex0, what = "expected_values")$table
```

TDA's `cen=` variable, `DELTA` in the manual, is 1 for an exact
observation; `censor=` here is the other way round, 1 (or `TRUE`) for a
censored one, so the manual's indicator is passed negated.

```{r gdf2, box = "gdf: right censored data (Boxes 4 and 5)"}
ex1 <- gdf1[gdf1$SEL == 1, ]
tda_gdf(~YL, data = ex1, censor = ex1$DELTA == 0)$table
tda_gdf(~YL, data = ex1, censor = ex1$DELTA == 0, what = "expected_values")$table
```

The third block of `gdf1.dat` is the two-dimensional case: four units,
each observed in two dimensions. `id=` and `dimension=` are TDA's
`grp=`, and `method=` its `opt=` -- `"joint1"` for the joint
distribution over the observed points.

```{r gdf3, box = "gdf: the joint distribution (Box 6)"}
ex2 <- gdf1[gdf1$SEL == 2, ]
tda_gdf(~YL, data = ex2, id = ex2$ID, dimension = ex2$L1,
        method = "joint1")$table
tda_gdf(~YL, data = ex2, id = ex2$ID, dimension = ex2$L1,
        method = "joint1", what = "survivor")$table
```

With censoring the same estimator needs a grid: `n_boxes=` is TDA's
`n=`, the number of boxes per dimension, and `sc=` the scaling. The
data are Pruitt's, the manual's Box 8.

```{r gdf4, box = "gdf: joint distribution, censored (Boxes 9 and 10)"}
pruitt <- read.table(ex("gdf2.dat"))
names(pruitt) <- c("ID", "L1", "Y", "CEN")
tda_gdf(~Y, data = pruitt, censor = pruitt$CEN == 0, id = pruitt$ID,
        dimension = pruitt$L1, method = "joint1",
        n_boxes = 64, options = list(sc = 0.5))$table
tda_gdf(~Y, data = pruitt, censor = pruitt$CEN == 0, id = pruitt$ID,
        dimension = pruitt$L1, method = "joint1", n_boxes = 64,
        options = list(sc = 0.5), what = "expected_values")$table
```

The manual's Box 12 sets the two joint methods against each other on a
hundred generated units. `tda_rng()` is TDA's generator, so the data
come out the same here as in `gdf3.cf`.

Both methods reproduce the box: 37.3177, 38.0035 and 46.2348 for method
2, and 37.6603, 37.6603 and 46.2104 for method 1.

A few of the later rows differ in the last digit or two -- 103.47752
here against the manual's 103.4777. `gdf3.cf` writes its generated data
to a file and `gdf4.cf` reads it back at `[10.6]`, so its input is
rounded to six decimals; this passes R's doubles straight to TDA. The
deviance in 6.15.2.1 differs for the same reason.

```{r gdf5, box = "gdf: the two joint methods (Box 12)"}
gen <- tda_rng()
X <- 1:100
Y <- X + sapply(X, function(i) gen$rd(a = -10, b = 10))
RD1 <- sapply(X, function(i) gen$rd())
RD2 <- sapply(X, function(i) gen$rd(a = 10, b = 90))
IS <- RD1 <= 0.2 & Y > RD2
two <- rbind(data.frame(ID = X, L1 = 1L, Y1 = as.numeric(X), CEN = 1L),
             data.frame(ID = X, L1 = 2L, Y1 = ifelse(IS, RD2, Y),
                        CEN = as.integer(!IS)))
ev <- function(m)
    tda_gdf(~Y1, data = two, censor = two$CEN == 0, id = two$ID,
            dimension = two$L1, method = m, what = "expected_values")$table
j1 <- ev("joint1")
j2 <- ev("joint2")

# the box lists every censored unit, the thirteen whose second
# dimension had to be estimated
i <- which(j1$exact2 == 0)
data.frame(j1[i, c("observed1", "observed2", "exact1", "exact2")],
           joint1 = j1$expected2[i], joint2 = j2$expected2[i],
           row.names = NULL)
```

`gdf3.cf` does not run to completion through the real TDA binary, so it has no box here.

### 6.2.3 Quantiles

```{r quant, box = "quant"}
tda_quant(cars)
```

### 6.2.4 Frequency Tables

`tda_freq()` is the one-way table, `tda_freq2()` the two-way one.

```{r freq, box = "freq and freq2"}
tda_freq(ds1[c("X1", "X2", "X3")])
tda_freq2(ds1$X2, ds1$X1)
```

### 6.2.5 Aggregated Tables

`tda_atab()` groups a variable into classes and aggregates a second one
over them.

```{r atab, box = "atab"}
tda_atab(ds1[c("X3", "X4")], breaks = seq(-10, 10, 0.5))
```

### 6.2.6 Covariance and Correlation

```{r corr, box = "corr"}
tda_corr(ds1)
```

### 6.2.7 Contingency Measures

The same two-way table as in 6.2.4, with `contingency = TRUE` adding the
measures of association.

```{r cont, box = "freq2 with contingency measures"}
tda_freq2(ds1$X2, ds1$X1, contingency = TRUE)
```

### 6.2.8 Scatterplots

The points come from TDA's random number generator, `tda_rng()`, so
that the picture is the manual's and not one of R's.

```{r scatterdata, box = "Points from TDA's generator"}
gen <- tda_rng()
n <- 200
X <- Y <- numeric(n)
for (i in seq_len(n)) {
    X[i] <- gen$rd(a = 0, b = 3)
    Y[i] <- sin(X[i]) + gen$rd()
}
df <- data.frame(x = X, y = Y)
```

```{r scatter, box = "A scatterplot", fig.width = 3.54, fig.height = 2.27, out.width = "340px", class.source = "folded"}
p <- tda_ps(data = df, width = 90, height = 50, xlim = c(0, 3), ylim = c(-1, 2.5))
p <- tda_pl_scatter(p, "x", "y", symbol = 5, size = 1)
p <- tda_pl_axes(p, sc = c(1), ic = c(10, 0))
plot(p)
```

**Example 1** A sunflower plot bins the same points and draws a petal per
point in the bin, which is what to reach for when the points overlap.

```{r sunflower, box = "A sunflower plot", fig.width = 3.54, fig.height = 2.27, out.width = "340px", class.source = "folded"}
p <- tda_ps(data = df, width = 90, height = 50, xlim = c(0, 3), ylim = c(-1, 2.5))
p <- tda_pl_scatter(p, "x", "y", type = "sunflower", grid = c(15, 10), size = 3)
p <- tda_pl_axes(p, sc = c(1), ic = c(10, 1))
plot(p)
```

**Example 2** `type = "lowess"` adds the smooth.

```{r lowess, box = "A scatterplot with a lowess smooth", fig.width = 3.54, fig.height = 2.27, out.width = "340px", class.source = "folded"}
p <- tda_ps(data = df, width = 90, height = 50, xlim = c(0, 3), ylim = c(-1, 2.5))
p <- tda_pl_scatter(p, "x", "y", symbol = 5, size = 1, type = "lowess")
p <- tda_pl(p, "plxa", sc = 1, ic = 10)
p <- tda_pl(p, "plya", sc = 1, ic = 0)
plot(p)
```
### 6.2.9 Histogram Plots

```{r hist, box = "A histogram (dh1.cf)", fig.width = 3.54, fig.height = 2.27, out.width = "340px", class.source = "folded"}
dh1 <- read.table(ex("dh1.dat"))
p <- tda_ps(data = dh1, width = 90, height = 50, xlim = c(0, 8), ylim = c(0, 0.4))
p <- tda_pl(p, "plxa", sc = 1)
p <- tda_pl(p, "plya", sc = 0.1)
p <- tda_pl_histogram(p, x = "V1", breaks = c(0, 2, 4, 6, 8))
plot(p)
```

### 6.2.10 Density Plots

The same data smoothed with a triangular kernel.

```{r dens, box = "A kernel density estimate (dh2.cf)", fig.width = 3.54, fig.height = 2.27, out.width = "340px", class.source = "folded"}
p <- tda_ps(data = dh1, width = 90, height = 50, xlim = c(0, 8), ylim = c(0, 0.6))
p <- tda_pl(p, "plxa", sc = 1)
p <- tda_pl(p, "plya", sc = 0.1)
p <- tda_pl_density(p, x = "V1", kernel = "triangle", at = seq(0, 8, 0.1))
plot(p)
```

### 6.4.2 Inequality Measures

```{r ineq, box = "ineq"}
tda_ineq(ds1)
```

### 6.4.3 Segregation Indices

```{r segr, box = "segr"}
ds2 <- read.table(ex("ds2.dat"))
names(ds2) <- c("G", paste0("X", 1:4))
tda_segr(ds2, group = "G")
```

### 6.5.1 Life Tables

`tda_ltb()` is the `ltb` command, and `tp=` gives the interval
boundaries.

```{r ltb, box = "A life table, 30-month intervals"}
lt <- tda_ltb(Surv(TFP, DES) ~ 1, data = d, tp = seq(0, 500, 30))
lt$summary
lt$table

# ltb.1 holds a second block: the estimates themselves, with their
# standard errors
lt$survivor
```

`tda_survivor()` reads the estimate back as a data frame carrying the
survivor function, the density and the rate, which is what the plotting
commands take.

```{r ltbsurv, box = "The estimate as a data frame (ple1.cf)"}
tda_survivor(lt)
```

### 6.5.2 Kaplan-Meier Procedure

`tda_ple()` is the product-limit estimator, `tda_km()` an alias for it.
Case weights are TDA's `cwt` command, and here a column name.

```{r ple, box = "Product-limit estimation with case weights"}
rrdat2 <- read.table(ex("rrdat.2"))
names(rrdat2) <- c("DUR", "CEN", "WT", "GRP")
fit <- tda_ple(Surv(DUR, CEN) ~ as.factor(GRP), data = rrdat2, weights = "WT")
fit$blocks
```

`ple2.cf` is the estimate on the job data with the single destination
state; the manual's Box 5 shows the first rows of its output file.

```{r ple2, box = "ple2.cf: the job data, one destination"}
ple2 <- tda_ple(Surv(TFP, DES) ~ 1, data = d)

# the manual prints part of ple.2: its first rows, a slice from the
# middle, and its last
b <- ple2$blocks[[1L]]
b[c(1:8, 41:46, 128:132), ]

# the table ends with the last observation, where only censoring
# happened: TDA writes no estimate there, so those columns are NA
tail(ple2$blocks[[1]], 2)
```

`ple3.cf` has the four destination states (Example 3): the estimate is
per transition, one sub-survivor function for each of the upward,
lateral and downward job changes, and `$summary` holds one row for each.
`ple3p.cf` plots the three: `tda_survivor()` gives the estimated
functions as a data frame, and each transition is selected by its
origin and destination.

```{r ple4, box = "Sub-survivor functions, four destination states"}
ple <- tda_ple(Surv(TFP, DES) ~ 1, data = d4)
ple$summary
```

```{r ple4plot, box = "One survivor function per destination state", fig.width = 3.15, fig.height = 1.87, out.width = "302px", class.source = "folded"}
srv <- tda_survivor(ple)
p <- tda_ps(srv, width = 80, height = 40, xlim = c(0, 300), ylim = c(0, 1))
p <- tda_pl(p, "plxa", sc = 60, ic = 5)
p <- tda_pl(p, "plya", sc = 1, ic = 10)
p <- tda_pl_lines(p, "time", "survivor", by = "group", select = "0,1", lty = 1)
p <- tda_pl_lines(p, "time", "survivor", by = "group", select = "0,2", lty = 5)
p <- tda_pl_lines(p, "time", "survivor", by = "group", select = "0,3", lty = 8)
plot(p)
```

### 6.5.3 Survivor Function Quantiles

`at_survival=` is TDA's `qo=`: survival probabilities, in decreasing
order, and the time each is reached at, one row per transition and
probability. A transition whose survivor function never drops to a
probability has no row for it, which is why the three blocks are of
different length. (`quantiles=` is the other direction, `qt=`: time
points, and the survivor function reached at each.)

```{r quantiles, box = "Quantiles of the survivor function"}
tda_ple(Surv(TFP, DES) ~ 1, data = d4,
        at_survival = seq(0.9, 0.1, by = -0.1))$quantiles
```

### 6.5.4 Comparing Survivor Functions

`compare = TRUE` runs TDA's four tests of whether the group survivor
functions differ. The manual's example, `ple6.cf`, is Kalbfleisch and
Prentice's two-group data (`rrdat.3`): 40 cases, survival time, status
and group.

```{r compare, box = "ple6.cf: comparing two survivor functions"}
rrdat3 <- read.table(ex("rrdat.3"), col.names = c("TF", "Status", "Group"))
tda_ple(Surv(TF, Status) ~ as.factor(Group), data = rrdat3,
        compare = TRUE)$comparison
```

The same four tests on the job data by sex, whose survivor functions the
next box draws:

```{r compare2, box = "Men and women"}
pl2 <- tda_ple(Surv(TFP, DES) ~ as.factor(SEX), data = d, compare = TRUE)
pl2$comparison
```

The bands are Greenwood's, and `band=` shades them. They widen where few
are still at risk, so the manual cuts the plot at 290 months.

```{r bands, box = "Survivor functions with confidence bands", fig.width = 3.54, fig.height = 2.27, out.width = "340px", class.source = "folded"}
sb <- tda_survivor(pl2)
sb <- sb[sb$time < 290, ]
p <- tda_ps(sb, width = 90, height = 50, xlim = c(0, 300), ylim = c(0, 1))
p <- tda_pl_lines(p, "time", "survivor", by = "group", select = "1",
                  band = c("upper", "lower"), lty = 1)
p <- tda_pl_lines(p, "time", "survivor", by = "group", select = "2",
                  band = c("upper", "lower"), lty = 5)
p <- tda_pl(p, "plxa", sc = 60, ic = 1)
p <- tda_pl(p, "plya", sc = 1, ic = 10)
p <- tda_pl_frame(p)
plot(p)
```



### 6.5.5 State Distributions

`tda_state_dist()` is `epsdat`: at each time point, how many episodes
are in each state. The manual's `ed5.cf` runs it on the small multi-state
file `ed1.dat`; with the origin state on the left-hand side
(`Surv(ts, tf, org, des)`) the counts are those of its Box 2, with the
file's `Total` and `Missing` columns.

```{r epsdat, box = "epsdat: state distributions (ed5.cf)"}
ed1 <- read.table(ex("ed1.dat"),
                  col.names = c("ID", "SN", "ORG", "DES", "TS", "TF", "S1", "S2"))
tda_state_dist(Surv(TS, TF, ORG, DES) ~ 1, ed1, times = seq(0, 30, 5))

# Box 2's second table: the same file read as multi-episode. Naming the
# id (and the spell number) puts a case's later spells with it instead
# of counting them as missing.
tda_state_dist(Surv(TS, TF, ORG, DES) ~ 1, ed1, times = seq(0, 30, 5),
               id = "ID", spell = "SN")
```

### 6.6.1 Sequence Length and Gaps

Sequence data are one row per case, one column per position.

```{r seqlg, box = "seqlg: length and gaps"}
seqd1 <- read.table(ex("seq.d1"))
names(seqd1) <- c("ID", paste0("Y", seq_len(ncol(seqd1) - 1L)))
tda_seqlg(seqd1, id = "ID")
```

### 6.6.2 General Characteristics

```{r seqgc, box = "seqgc"}
tda_seqgc(seqd1, id = "ID")
```

### 6.6.3 State Distributions

```{r seqsd, box = "seqsd and seqen"}
tda_seqsd(seqd1[, -1])
tda_seqen(seqd1[, -1])
```

### 6.6.4 Pattern Matching

Patterns are given as a list: `"-"` is a gap, a vector is a run of
states, and `"*"` matches anything.

```{r seqpm, box = "seqpm"}
tda_seqpm(seqd1[, -1L], patterns = list("-", c(3, 3), c(3, "*", 3)))
```

### 6.7.2.2 The seqm Command

`tda_seqm()` is `seqm`: the optimal matching distance between every pair
of sequences.

```{r seqm, box = "seqm (seqm1.cf)"}
seqmd1 <- read.table(ex("seqm.d1"))
tda_seqm(seqmd1)
```

`print=` asks for the alignment itself rather than the distance, either
as the sequence of operations or as the longest common subsequence.

```{r seqmprint, box = "The alignment behind the distance"}
s <- tda_seqm(seqmd1, print = "sequential")
attr(s, "pairs")
attr(s, "dist_by_time")[, , "t4"]
s <- tda_seqm(seqmd1, print = "lcs")
attr(s, "lcs")
```

### 6.7.2.3 Examples

With `subcost=` the substitution cost is a matrix rather than a constant,
and `dp_matrix = TRUE` keeps the dynamic programming table behind each
distance.

```{r seqmcost, box = "A substitution cost matrix"}
seqmd3 <- read.table(ex("seqm.d3"))
scost <- matrix(0, 10, 10)
for (i in 1:10) for (j in 1:10) scost[i, j] <- 0.1 * abs(i - j)

r <- tda_seqm(seqmd3, indel = 1, subcost = scost, dp_matrix = TRUE)
dp <- attr(r, "dp_matrix")[["2_1"]]
dp$seqA        # the two sequences the matrix aligns
dp$seqB
dp$D
```

`s=` asks seqm to print, for every pair, more than the one distance.
`print = "sequential"` gives the distance after each time point
(`seqm1b.cf`); `print = "lcs"` gives the length of the longest common
subsequence and the positions it occupies (`seqm1c.cf`).

```{r seqms, box = "seqm's s= option (seqm1b.cf, seqm1c.cf)"}
seqmd1 <- read.table(ex("seqm.d1"), col.names = paste0("Y", 1:5))
attr(tda_seqm(seqmd1, print = "sequential"), "pairs")
lcs <- attr(tda_seqm(seqmd1, print = "lcs"), "pairs")
lcs
```

### 6.7.2.4 Consecutive Indel Operations

A second `indel` cost charges consecutive insertions or deletions
differently from the first.

```{r indel, box = "indel costs (seqm4.cf)"}
seqmd4 <- read.table(ex("seqm.d4"))
tda_seqm(seqmd4)
tda_seqm(seqmd4, indel = c(1, 0.5))
tda_seqm(seqmd4, indel = c(1, 0))
```

### 6.7.2.5 Data-based Substitution Costs

`subcost = 2` is TDA's `scost = 2`: substitution costs derived from the
data, 2 - p(a,b) - p(b,a) with p the observed transition probabilities
between the two states, the manual's `seqm5.cf`. The distances are those
of its output file `seqm5.d`.

```{r subcost, box = "seqm5.cf: substitution costs based on the data"}
tda_seqm(seqmd4, subcost = 2)
```

### 6.9.1 The lsreg Command

`tda_lsreg()` is least squares regression, and `summary()` prints it the
way R prints a regression.

```{r lsreg, box = "lsreg"}
lsreg1 <- read.table(ex("lsreg1.dat"))
names(lsreg1) <- c("Height", "Weight")
summary(tda_lsreg(Weight ~ Height, data = lsreg1))
```

### 6.9.1.2 Parameter Constraints

Constraints are written the way TDA writes them, as strings over the
coefficient names `b1`, `b2` and so on.

```{r lsregcon, box = "Equality and inequality constraints"}
lsreg2 <- read.table(ex("lsreg2.dat"))
names(lsreg2) <- c("X1", "X2", "Y")
tda_lsreg(Y ~ 0 + X1 + X2, data = lsreg2,
          equality   = "b1 + b2 = 6",
          inequality = c("b1 = 2", "b2 = 2"))
```

### 6.9.1.3 Dummy Variables

`dgroup=` is TDA's group-of-dummies parameter: one regression per group,
reported in one table.

```{r lsregd, box = "A regression per region"}
lsreg3 <- read.table(ex("lsreg3.dat"))
names(lsreg3) <- c("X1", "G", "Y", "NE", "NC", "SO", "WE")
lsreg3$Y2 <- lsreg3$Y * lsreg3$Y
reg <- tda_lsreg(G ~ Y + Y2, data = lsreg3,
                 dgroup = c("NE", "NC", "SO", "WE"))
summary(reg)

# the per-group table TDA prints for dgrp=: the cases in each and its
# weight
reg$groups

# and the fit statistics the manual's box reports alongside them
unlist(reg$stats)
```

### 6.10.1 L1-Norm Regression

```{r l1reg, box = "l1reg"}
l1reg4 <- read.table(ex("l1reg4.dat"))
names(l1reg4) <- c("Y", "X1", "X2")
tda_l1reg(Y ~ X1 + X2, data = l1reg4)
```

### 6.10.2 Nonparametric Regression

`tda_npreg()` smooths with a kernel, against the curve the data were
generated from.

```{r npregdata, box = "A curve that is hard to recover, plus noise"}
gen <- tda_rng()
n <- 256
X <- E <- numeric(n)
for (i in seq_len(n)) {
    X[i] <- gen$rd()
    E[i] <- 0.32 * gen$rdn()
}
Z <- sin(2 * pi * X * X * X)
npreg <- data.frame(X = X, Y = Z * Z * Z + E)
```

```{r npregfit, box = "npreg: the kernel estimate"}
np <- tda_npreg(Y ~ X, data = npreg, method = "mean", kernel = "quartic",
                bandwidth = 0.1, x = seq(0, 1, 0.01))
head(np$table)
```

```{r npreg, box = "npreg, against the curve it is estimating", fig.width = 3.54, fig.height = 2.27, out.width = "340px", class.source = "folded"}
p <- tda_ps(npreg, width = 90, height = 50, xlim = c(0, 1), ylim = c(-1.2, 1.8))
p <- tda_pl_axes(p, sc = c(0.1, 0.5), ic = 0)
p <- tda_pl_points(p, "X", "Y", symbol = 5, size = 0.5, lty = 0)
p <- tda_pl_function(p, "sin(2*pi*x*x*x)*sin(2*pi*x*x*x)*sin(2*pi*x*x*x)",
                     range = c(0, 1), step = 0.01, lty = "dashed")
p <- tda_pl_lines(p, x = np$table$X, y = np$table$YM, lty = "solid", lw = 1)
plot(p)
```

### 6.11.2 The fml Command

`tda_fml()` maximises a likelihood the user writes. The block in braces
is TDA's function definition, with `fn` the value summed over the data
matrix; the names that are neither variables nor already defined are the
parameters.

The likelihood below is the exponential transition rate model of 6.17.2.1
written out by hand, and it reaches the estimates `tda_rate(model = 2)`
reaches.

```{r fml, box = "An exponential model written as a likelihood (frml1.cf)"}
d$DUR <- d$TFin - d$TStart + 1L
d$W <- as.integer(d$SEX == 2)

tda_fml({
    rate = exp(a0 + COHO2 * a1 + COHO3 * a2 + W * a3)
    l1 = ifelse(DES, log(rate), 0)
    fn = l1 - rate * DUR
}, data = d, start = c(-4, 0, 0, 0))
```

### 6.12.2 Binary Logit and Probit

`tda_qreg()` is the `qreg` command, model 1 the logit and model 2 the
probit. `predictions=` and `standardized=` ask for the two further
tables.

```{r qreg, box = "Binary logit (qr1.cf)"}
qr1 <- read.table(ex("qr1.dat"))
names(qr1) <- c("Dose", "Weight", "Response")
qr1$Log10Dose <- log(qr1$Dose) / log(10)

logit <- tda_qreg(Response ~ Log10Dose, data = qr1, weights = "Weight",
                  predictions = TRUE, standardized = TRUE)
logit
logit$categories
attr(logit$categories, "observations")

# qr1.cf fits the probit beside the logit, on the same data
tda_qreg(Response ~ Log10Dose, data = qr1, weights = "Weight",
         model = "probit")
logit$standardized
logit$predictions
```

```{r probit, box = "The same data, probit"}
tda_qreg(Response ~ Log10Dose, data = qr1, weights = "Weight", model = 2)
```

The two link functions side by side, the logistic and, through TDA's
`nd()`, the standard normal distribution function.

```{r plot_bin, box = "plot-bin.cf", fig.width = 3.54, fig.height = 2.27, out.width = "340px", class.source = "folded"}
p <- tda_ps(xlim = c(-4, 4), ylim = c(0, 1), width = 90, height = 50)
p <- tda_pl_axes(p, sc = c(1, 1), ic = c(5, 10))
p <- tda_pl_function(p, "exp(x1)/(1+exp(x1))", range = c(-4, 4), step = 0.05)
p <- tda_pl_function(p, "nd(x1)", range = c(-4, 4), step = 0.05, lty = 6)
p <- tda_pl_function(p, "nd(x1*sqrt(3)/pi)", range = c(-4, 4), step = 0.05,
                     lty = 2)

plot(p)
```

### 6.12.2.1 Grouped Data

With grouped data the likelihood is binomial, and a random effect can be
integrated out with `intn`, TDA's numerical integration operator. The
starting values come from the fit without the random effect.

```{r qreggrouped, box = "A grouped logit, with and without a random effect (qr5.cf)"}
qr5 <- read.table(ex("qr5.dat"))
names(qr5) <- c("Y", "N", "S", "X")
qr5$SX <- qr5$S * qr5$X

fml <- tda_fml({
    xb = beta0 + S * beta1 + X * beta2 + SX * beta3
    ee = bc(N, Y) * (exp(xb)^Y) / ((1 + exp(xb))^N)
    fn = log(ee)
}, data = qr5)
fml

tda_fml({
    xb = beta0 + S * beta1 + X * beta2 + SX * beta3
    ee = intn(7, bc(N, Y) * (exp(xb + t * gam)^Y) / ((1 + exp(xb + t * gam))^N))
    fn = log(ee)
}, data = qr5, start = c(coef(fml), gam = 0.1), options = list(mina = 4))
```

### 6.12.3 Ordinal Logit and Probit

Models 3 and 4.

```{r ordinal, box = "Ordinal logit and probit (qr2.cf)"}
qr2 <- read.table(ex("qr2.dat"), col.names = c("Y", "X1", "X2"))

qr_ol <- tda_qreg(Y ~ X1 + X2, data = qr2, model = 3, predictions = TRUE)
qr_ol
qr_ol$categories

# the manual's Box 5 is this table in full: the observed values, the
# fitted probability of the observed category, and the cumulative
# probabilities per category
qr_ol$predictions
tda_qreg(Y ~ X1 + X2, data = qr2, model = 4)
```

The same ordinal logit written as a likelihood, which is how the manual
shows what the model is. `eexp` is TDA's logistic function.

```{r ordinalfml, box = "The ordinal logit as a likelihood"}
tda_fml({
    xb = X1 * beta1 + X2 * beta2
    theta0 = 1
    theta1 = eexp(alpha1 + xb)
    theta2 = eexp(alpha2 + xb)
    theta3 = eexp(alpha3 + xb)
    ll = ifelse(Y == 0, theta0 - theta1,
         ifelse(Y == 1, theta1 - theta2,
         ifelse(Y == 2, theta2 - theta3, theta3)))
    fn = log(ll)
}, data = qr2, start = c(-2, -3, -4, 0, 0))
```

### 6.12.4 Multinomial Logit

Model 5, with `nq=` the number of response categories.

```{r multinom, box = "Multinomial logit (qr3.cf)"}
qr3 <- read.table(ex("qr3.dat"), col.names = c("X", "Weight", "Y"))
ml <- tda_qreg(Y ~ X, data = qr3, weights = "Weight", model = 5, nq = 5)
ml

# the distribution over the response's categories, which qreg prints
# before it estimates
ml$categories
attr(ml$categories, "observations")   # the weighted total TDA prints
```

`lvl()` is TDA's level operator, which takes a set of columns as the
levels of one response.

```{r multinomlvl, box = "A multinomial logit over lvl() columns (qr4.cf)"}
qr4 <- read.table(ex("qr4.dat"), col.names = c("Z1", "Z2", "Z3", "Y"))
cml <- tda_qreg(Y ~ 0 + lvl(Z1, Z2, Z3), data = qr4, model = 5, nq = 3,
                predictions = TRUE)
cml
cml$categories
attr(cml$categories, "observations")

# the manual's Box 5 is this table in full
cml$predictions
```

### 6.12.5 Multivariate Probit

Model 6. The constraints zero the terms that are not wanted.

```{r mvprobit, box = "Multivariate probit (qr6.cf)"}
mvp <- tda_qreg(Y ~ X, data = qr3, model = 6, nq = 5, weights = "Weight",
         constraints = c("b9 = 0", "b10 = 0", "b11 = 0", "b12 = 0", "b13 = 0",
                         "b14 = 0", "b15 = 0", "b16 = 0", "b17 = 0",
                         "b18 = 0"))
mvp
unlist(mvp$stats)
```

```{r mvprobit2, box = "Multivariate probit over lvl() columns (qr7.cf)"}
mv <- tda_qreg(Y ~ 0 + lvl(Z1, Z2, Z3), data = qr4, model = 6, nq = 3,
               start = c(-0.1716, 0, 0, 0), constraints = c("b3 = 0", "b4 = 0"))
mv

# the correlation among the latent equations, which TDA writes into its
# ppar= file with the rows commented
mv$correlation
```

### 6.14.1 Poisson Regression

`poisson(mu, y)` is TDA's log Poisson density, so a count model is a
likelihood like any other. `LOGS` becomes an offset once its coefficient
is constrained to one.

```{r poisson, box = "Poisson regression, free and with an offset (cd1.cf)"}
cd1 <- read.table(ex("cd1.dat"),
                  col.names = c("NDI", "Service", "B", "C", "D", "E",
                                "C60", "C65", "C70", "P75"))
cd1$LOGS <- log(cd1$Service)
cd1 <- cd1[cd1$NDI >= 0, ]

pois <- function(...) tda_fml({
    xb = b0 + B * bb + C * bc + D * bd + E * be +
         C60 * bc60 + C65 * bc65 + C70 * bc70 + P75 * bp75 + LOGS * blogs
    exb = exp(xb)
    fn = poisson(exb, NDI)
}, data = cd1, ...)

free <- pois()
free
# what the manual's box reports around the estimates
unlist(free$convergence)

pois(constraints = "b9 = 1")
```

### 6.14.2 Compound Poisson Models

`cd2.cf` is the same count model with a gamma-distributed rate -- the
negative binomial, `negbin(alpha, gamma, y)` in TDA's expression
language -- and `mina = 8` for a more robust algorithm. The
overdispersion parameter `sig` is not identified on these data: its
error is two orders of magnitude larger than its value, and its last
digits depend on where the algorithm stops (the manual's box reads
-14.77, this run -15.06); the regression coefficients are the manual's.

```{r negbin, box = "cd2.cf: a negative binomial count model"}
tda_fml({
    xb = b0 + B * bb + C * bc + D * bd + E * be +
         C60 * bc60 + C65 * bc65 + C70 * bc70 + P75 * bp75 + LOGS * blogs
    gamma = exp(xb)
    sigma = exp(sig)
    alpha = 1 / sigma
    fn = negbin(alpha, gamma, NDI)
}, data = cd1, options = list(mina = 8))
```

### 6.15.2.1 Generalized Linear Models

`tda_glm()` is TDA's `glm`. Without a family it is the Gaussian model
with the identity link, which is least squares; `link = 2` is the log
link, which needs starting values here.

```{r glm, box = "glm, Gaussian, identity and log links"}
plain <- tda_glm(Weight ~ Height, data = lsreg1)
plain

# the manual prints the deviance as 2142.4880. Reading its data file TDA
# stores each variable as a 4-byte float; the wrapper hands R's doubles
# over directly, and stats::glm on the same data agrees with it exactly.
unlist(plain$stats)

gen <- tda_rng()
n <- 100
X <- Y <- numeric(n)
for (i in seq_len(n)) {
    X[i] <- gen$rd(a = -3, b = 3)
    Y[i] <- exp(3 + X[i]) + gen$rd()
}
bad <- tda_glm(Y ~ X, data = data.frame(X, Y), start = c(5, 5), link = 2)
bad

# what the manual's Box 4 reports around the estimates: the degrees of
# freedom, the deviance and Pearson statistics, the scale estimates and
# the rank, and how far the last step moved
unlist(bad$stats)
unlist(bad$convergence)[c("converged", "iterations", "final_change")]
```

### 6.15.2.2 Binomial Models

The number of trials is given separately, and a link can be written out
rather than selected: `custom_link=` takes a TDA expression in `mue`.

```{r glmbin, box = "Binomial models, selected and hand-written links"}
glmbin1 <- tda_glm(Response ~ Log10Dose, data = qr1, weights = "Weight", family = 2)
glmbin1
unlist(glmbin1$stats)

glm3 <- read.table(ex("glm3.dat"), col.names = c("D", "N", "Y"))
glm3$D <- log(glm3$D)

glmbin2 <- tda_glm(Y ~ D, data = glm3, family = 2, link = 3, trials = glm3$N)
glmbin2
unlist(glmbin2$stats)
glmbin1 <- tda_glm(Y ~ D, data = glm3, family = 2, trials = glm3$N,
        custom_link = "log(mue / (1 - mue))")
glmbin1
unlist(glmbin1$stats)
```

### 6.15.2.3 Poisson Models

The offset is the same constrained coefficient as in 6.14.1, given here
as an equality constraint on the GLM.

```{r glmpois, box = "The Poisson family"}
glm5 <- read.table(ex("glm5.dat"), col.names = c("X", "Y"))
glmpois1 <- tda_glm(Y ~ X, data = glm5, family = "poisson")
glmpois1
unlist(glmpois1$stats)

glmpois1 <- tda_glm(NDI ~ B + C + D + E + C60 + C65 + C70 + P75 + LOGS, data = cd1,
        family = "poisson", equality = c("b9 = 1"),
        control = tda_control(maxit = 100))
glmpois1
unlist(glmpois1$stats)
```

### 6.15.2.4 Gamma Models

```{r glmgamma, box = "The Gamma family, log and inverse links"}
gen <- tda_rng()
n <- 1000
X <- Y <- numeric(n)
for (i in seq_len(n)) {
    X[i] <- gen$rd()
    Y[i] <- -exp(1 + X[i]) * (log(gen$rd()) + log(gen$rd())) / 2
}
glmgamma1 <- tda_glm(Y ~ X, data = data.frame(X, Y), family = "Gamma", link = 2,
        start = c(1, 1), control = tda_control(maxit = 50))
glmgamma1
unlist(glmgamma1$stats)

# glm7.cf's second model: the same generator continued, with expectation
# 1/(1+X), and the gamma family's canonical (reciprocal) link
X2 <- Y2 <- numeric(n)
for (i in seq_len(n)) {
    X2[i] <- gen$rd()
    Y2[i] <- -(log(gen$rd()) + log(gen$rd())) / ((1 + X2[i]) * 2)
}
glmgamma2 <- tda_glm(Y2 ~ X2, data = data.frame(X2, Y2), family = "Gamma",
        start = c(1, 1), control = tda_control(maxit = 50))
glmgamma2
unlist(glmgamma2$stats)

glm8a <- read.table(ex("glm8a.dat"), col.names = c("U", "Y", "L"))
glm8a$X <- log(glm8a$U)
glmgamma3 <- tda_glm(Y ~ X, data = glm8a[glm8a$L == 0, ], family = "Gamma",
        start = c(-0.02, 0.02))
glmgamma3
unlist(glmgamma3$stats)
glmgamma4 <- tda_glm(Y ~ X, data = glm8a[glm8a$L == 1, ], family = "Gamma",
        start = c(-0.02, 0.02))
glmgamma4
unlist(glmgamma4$stats)
```

### 6.15.2.5 Inverse Gaussian Models

`glm9.cf` draws from an inverse Gaussian with mean 1 + X through TDA's
own generator, and fits it with the identity link from the starting
values 0.5, 0.5.

```{r glmig, box = "The inverse Gaussian family (glm9.cf)"}
gen <- tda_rng()
n <- 1000
X <- Mu <- Z <- X1 <- Y <- numeric(n)
for (i in seq_len(n)) {
    X[i] <- gen$rd()
    Mu[i] <- 1 + X[i]
    Z[i] <- gen$rdn()^2
    X1[i] <- Mu[i] + Mu[i]^2 * Z[i] -
        Mu[i] * sqrt(2 * Mu[i] * Z[i] + Mu[i]^2 * Z[i]^2)
    Y[i] <- ifelse(gen$rd() <= Mu[i] / (Mu[i] + X1[i]), X1[i], Mu[i]^2 / X1[i])
}
ig <- tda_glm(Y ~ X, data = data.frame(X, Y), family = "inverse.gaussian",
              link = 1, start = c(0.5, 0.5))
ig
unlist(ig$stats)
unlist(ig$convergence)[c("converged", "iterations", "final_change")]
```

### 6.16.1 The freg Command

`tda_freg()` is nonlinear least squares: the function block gives the
model and `fn` the squared residual, summed over the data. The manual's
example is Gallant's (1987, p. 4) data, 30 observations, which the
manual's `freg1.dat` holds; they are typed in here.

```{r freg, box = "freg: Gallant's example (freg1.cf)"}
freg1dat <- read.csv(ex("gallant.dat"))
tda_freg({
    ax = a1 * X1 + a2 * X2 + a4 * exp(a3 * X3)
    fn = (Y - ax)^2
}, data = freg1dat, start = c(-0.1, 1, -1, -1),
   control = tda_control(algorithm = 5))
```

### 6.17.2.1 The Basic Exponential Model

The simplest transition rate model lets the rate vary with the covariates
but not with the duration of the episode. The duration is exponentially
distributed, with density `f(t) = a exp(-a t)`, survivor function
`G(t) = exp(-a t)` and rate `r(t) = a`, and the covariates enter through
an exponential link,

> `r(jk) = exp(A(jk) alpha(jk))`,

where `A` holds a constant one in its first component for the intercept.
This is TDA's model number 2. The `xa(0,1) = ...` parameter of the `rate`
command, which links covariates to the transition from 0 to 1, is the
right hand side of the formula.

**Example 1** The manual's `rt1.cf` links three covariates to the only
transition in the data.

```{r rate2, box = "The command file rt1.cf, in R"}
fit <- tda_rate(Surv(TFP, DES) ~ COHO2 + COHO3 + W, data = d, model = 2,
                residuals = TRUE,
                # rt1.cf's constellation: COHO3 = 1 and W = 1
                prate = list(tp = seq(0, 100, 5), COHO3 = 1, W = 1))
fit
```

`rt1.cf` asks for two things besides the estimates: the generalized
residuals for every episode, and the fitted rate over a time axis at a
chosen covariate constellation.

```{r rate2res, box = "rt1.cf: generalized residuals and the rate table"}
# the manual prints part of res.d: its first rows and its last
head(fit$residuals, 6)
tail(fit$residuals, 6)

# the rate table, and the constellation it was computed at -- which
# covariate values, and the coefficients they multiply
attr(fit$rates, "constellation")
fit$rates
```

Convergence is reached in 5 iterations with TDA's default algorithm, and
the coefficients are the ones printed in the manual. They are not read
back from that printout: `coef()` holds the doubles TDA computed.

```{r rate2coef, box = "The estimates, at full precision"}
coef(fit)
logLik(fit)
fit$logLik_null   # the exponential null model TDA starts from
```

**Example 2** `rt1m.cf` estimates the same three covariates over the
three destination states. Its `DES` differs from `ed2.cf`'s in one
comparison -- `ge(PRESN/PRES - 1, 0.2)` against `gt` -- and one
episode lies on 0.2 to within TDA's comparison tolerance, so the manual's
episode table here reads 85/218 where 3.3.3's read 84/219;
`tda_rrdat(upward = "ge")` is that coding. `relative_risk=` adds the
manual's table of rate ratios.

```{r rate2rr, box = "rt1m.cf: three transitions, and the relative risks (frml1m.cf)"}
d4m <- tda_rrdat(states = 4, upward = "ge")
d4m$W <- as.integer(d4m$SEX == 2)
fitm <- tda_rate(Surv(TFP, DES) ~ COHO2 + COHO3 + W, data = d4m, model = 2,
                 relative_risk = TRUE)
fitm$episodes
fitm$estimates
fitm
fitm$relative_risk
```

### 6.17.2.2 Models with Time Periods

Model 3 is the piecewise constant exponential: a separate rate in each
period given by `tp=`, the covariate effects held common.

```{r rate3, box = "A piecewise constant exponential (rt2.cf)"}
rt2 <- tda_rate(Surv(TFP, DES) ~ COHO2 + COHO3 + W, data = d, model = 3,
                tp = seq(0, 96, 12))
rt2$periods
rt2
```

`rt2c.cf` constrains all nine period coefficients to be equal, which
gives the exponential model of `rt1.cf` back -- the manual's way of
showing what a constraint does.

```{r rate3con, box = "rt2c.cf: all periods constrained equal"}
tda_rate(Surv(TFP, DES) ~ COHO2 + COHO3 + W, data = d, model = 3,
         tp = seq(0, 96, 12),
         constraints = sprintf("b%d - b%d = 0", 1:8, 2:9))
```

### 6.17.2.3 Period-specific Effects

Model 16 gives each period its covariate effects as well as its own
baseline.

```{r rate16, box = "Period-specific effects (rt3.cf)"}
tda_rate(Surv(TFP, DES) ~ COHO2 + COHO3 + W, data = d, model = 16,
         tp = seq(0, 96, 12))
```

### 6.17.3 Parametric Duration Dependence

The parametric family is a matter of the model number, and `TDA_MODELS`
names them, so `model = "weibull"` and `model = 7` are the same request.
Each family has its section in the manual, with the same three
covariates as `rt1.cf`; the boxes below are those sections' models, one
each.

```{r models, box = "The model numbers"}
TDA_MODELS
```

### 6.17.3.1 Polynomial Rates

Models 4 and 5, with `degree=`: the rate, or its logarithm, a polynomial
in time.

```{r poly, box = "Polynomial rates"}
poly1 <- tda_rate(Surv(TFP, DES) ~ COHO2 + COHO3 + W, data = d,
                  model = 4, degree = 2)
poly1
unlist(poly1$convergence)   # what the manual prints above the estimates

poly2 <- tda_rate(Surv(TFP, DES) ~ COHO2 + COHO3 + W, data = d,
                  model = 5, degree = 2)
poly2
unlist(poly2$convergence)
```

### 6.17.3.2 Gompertz-Makeham Models

Model 6. `on=` says which terms of the model the covariates are linked
to; the manual's example puts them on the manual's `xb` and `xc`.

```{r gompertz, box = "Gompertz-Makeham (rt6.cf)"}
tda_rate(Surv(TFP, DES) ~ COHO2 + COHO3 + W, data = d, model = 6,
         on = c("xb", "xc"))
```

Gompertz transition rates at three shape parameters.
```{r plot_gm, box = "plot-gm.cf", fig.width = 3.54, fig.height = 2.27, out.width = "340px", class.source = "folded"}
p <- tda_ps(xlim = c(0, 5), ylim = c(0, 3), width = 90, height = 50)
p <- tda_pl_axes(p, sc = c(1, 1))
p <- tda_pl_frame(p)
p <- tda_pl_function(p, "1+0*x1", range = c(0, 5), step = 1)
p <- tda_pl_function(p, "exp(0.2*x1)", range = c(0, 5), step = 0.1, lty = 5)
p <- tda_pl_function(p, "exp(-0.5*x1)", range = c(0, 5), step = 0.1, lty = 6)

p <- tda_pl_text(p, "c = 0.0", at = c(4.3, 1.1))
p <- tda_pl_text(p, "c = 0.2", at = c(4.3, 2.1))
p <- tda_pl_text(p, "c = -0.5", at = c(4.3, 0.2))

plot(p)
```

### 6.17.3.3 Weibull Models

```{r weibull, box = "Weibull (rt7.cf)"}
tda_rate(Surv(TFP, DES) ~ COHO2 + COHO3 + W, data = d, model = 7)
```

Weibull transition rates at five shape parameters.
```{r plot_wei, box = "plot-wei.cf", fig.width = 3.54, fig.height = 2.27, out.width = "340px", class.source = "folded"}
wei <- function(b) sprintf("%s*1^%s*x1^(%s-1)", b, b, b)

p <- tda_ps(xlim = c(0, 3), ylim = c(0, 3), width = 90, height = 50)
p <- tda_pl_axes(p, sc = c(1, 1))
p <- tda_pl_frame(p)
p <- tda_pl_function(p, "1+0*x1", range = c(0.01, 3), step = 1)
p <- tda_pl_function(p, wei(1.2), range = c(0.01, 3), step = 0.01, lty = 5)
p <- tda_pl_function(p, wei(1.5), range = c(0.01, 3), step = 0.01, lty = 6)
p <- tda_pl_function(p, wei(0.7), range = c(0.01, 3), step = 0.01, lty = 5)
p <- tda_pl_function(p, wei(0.5), range = c(0.01, 3), step = 0.01, lty = 6)

p <- tda_pl_text(p, "b = 0.0", at = c(2.6, 1.1))
p <- tda_pl_text(p, "b = 1.2", at = c(2.6, 1.6))
p <- tda_pl_text(p, "b = 1.5", at = c(2.6, 2.2))
p <- tda_pl_text(p, "b = 0.7", at = c(2.6, 0.6))
p <- tda_pl_text(p, "b = 0.5", at = c(2.6, 0.1))

plot(p)
```

### 6.17.3.4 Sickle Models

```{r sickle, box = "Sickle (rt8.cf)"}
tda_rate(Surv(TFP, DES) ~ COHO2 + COHO3 + W, data = d, model = 8)
```

Sickle transition rates at three scale parameters.
```{r plot_sic, box = "plot-sic.cf", fig.width = 3.54, fig.height = 2.27, out.width = "340px", class.source = "folded"}
sic <- function(b) sprintf("1*x1*exp(-x1/%s)", b)

p <- tda_ps(xlim = c(0, 12), ylim = c(0, 0.8), width = 90, height = 50)
p <- tda_pl_axes(p, sc = c(1, 0.4), ic = c(2, 4))
p <- tda_pl_frame(p)
p <- tda_pl_function(p, sic(1.0), range = c(0, 12), step = 0.1)
p <- tda_pl_function(p, sic(0.5), range = c(0, 12), step = 0.1, lty = 5)
p <- tda_pl_function(p, sic(2.0), range = c(0, 12), step = 0.1, lty = 6)

p <- tda_pl_text(p, "b = 1.0", at = c(4, 0.11))
p <- tda_pl_text(p, "b = 0.5", at = c(2, 0.06))
p <- tda_pl_text(p, "b = 2.0", at = c(10, 0.14))

plot(p)
```

### 6.17.3.5 Log-logistic Models

```{r loglogistic, box = "Log-logistic (rt9.cf)"}
tda_rate(Surv(TFP, DES) ~ COHO2 + COHO3 + W, data = d, model = 9)
```

Log-logistic transition rates at five shape parameters.
```{r plot_ll, box = "plot-ll.cf", fig.width = 3.54, fig.height = 2.27, out.width = "340px", class.source = "folded"}
ll <- function(b) sprintf("%s*x1^(%s-1)/(1+x1^%s)", b, b, b)

p <- tda_ps(xlim = c(0, 3), ylim = c(0, 2), width = 90, height = 50)
p <- tda_pl_axes(p, sc = c(1, 1))
p <- tda_pl_frame(p)
p <- tda_pl_function(p, ll(1.0), range = c(0.01, 3), step = 0.02)
p <- tda_pl_function(p, ll(1.5), range = c(0.01, 3), step = 0.02, lty = 5)
p <- tda_pl_function(p, ll(2.0), range = c(0.01, 3), step = 0.02, lty = 6)
p <- tda_pl_function(p, ll(0.7), range = c(0.01, 3), step = 0.02, lty = 5)
p <- tda_pl_function(p, ll(0.4), range = c(0.01, 3), step = 0.02, lty = 6)

p <- tda_pl_text(p, "0.4", at = c(2.8, 0.016), fs = 1.5)
p <- tda_pl_text(p, "0.7", at = c(2.8, 0.1), fs = 1.5)
p <- tda_pl_text(p, "1.0", at = c(2.8, 0.18), fs = 1.5)
p <- tda_pl_text(p, "1.5", at = c(2.8, 0.33), fs = 1.5)
p <- tda_pl_text(p, "2.0", at = c(2.8, 0.52), fs = 1.5)

plot(p)
```

### 6.17.3.6 Log-normal Models

```{r lognormal, box = "Log-normal (rt12.cf)"}
tda_rate(Surv(TFP, DES) ~ COHO2 + COHO3 + W, data = d, model = 12)
```

Log-normal transition rates at three scale parameters.
```{r plot_ln, box = "plot-ln.cf", fig.width = 3.54, fig.height = 2.27, out.width = "340px", class.source = "folded"}
ln <- function(b) sprintf("ndf(log(x1)/%s)/(%s*x1*(1-nd(log(x1)/%s)))",
                          b, b, b)

p <- tda_ps(xlim = c(0, 7), ylim = c(0, 1.5), width = 90, height = 50)
p <- tda_pl_axes(p, sc = c(1, 0.5))
p <- tda_pl_frame(p)
p <- tda_pl_function(p, ln(1.0), range = c(0.02, 7), step = 0.02)
p <- tda_pl_function(p, ln(0.7), range = c(0.02, 7), step = 0.02, lty = 5)
p <- tda_pl_function(p, ln(1.3), range = c(0.02, 7), step = 0.02, lty = 6)

p <- tda_pl_text(p, "b = 1.3", at = c(6, 0.125))
p <- tda_pl_text(p, "b = 1.0", at = c(6, 0.4))
p <- tda_pl_text(p, "b = 0.7", at = c(6, 0.725))

plot(p)
```

### 6.17.3.7 Generalized Gamma Models

```{r gengamma, box = "Generalized gamma (rt13.cf)"}
tda_rate(Surv(TFP, DES) ~ COHO2 + COHO3 + W, data = d, model = 13)
```

The generalised gamma rate at three values of the shape parameter,
written out with TDA's `lgam()` and `icg()`, the log gamma and the
incomplete gamma function.

```{r plot_g1, box = "plot-g1.cf", fig.width = 3.54, fig.height = 2.27, out.width = "340px", class.source = "folded"}
gg <- function(k) sprintf(
    "%s^(%s-0.5)*exp(sqrt(%s)*log(x1)-%s*exp(log(x1)/sqrt(%s)))/(x1*exp(lgam(%s))*(1-icg(%s*exp(log(x1)/sqrt(%s)),%s)))",
    k, k, k, k, k, k, k, k, k)

p <- tda_ps(xlim = c(0, 5), ylim = c(0, 2), width = 90, height = 50)
p <- tda_pl_axes(p, sc = c(1, 1))
p <- tda_pl_frame(p)
p <- tda_pl_function(p, gg(1), range = c(0.01, 5), step = 0.02)
p <- tda_pl_function(p, gg(0.5), range = c(0.01, 5), step = 0.02, lty = 5)
p <- tda_pl_function(p, gg(10), range = c(0.01, 5), step = 0.02, lty = 6)

p <- tda_pl_text(p, "a = 0.0", at = c(0.5, 1.73))
p <- tda_pl_text(p, "b = 1.0", at = c(0.5, 1.57))
p <- tda_pl_text(p, "k = 10.0", at = c(4, 0.63))
p <- tda_pl_text(p, "k = 1.0", at = c(4, 1.07))
p <- tda_pl_text(p, "k = 0.5", at = c(4, 1.5))

plot(p)
```

Generalised gamma transition rates at three scale parameters.
```{r plot_g2, box = "plot-g2.cf", fig.width = 3.54, fig.height = 2.27, out.width = "340px", class.source = "folded"}
gg2 <- function(b) sprintf(
    "1^(1-0.5)*exp(sqrt(1)*log(x1)/%s-1*exp((log(x1)/%s)/sqrt(1)))/(x1*exp(lgam(1))*(1-icg(1*exp((log(x1)/%s)/sqrt(1)),1)))",
    b, b, b)

p <- tda_ps(xlim = c(0, 5), ylim = c(0, 3), width = 90, height = 50)
p <- tda_pl_axes(p, sc = c(1, 1))
p <- tda_pl_frame(p)
p <- tda_pl_function(p, gg2(1), range = c(0.01, 5), step = 0.02)
p <- tda_pl_function(p, gg2(0.7), range = c(0.01, 5), step = 0.02, lty = 5)
p <- tda_pl_function(p, gg2(2), range = c(0.01, 5), step = 0.02, lty = 6)

p <- tda_pl_text(p, "a = 0.0", at = c(0.5, 2.6))
p <- tda_pl_text(p, "k = 1.0", at = c(0.5, 2.35))
p <- tda_pl_text(p, "b = 2.0", at = c(4, 0.6))
p <- tda_pl_text(p, "b = 1.0", at = c(4, 1.1))
p <- tda_pl_text(p, "b = 0.7", at = c(4, 2))

plot(p)
```

### 6.17.3.8 Inverse Gaussian Models

```{r invgauss, box = "Inverse Gaussian (rt14.cf)"}
tda_rate(Surv(TFP, DES) ~ COHO2 + COHO3 + W, data = d, model = 14)
```

The inverse Gaussian rate, written out with TDA's `ndf()` and `nd()`.

```{r plot_ig, box = "plot-ig.cf", fig.width = 3.54, fig.height = 2.27, out.width = "340px", class.source = "folded"}
ig <- function(a) sprintf(
    "(ndf((%s*x1-1)/sqrt(x1))/x1^(3/2))/(nd((1-%s*x1)/sqrt(x1))-exp(2*%s)*nd((-1-%s*x1)/sqrt(x1)))",
    a, a, a, a)

p <- tda_ps(xlim = c(0, 6), ylim = c(0, 2.5), width = 90, height = 50)
p <- tda_pl_axes(p, sc = c(1, 0.5))
p <- tda_pl_frame(p)
p <- tda_pl_function(p, ig(1), range = c(0.02, 6), step = 0.02)
p <- tda_pl_function(p, ig(0.1), range = c(0.02, 6), step = 0.02, lty = 5)
p <- tda_pl_function(p, ig(1.5), range = c(0.02, 6), step = 0.02, lty = 6)

p <- tda_pl_text(p, "a = 0.1", at = c(5, 0.21))
p <- tda_pl_text(p, "a = 1.0", at = c(5, 0.83))
p <- tda_pl_text(p, "a = 1.5", at = c(5, 1.46))
p <- tda_pl_text(p, "b = 1.0", at = c(5, 2.08))

plot(p)
```

### 6.17.5.1 The frml Command

`tda_frml()` is the rate-model counterpart of `tda_fml()`: the episode is
named by the formula, and the likelihood is written out. `tf`, `ts` and
`des` are available inside the block, which is what makes a model with
its duration dependence writable.

**Example 1** The exponential model again, now by hand.

```{r frml, box = "An exponential model as frml (fml1.cf)"}
tda_frml(Surv(TFP, DES) ~ COHO2 + COHO3 + W, {
    rate = exp(a0 + COHO2 * a1 + COHO3 * a2 + W * a3)
    l1 = ifelse(DES, log(rate), 0)
    fn = l1 - rate * TFP
}, data = d, start = c(-4, 0, 0, 0))
```

**Example 2** `frml1m.cf`: three competing destinations, each with its
own rate, the episode contributing the log of whichever rate it
realised and the sum of all three to the survivor term. Its `DES` is
`rt1m.cf`'s (the `ge` coding).

```{r frml3, box = "Three competing rates (frml1m.cf)"}
tda_frml(Surv(TFP, DES) ~ COHO2 + COHO3 + W, {
    rate1 = exp(a10 + COHO2 * a11 + COHO3 * a12 + W * a13)
    rate2 = exp(a20 + COHO2 * a21 + COHO3 * a22 + W * a23)
    rate3 = exp(a30 + COHO2 * a31 + COHO3 * a32 + W * a33)
    rate = rate1 + rate2 + rate3
    l1 = ifelse(DES == 1, log(rate1),
         ifelse(DES == 2, log(rate2),
         ifelse(DES == 3, log(rate3), 0)))
    fn = l1 - rate * TFP
}, data = d4m, start = rep(c(-4, 0, 0, 0), 3),
   control = tda_control(maxit = 100))
```

**Example 2a** `frml2.cf` writes the piecewise constant model of
`rt2.cf` out by hand: one indicator per period, and the log rate as the
sum of the period constants and the covariate effects; `frml2c.cf` adds
the constraints that equate the periods. Both reach `rt2` and `rt2c`.

```{r frml2, box = "frml2.cf, frml2c.cf: the piecewise constant model, by hand"}
pw <- function(...) tda_frml(Surv(TFP, DES) ~ COHO2 + COHO3 + W, {
    bb = COHO2 * b1 + COHO3 * b2 + W * b3
    p1 = ge(tf, 0) & lt(tf, 12);  p2 = ge(tf, 12) & lt(tf, 24)
    p3 = ge(tf, 24) & lt(tf, 36); p4 = ge(tf, 36) & lt(tf, 48)
    p5 = ge(tf, 48) & lt(tf, 60); p6 = ge(tf, 60) & lt(tf, 72)
    p7 = ge(tf, 72) & lt(tf, 84); p8 = ge(tf, 84) & lt(tf, 96)
    p9 = ge(tf, 96)
    rr = ifelse(p1, a1, ifelse(p2, a2, ifelse(p3, a3, ifelse(p4, a4,
         ifelse(p5, a5, ifelse(p6, a6, ifelse(p7, a7, ifelse(p8, a8,
         ifelse(p9, a9, 0)))))))))
    s1 = 12 * exp(a1 + bb); s2 = s1 + 12 * exp(a2 + bb)
    s3 = s2 + 12 * exp(a3 + bb); s4 = s3 + 12 * exp(a4 + bb)
    s5 = s4 + 12 * exp(a5 + bb); s6 = s5 + 12 * exp(a6 + bb)
    s7 = s6 + 12 * exp(a7 + bb); s8 = s7 + 12 * exp(a8 + bb)
    lsurv = ifelse(p1, tf * exp(a1 + bb),
            ifelse(p2, s1 + (tf - 12) * exp(a2 + bb),
            ifelse(p3, s2 + (tf - 24) * exp(a3 + bb),
            ifelse(p4, s3 + (tf - 36) * exp(a4 + bb),
            ifelse(p5, s4 + (tf - 48) * exp(a5 + bb),
            ifelse(p6, s5 + (tf - 60) * exp(a6 + bb),
            ifelse(p7, s6 + (tf - 72) * exp(a7 + bb),
            ifelse(p8, s7 + (tf - 84) * exp(a8 + bb),
            ifelse(p9, s8 + (tf - 96) * exp(a9 + bb), 0)))))))))
    l1 = ifelse(DES, rr + bb, 0)
    fn = l1 - lsurv
}, data = d, start = c(rep(-4, 9), 0, 0, 0), ...)
pw()                                              # frml2.cf
pw(constraints = sprintf("b%d - b%d = 0", 2:9, 1:8))   # frml2c.cf
```

**Example 3** A Weibull and a log-normal, written out. The second uses
`nd` and `ndf`, TDA's normal distribution and density operators.

```{r frmlwei, box = "Weibull and log-normal, written out"}
tda_frml(Surv(TFP, DES) ~ COHO2 + COHO3 + W, {
    aa = exp(a0 + COHO2 * a1 + COHO3 * a2 + W * a3)
    bb = exp(b0)
    l1 = ifelse(DES, log(bb * aa^bb * tf^(bb - 1)), 0)
    fn = l1 - (aa * tf)^bb
}, data = d, start = c(-4, 0, 0, 0, 0), control = tda_control(maxit = 100))

tda_frml(Surv(TFP, DES) ~ COHO2 + COHO3 + W, {
    aa = a0 + COHO2 * a1 + COHO3 * a2 + W * a3
    bb = exp(b0)
    zz = (log(tf) - aa) / bb
    qz = 1 - nd(zz)
    vz = ndf(zz) / qz
    l1 = ifelse(DES, log(vz / (bb * tf)), 0)
    fn = l1 + log(qz)
}, data = d, start = c(4, 0, 0, 0, 0), control = tda_control(maxit = 100))
```

**Example 4** `frml5.cf` is the generalized gamma model written out,
with `icg`, the incomplete gamma integral, and `lgam`; with the shape
`kk = 1` it is the Weibull of Example 3, which is what the box shows.

```{r frml5, box = "frml5.cf: the generalized gamma, written out"}
tda_frml(Surv(TFP, DES) ~ COHO2 + COHO3 + W, {
    kk = 1
    aa = a0 + COHO2 * a1 + COHO3 * a2 + W * a3
    bb = exp(b0)
    zz = (log(tf) - aa) / bb
    qt = kk * exp(zz / sqrt(kk))
    rr = (kk - 0.5) * log(kk) + (sqrt(kk) * zz - qt) - (b0 + log(tf) + lgam(kk))
    icc = icg(qt, kk)
    fn = ifelse(DES, rr, log(1 - icc))
}, data = d, start = c(5, 0, 0, 0, 0), control = tda_control(algorithm = 4))
```

### 6.17.5.2 The Hernes Model

```{r hernes, box = "The Hernes model (frml8.cf)"}
tda_frml(Surv(TFP, DES) ~ COHO2 + COHO3 + W, {
    sigma = exp(s)
    lambda = exp(l) / (1 + exp(l))
    beta = exp(b)
    ff = sigma * exp(beta * lambda^tf)
    ff1 = ff + 1
    dens = -(ff * beta * lambda^tf * log(lambda)) / (ff1 * ff1)
    surv = 1 - 1 / ff1
    fn = ifelse(des, log(dens), log(surv))
}, data = d, control = tda_control(maxit = 100, algorithm = 4))
```

### 6.17.5.3 The Coale-McNeil Model

The manual's `frml9.cf` gives two starting values for three parameters
and lets TDA pad the third with zero, which TDA warns about; the same
values, spelled out, reach the same estimates without the warning.

```{r coale, box = "The Coale-McNeil model (frml9.cf)"}
tda_frml(Surv(TFP, DES) ~ COHO2 + COHO3 + W, {
    bg = 0.604
    sigma = exp(s) / (1 + exp(s))
    lambda = exp(l)
    mue = exp(m)
    beta = bg * lambda
    tl = exp(-lambda * (tf - mue))
    ldens = log(sigma) + l - beta * (tf - mue) - tl - lgam(bg)
    surv = 1 - sigma * (1 - icg(tl, bg))
    fn = ifelse(DES, ldens, log(surv))
}, data = d, start = c(l = -3, m = 2, s = 0),
   control = tda_control(maxit = 100, algorithm = 4))
```

### 6.17.6.2 Logistic Regression Models

Model 20: the discrete-time logistic model, the rate at each time point a
logit in the covariates.

```{r discrete, box = "Discrete time logistic"}
tda_rate(Surv(TFP, DES) ~ COHO2 + COHO3 + W, data = d, model = 20)
```

### 6.17.6.3 Complementary Log-Log Models

Model 21.

```{r cloglog, box = "Complementary log-log"}
tda_rate(Surv(TFP, DES) ~ COHO2 + COHO3 + W, data = d, model = 21)
```

### 6.17.7.1 Partial Likelihood Estimation

Model 1 is the Cox model, estimated by partial likelihood, here with one
transition and then with three.

```{r cox, box = "pl1.cf and pl1m.cf: Cox models with one and three transitions"}
tda_rate(Surv(TFP, DES) ~ COHO2 + COHO3 + W, data = d, model = 1)
tda_rate(Surv(TFP, DES) ~ COHO2 + COHO3 + W, data = d4m, model = 1)
```

### 6.17.7.2 Time-dependent Covariates

`define=` writes a covariate as a TDA expression in `time`, evaluated at
every risk set rather than once per episode, and `helpers=` names the
columns that expression needs. `MDATE` does not depend on `time` and so is
computed in R; `MARR` does, and so is defined for TDA.

```{r coxtv, box = "A time-dependent covariate"}
d$MDATE <- ifelse(d$TMAR <= 0, 10000, d$TMAR - d$TStart)
tda_rate(Surv(TFP, DES) ~ COHO2 + COHO3 + W + MARR, data = d, model = 1,
         define = list(MARR = "gt(time,MDATE)"), helpers = "MDATE")
```

### 6.17.7.3 Episode Splitting

`pl4.cf` splits every episode at the date of marriage (`split = MDATE`
in `edef`) and writes the pieces out; `pl5.cf` reads them back and fits
the Cox model twice, without and with `MARR`, the piece's marriage
status. `tda_split(at=)` is the split, and the pieces are left
truncated, so the model is told `Surv(ts, tf, des)`.

```{r pl45, box = "pl4.cf, pl5.cf: episode splitting"}
d$MDATE <- ifelse(d$TMAR <= 0, 10000, d$TMAR - d$TStart)
sp <- tda_split(Surv(TFP, DES) ~ COHO2 + COHO3 + W, d, at = "MDATE")
sp$MARR <- as.integer(sp$MDATE <= sp$ts)
nrow(sp)
tda_rate(Surv(ts, tf, des) ~ COHO2 + COHO3 + W, sp, model = 1)
tda_rate(Surv(ts, tf, des) ~ COHO2 + COHO3 + W + MARR, sp, model = 1)
```

### 6.17.7.4 The Proportionality Assumption

The usual test is an interaction with log time: if the effect is
proportional, its coefficient is zero.

```{r coxprop, box = "A test of proportionality (pl6.cf)"}
tda_rate(Surv(TFP, DES) ~ COHO2 + COHO3 + W + WTEST, data = d, model = 1,
         define = list(WTEST = "W * (log(time) - 4.22)"))
```

The same question asked by splitting the time axis: `tp=` estimates
within periods.

```{r coxtp, box = "The effect within periods (pl7.cf)"}
rrdat5 <- read.table(ex("rrdat.5"), col.names = c("TF", "DES", "G"))
rrdat5$G1 <- as.integer(rrdat5$G == 1)
pl7 <- tda_rate(Surv(TF, DES) ~ G1, data = rrdat5, model = 1,
                tp = c(0, 170.1, 354.1, 535.1))
pl7
pl7$gof
```

`$gof` holds the global goodness-of-fit statistic that comes with a
period split -- the manual's Box 5, from `pl8.cf` for the single
transition and `pl8m.cf` for the three.

```{r coxgof, box = "pl8.cf, pl8m.cf: goodness of fit over periods"}
tda_rate(Surv(TFP, DES) ~ COHO2 + COHO3 + W, data = d, model = 1,
         tp = seq(0, 96, 12))$gof
tda_rate(Surv(TFP, DES) ~ COHO2 + COHO3 + W, data = d4m, model = 1,
         tp = seq(0, 96, 12))$gof
```

### 6.17.7.5 Stratification

`strata()` on the right hand side is survival's spelling and TDA's `grp=`:
a separate baseline rate per stratum, one set of coefficients.

```{r coxstrata, box = "A stratified Cox model (pl9.cf)"}
tda_rate(Surv(TFP, DES) ~ COHO2 + COHO3 + strata(SEX), data = d, model = 1)
```

### 6.17.7.6 The Baseline Rate

`prate=` evaluates the baseline rate at a covariate vector.

```{r coxbase, box = "The baseline rate at given covariate values (pl1r.cf)"}
fit <- tda_rate(Surv(TFP, DES) ~ COHO2 + COHO3 + W, data = d, model = 1,
                prate = list(COHO3 = 1, W = 1))
head(fit$rates)
```

### 6.18.2 Investigating Events

Event sequence data are given as a list of column groups, one per
sequence data structure -- the manual's `seq8.cf` defines two from
`seq.d4`, `Y0..Y5` and `S0..S5`. `tda_seqev()` counts the events (Box
3), and `tda_seqevd()` their distribution over time for the first
structure (Box 4).

```{r seqev, box = "seq_info, seqev and seqevd"}
seqd4 <- read.table(ex("seq.d4"))
names(seqd4) <- c("ID", paste0("Y", 0:5), paste0("S", 0:5), "V1", "V2")
multi <- list(paste0("Y", 0:5), paste0("S", 0:5))

tda_seq_info(multi, data = seqd4)
tda_seqev(multi, data = seqd4)
tda_seqevd(multi, data = seqd4)
```

### 6.18.3 Data Generation

`tda_seqmd()` turns sequences into episode data around chosen events,
which is how a sequence becomes something a rate model can take.

```{r seqmd, box = "Episodes derived from sequences"}
tda_seqmd(paste0("Y", 0:5), event = c(1, 2), data = seqd4, summary = TRUE)
tda_seqmd(paste0("Y", 0:5), event = c(1, 2), data = seqd4)
tda_seqmd(paste0("Y", 0:5), event = c(1, 2), data = seqd4,
           covariates = c("V1", "V2"))
```

`xe=` adds time-varying covariates derived from a date: `xe=[V1],[V2]`
(the manual's `seq8c.cf`, Box 4) gives, for each episode, the process
time since the date each variable holds -- `event_covariates=`, named
after the columns TDA writes.

```{r seqmd2, box = "seqmd with xe= (seq8c.cf)"}
tda_seqmd(paste0("Y", 0:5), event = c(1, 2), data = seqd4,
           covariates = c("V1", "V2"),
           event_covariates = list(V1_D = "V1", V2_D = "V2"))
```

A third form of `xe=` counts the events of a given type in another
sequence structure before the present time and the time since the
last one, with effect-shape dummies (`seq8d.cf`, the manual's Box 5):
`[1,2,1,4]` is the event 1 -> 2 in structure 1, with 4 dummies.

```{r seqmd3, box = "seqmd with event counting (seq8d.cf)"}
tda_seqmd(paste0("Y", 0:5), event = c(1, 2), data = seqd4,
           covariates = c("V1", "V2"),
           event_covariates = list(V1_D = "V1", V2_D = "V2"),
           options = list(xe = "[V1],[V2],[1,2,1,4]"))
```

### 6.19.1 Contingency Tables

```{r loglin, box = "loglin on a contingency table"}
ll1 <- read.table(ex("ll1.dat"), col.names = c("X1", "X2", "H"))
tda_loglin(~ X1 + X2, data = ll1, weights = "H")$table
```

### 6.19.2 Loglinear Models

`ll2.dat` carries a fourth column, `S`, the scale factors of 6.19.1's
structural zeros; `ll2.cf` defines it and does not use it, and the
residual table's `Scale` column is all ones accordingly. `scale=` would
pass it.

```{r loglin2, box = "A loglinear model with residuals"}
ll2 <- read.table(ex("ll2.dat"), col.names = c("X1", "X2", "F", "S"))
fit <- tda_loglin(~ X1 + X2, data = ll2, weights = "F", residuals = TRUE)
fit
coef(fit)
head(residuals(fit))
```

## 7 Relational Data

The manual's example graph is `gd1.dat`: an edge list `I, J` with two
value variables, `V1` and `V2`, so every node count comes once per value
variable. `tda_graph()` is the `gdd` command.

### 7.2.1.1 Degree of Nodes

`gni` reports each node of the current graph: its number, its label, and
per value variable the in-degree, out-degree and number of loops;
`tda_g_degrees()` is `gni`, and `gd4.cf` runs it on `gd1.dat`. A loop
(node 8 to itself) is counted in its own column and not in the degrees:
the manual's Box 1 prints node 8's first-graph degrees as 1 and 1,
while `gd4.cf` run through TDA gives 0 and 0, as does a build of
Rohwer's unmodified sources, so the box is out of date.

```{r g1, box = "gni: in-degree, out-degree and loops per value variable (gd4.cf)"}
gd1 <- read.table(ex("gd1.dat"), col.names = c("I", "J", "V1", "V2"))
g1 <- tda_graph(gd1, directed = TRUE)
tda_g_degrees(g1)
```

### 7.2.1.2 Direct Links

`tda_g_links()` is `gdln`: for each node the nodes it links to, one
ragged record per node with any forward links.

```{r g2, box = "gdln: forward links"}
tda_g_links(g1)$table                     # forward links
tda_g_links(g1, backward = TRUE)$table    # and backward
```

### 7.2.2.1 Topological Sorting

`tda_g_toposort()` is `gsort`, on the manual's `gd6.dat`: a directed
acyclic graph, its nodes in an order that respects every edge.

```{r g3, box = "gsort (gd22.cf)"}
gd6 <- read.table(ex("gd6.dat"), col.names = c("I", "J", "V"))
gsrt <- tda_graph(gd6, directed = TRUE)
tda_g_toposort(gsrt)$table

# option 2 relabels the graph's edges, keeping their values
tda_g_toposort(gsrt, relabel_graph = TRUE)$table
```

### 7.2.3.1 Connected Components

`tda_g_components()` is `gcon`. `gd7.cf` declares `gd1.dat` undirected
and valued (`gt = 2`) and asks for the components of its second graph,
the one `V2` defines (`gn = 2`): node 8 joins component 1 through the
edge 8-7 that `V2` carries and `V1` does not.

```{r g4, box = "gcon: components of graph 2 (gd7.cf)"}
gcomp <- tda_graph(gd1, directed = FALSE)
tda_g_components(gcomp, gn = 2)$table

# option 2 lists each component's edges instead, with their values
tda_g_components(gcomp, gn = 2, edges = TRUE)$table
```

### 7.2.3.2 Reachable Nodes in Directed Graphs

`gdcon` (`tda_g(g, "gdcon")`) is the directed counterpart of `gcon`:
strongly connected components (`opt = 1`) or, with `opt = 2`, the
nodes each node reaches. `gd8.cf` runs it on graph 2 of `gd1.dat`.

```{r g9, box = "gdcon: reachable nodes (gd8.cf)"}
g1d <- tda_graph(gd1, directed = TRUE)
tda_g(g1d, "gdcon", list(gn = 2, opt = 2))$table
```

### 7.2.3.3 Cut Nodes and Blocks

`tda_g_cutpoints()` is `gcut`, on `gd8.dat`: the nodes whose removal
disconnects the graph, and with `opt = 2` the blocks.

```{r g10, box = "gcut (gd26.cf)"}
gd8 <- read.table(ex("gd8.dat"), col.names = c("I", "J", "V"))
g8 <- tda_graph(gd8, directed = FALSE)
tda_g_cutpoints(g8)$table
tda_g_cutpoints(g8, opt = 2)$table
```

### 7.2.4.1 Enumeration of Paths

`tda_g_paths()` is `gep`, on `gd5.dat`, a small directed graph;
`gd20.cf` runs its six output options: the paths between all pairs of
nodes, their number per pair, the shortest and longest path per pair,
and the two full listings with each path's length and value.

```{r g11, box = "gep (gd20.cf)"}
gd5 <- read.table(ex("gd5.dat"), col.names = c("I", "J", "V"))
g5 <- tda_graph(gd5, directed = TRUE)
for (o in 1:6)
    print(tda_g_paths(g5, opt = o)$table)
```

The graph itself, as the manual draws it (`gd18.cf`): five nodes,
curved arrows for the edges that run both ways, the edge values as
labels.

```{r plg18, box = "The directed graph of gd5.dat, drawn (gd18.cf)", fig.width = 3.15, fig.height = 1.87, out.width = "302px", class.source = "folded"}
p <- tda_ps(xlim = c(0, 8), ylim = c(-1, 5), width = 80, height = 40)
nodes <- data.frame(id = 1:5, x = c(3, 5, 1, 7, 7), y = c(3, 1, 1, 4, 1),
                    grey = 0.8)
edges <- data.frame(from  = c(1, 2, 3, 3, 2, 4, 5, 2, 3),
                    to    = c(2, 3, 2, 1, 4, 5, 3, 5, 4),
                    curve = c(-3, 0, -3, -3, 0, 0, -3, 0, -3),
                    label = c(1, 2, 0.5, 6, 3, 4, 5, 0.5, 7),
                    arrow = "1.5,1.0")
plot(tda_pl_graph(p, nodes, edges))
```

### 7.2.4.2 All Shortest Paths

`tda_g_shortest()` is `gsp`, the same graph, `gd27.cf`.

```{r g12, box = "gsp (gd27.cf)"}
tda_g_shortest(g5, opt = 1, fmt = "4.1")$table
```

### 7.2.4.3 Transitive Closure

`tda_g_transitive()` is `gtcl`; `gd16.cf` runs it on `gd5.dat` read
as an undirected, an undirected valued and a directed graph, `opt = 2`
for the closure as a matrix.

```{r g13, box = "gtcl (gd16.cf)"}
# gdd's gt= is the graph type: 1 undirected unvalued, 2 undirected
# valued, 3 directed unvalued, 4 directed valued. The manual's box runs
# gtcl over all four.
for (gt in 1:4)
    print(tda_g_transitive(tda_graph(gd5, options = list(gt = gt)),
                           opt = 2, fmt = "4.1")$table)
```

### 7.2.5.1 Depth-first Spanning Trees

`tda_g_spanning()` is `gst`, on `gd7.dat`: one spanning tree per
component, the edges listed with their values.

```{r g5, box = "gst (gd24.cf)"}
gd7 <- read.table(ex("gd7.dat"), col.names = c("I", "J", "V"))
tda_g_spanning(tda_graph(gd7, directed = FALSE))$table
```

The manual's Figure 7.2.5.1-1 is `gd7.dat`'s first component drawn
by `gd21.cf`, with its spanning tree's edges.

```{r plg21, box = "gd7.dat drawn (gd21.cf)", fig.width = 2.36, fig.height = 1.48, out.width = "227px", class.source = "folded"}
p <- tda_ps(xlim = c(0, 6), ylim = c(0, 5), width = 60, height = 30)
nodes <- data.frame(id = c(5, 7, 4, 1, 2), x = c(1, 3, 5, 2, 4),
                    y = c(3, 3, 3, 1, 1), grey = 0.8)
edges <- data.frame(from  = c(5, 4, 5, 7, 7, 4, 1, 4),
                    to    = c(7, 7, 1, 1, 2, 2, 2, 5),
                    curve = c(0, 0, 0, 0, 0, 0, 0, 3),
                    label = c(1, 5, 3, 2, 4, 7, 8, 6),
                    arrow = "1.5,1.0")
plot(tda_pl_graph(p, nodes, edges))
```

```{r gd23, box = "gd23.cf", fig.width = 2.36, fig.height = 1.48, out.width = "227px", class.source = "folded"}
p <- tda_ps(xlim = c(0, 7), ylim = c(0, 3.7), width = 60, height = 30)
nodes <- data.frame(id = 1:8, x = c(0.5, 2, 2, 3.5, 4.5, 5, 6, 6),
                    y = c(1.5, 2.5, 0.5, 1.5, 0.5, 1.5, 0.5, 2.5), grey = 0.8)
edges <- data.frame(
    from  = c(1, 2, 4, 5, 3, 1, 2, 6, 6, 8),
    to    = c(2, 4, 5, 3, 4, 3, 3, 8, 7, 7),
    label = 1:10,
    arrow = "1.5,1.0")
p <- tda_pl_graph(p, nodes, edges)

plot(p)
```

### 7.2.5.2 Minimum Spanning Trees

`tda_g_mst()` is `gmst`, `gd25.cf`, on the same `gd7.dat`.

```{r g14, box = "gmst (gd25.cf)"}
tda_g_mst(tda_graph(gd7, directed = FALSE))$table
```

### 7.2.5.3 Enumeration of Spanning Trees

`tda_g_spantrees()` is `gnst`, on `gd10.dat` (`gd32.cf`): the number
of spanning trees, and with `opt = 2` the trees themselves.

```{r g15, box = "gnst (gd32.cf)"}
gd10 <- read.table(ex("gd10.dat"), col.names = c("I", "J", "V"))
g10 <- tda_graph(gd10, directed = FALSE)
tda_g_spantrees(g10, opt = 1)$table
tda_g_spantrees(g10, opt = 2)$table
```

```{r gd31, box = "gd31.cf", fig.width = 3.15, fig.height = 1.87, out.width = "302px", class.source = "folded"}
p <- tda_ps(xlim = c(0, 8), ylim = c(-1, 5), width = 80, height = 40)
nodes <- data.frame(id = 1:4, x = c(3, 5, 3, 5), y = c(3, 3, 1, 1), grey = 0.8)
edges <- data.frame(from = c(1, 1, 1, 2, 3), to = c(2, 3, 4, 3, 4))
p <- tda_pl_graph(p, nodes, edges)

plot(p)
```

### 7.2.6.1 Fundamental Set of Cycles

`tda_g_cycles()` is `gcyc`; `gd30.cf` runs it on `gd9.dat`, an
undirected graph with 6 nodes and 11 edges. `opt = 1` lists a
fundamental set of cycles, one ragged record each: index, cycle number,
length, and the nodes.

`gcyc`'s four options are four views of the same cycle space. Option 1
lists the nodes of each fundamental cycle, so its rows are ragged;
option 2 gives the edge and one indicator per fundamental cycle; option 3
the same over every cycle found; option 4 writes the edges first and
then one indicator row per cycle, two row shapes in the one file, which
is why it arrives ragged as well.

```{r g6, box = "gcyc, options 1 to 4"}
gd9 <- read.table(ex("gd9.dat"), col.names = c("I", "J", "V"))
g9 <- tda_graph(gd9, directed = FALSE)
tda_g_cycles(g9, "fundamental_nodes")$table
tda_g_cycles(g9, "fundamental_edges")$table
tda_g_cycles(g9, "all_v1")$table

# option 4's records are of two shapes -- a cycle's row and its edge
# list -- so they stay a ragged list; the manual prints all of them
print(tda_g_cycles(g9, "all_v2")$table, n = Inf)
```

```{r gd29, box = "gd29.cf", fig.width = 3.15, fig.height = 1.87, out.width = "302px", class.source = "folded"}
p <- tda_ps(xlim = c(0, 8), ylim = c(-1, 5), width = 80, height = 40)
nodes <- data.frame(id = c(1, 3, 4, 6, 2, 5), x = c(3, 3, 5, 1, 5, 7),
                    y = c(3, 1, 1, 1, 3, 1), grey = 0.8)
edges <- data.frame(
    from  = c(1, 3, 1, 1, 3, 3, 2, 2, 4, 6, 6),
    to    = c(6, 6, 3, 2, 4, 2, 4, 5, 5, 4, 5),
    curve = c(0, 0, 0, 0, 0, 0, 0, 0, 0, 2, 5))
p <- tda_pl_graph(p, nodes, edges)

plot(p)
```

### 7.2.6.2 Enumeration of Cycles in Directed Graphs

`tda_g_dcycles()` is `gdcyc`, on `gd5.dat` (`gd33.cf`).

```{r g16, box = "gdcyc (gd33.cf)"}
tda_g_dcycles(g5, opt = 1)$table
tda_g_dcycles(g5, opt = 2)$table
```

### 7.2.8.1 Maximal Flows

`tda_g_flow()` is `gflow`, on `gd11.dat`: for every ordered pair of
nodes the maximal flow, one record per pair -- index, the two nodes'
indices, the two nodes, the flow.

```{r g7, box = "gflow (gd35.cf)"}
gd11 <- read.table(ex("gd11.dat"), col.names = c("I", "J", "V"))
g11 <- tda_graph(gd11, directed = TRUE)
tda_g_flow(g11)$table

# option 2 gives the same flows as a matrix
tda_g_flow(g11, matrix = TRUE)$table
```

```{r gd34, box = "gd34.cf", fig.width = 3.15, fig.height = 1.87, out.width = "302px", class.source = "folded"}
p <- tda_ps(xlim = c(0, 11), ylim = c(-1, 7), width = 80, height = 40)
nodes <- data.frame(id = c(4, 1, 8, 6, 3, 7, 2, 5),
                    x = c(3, 3, 1, 8, 8, 10, 6, 4.5),
                    y = c(1, 5, 3, 1, 5, 3, 5, 3), grey = 0.8)
edges <- data.frame(
    from  = c(8, 8, 1, 6, 2, 6, 3, 2, 4, 5),
    to    = c(1, 4, 2, 4, 3, 7, 7, 6, 5, 3),
    label = c(40, 20, 30, 17, 10, 32, 22, 37, 54, 50),
    arrow = "1.5,1.0")
p <- tda_pl_graph(p, nodes, edges)

plot(p)
```

### 7.2.9.1 Cliques

`tda_g_cliques()` is `gcliq`, on `gd16.dat`; the manual's Box 1 runs
both of its algorithms, `alg = 1` (Bron-Kerbosch) and `alg = 2`
(Harary-Ross), whose listings differ in order, not in content.

```{r g8, box = "gcliq (gd42.cf)"}
gd16 <- read.table(ex("gd16.dat"), col.names = c("I", "J", "V"))
g16 <- tda_graph(gd16, directed = FALSE)
tda_g_cliques(g16, algorithm = "bron_kerbosch")$table
tda_g_cliques(g16, algorithm = "harary_ross")$table
```

```{r gd43, box = "gd43.cf", fig.width = 3.15, fig.height = 1.87, out.width = "302px", class.source = "folded"}
p <- tda_ps(xlim = c(0, 8), ylim = c(-1.5, 7), width = 80, height = 40)
nodes <- data.frame(id = 1:6, x = c(1, 3, 5, 3, 7, 7), y = c(3, 6, 3, 0, 6, 0),
                    grey = 0.8)
edges <- data.frame(
    from = c(1, 1, 1, 2, 2, 2, 4, 4, 3, 3, 5),
    to   = c(3, 2, 4, 4, 3, 5, 3, 6, 5, 6, 6))
p <- tda_pl_graph(p, nodes, edges)

plot(p)
```

### 7.2.9.2 Compact Sets

`tda_g_independent()` is `giset`, independent sets on `gd12.dat`
(`gd37.cf`), and `tda_g_compact()` is `gcset`, compact sets on
`gd13.dat` (`gd39.cf`).

```{r g17, box = "giset and gcset"}
gd12 <- read.table(ex("gd12.dat"), col.names = c("I", "J", "V"))
tda_g_independent(tda_graph(gd12, directed = FALSE))$table
gd13 <- read.table(ex("gd13.dat"), col.names = c("I", "J", "V"))
g13 <- tda_graph(gd13, directed = FALSE)
print(tda_g_compact(g13, opt = 1)$table, n = Inf)   # the sets themselves
tda_g_compact(g13, opt = 2)$table                   # one row per node
```

```{r gd38, box = "gd38.cf", fig.width = 3.15, fig.height = 1.87, out.width = "302px", class.source = "folded"}
p <- tda_ps(xlim = c(0, 8), ylim = c(-1.5, 7), width = 80, height = 40)
nodes <- data.frame(id = 1:6, x = c(1, 3, 5, 3, 7, 7), y = c(3, 6, 3, 0, 6, 0),
                    grey = 0.8)
edges <- data.frame(
    from  = c(1, 1, 1, 2, 2, 4, 4, 3, 3, 5),
    to    = c(3, 2, 4, 3, 5, 3, 6, 5, 6, 6),
    label = c(1, 3, 12, 7, 5, 11, 2, 4, 15, 8))
p <- tda_pl_graph(p, nodes, edges)

plot(p)
```

### 7.2.9.3 Regular Equivalence

```{r gd36, box = "gd36.cf", fig.width = 3.15, fig.height = 1.87, out.width = "302px", class.source = "folded"}
p <- tda_ps(xlim = c(0, 8), ylim = c(-1.5, 7), width = 80, height = 40)
nodes <- data.frame(id = 1:6, x = c(1, 2.5, 4, 4, 5.5, 7),
                    y = c(1, 3, 1, 5, 3, 1), grey = 0.8)
edges <- data.frame(
    from  = c(1, 1, 2, 2, 2, 3, 3, 4, 5, 1),
    to    = c(2, 3, 3, 4, 5, 4, 5, 5, 6, 6),
    curve = c(0, 0, 0, 0, 0, 0, 0, 0, 0, 3))
p <- tda_pl_graph(p, nodes, edges)

plot(p)
```

### 7.3.1.1 The Linear Assignment Problem

`tda_gap()` is `gap`: given a cost matrix, the assignment of rows to
columns with minimal total cost. The manual's `co1.dat` is a 6 x 6
cost matrix, row by row; the result is the permutation of its Box 2
and the matrix with its rows permuted accordingly.

```{r gap, box = "gap (co1.cf)"}
cost <- matrix(scan(ex("co1.dat"), quiet = TRUE), 6, 6, byrow = TRUE)
ap <- tda_gap(cost)
ap$assignment
ap$permuted
```

### 7.3.1.2 The Quadratic Assignment Problem

`tda_gqap()` is `gqap`, on two symmetric matrices, a flow and a
distance matrix. `co2.dat` gives their lower triangles in one file, one
pair per line in the order (2,1), (3,1), (3,2); the manual's Box 2 is
the permutation 2, 1, 3.

```{r gqap, box = "gqap (co2.cf)"}
co2 <- read.table(ex("co2.dat"), col.names = c("F", "D"))
lower <- function(v) { m <- matrix(0, 3, 3); m[lower.tri(m)] <- v; m + t(m) }
qap <- tda_gqap(lower(co2$F), lower(co2$D))
qap$assignment
qap$cost

# gqap's second output file: the multigraph under that permutation, with
# the flow and distance between each pair
qap$permuted
```

### 7.5.1.1 SAHN Algorithms

`tda_cluster()` covers TDA's clustering commands. The manual's example
dissimilarity matrix is `cl1.dat`, the upper triangle of an 11 x 11
matrix, row by row; `cl1.cf` runs `hcls` with `opt = 2`, complete
linkage. The output is the manual's Box 3: at each level the two
clusters merged, named by their representative nodes.

```{r cl1, box = "hcls: complete-link clustering (cl1.cf)"}
D <- scan(ex("cl1.dat"), quiet = TRUE)
dm <- matrix(0, 11, 11)
dm[lower.tri(dm)] <- D
dm <- dm + t(dm)
tda_cluster(dm, method = "hierarchical_single", options = list(opt = 2))$table
```

`cl1a.cf` reads the dendrogram back and cuts it into four levels with
`hclsp`; `tda_cutree()` is both steps.

```{r cl1a, box = "hclsp: the dendrogram cut at four levels (cl1a.cf)"}
tda_cutree(dm, nlev = 4)$table
```

### 7.5.1.2 Nearest-Neighbor Clustering

`cl2.cf` runs `nncl` twice: with a threshold (`sc = 10`) and with a
fixed number of neighbours (`ns = 2`).

```{r cl2, box = "nncl (cl2.cf)"}
tda_cluster(dm, method = "nearest_neighbour", options = list(alg = 1, sc = 10))$table
tda_cluster(dm, method = "nearest_neighbour", options = list(alg = 2, ns = 2))$table
```

### 7.5.2.1 Maximally Different Cluster Centers

`cl3.cf` is `hcld` with its first algorithm: divisive clustering around
maximally different centres. The manual's Box 2 lists each cluster's
nodes.

```{r cl3, box = "hcld (cl3.cf)"}
hc <- tda_cluster(dm, method = "hierarchical", algorithm = "centers")
hc$clusters
hc$merges     # one row per split: the two clusters, their sizes, their diameters
```

### 7.5.2.2 Maximally Homogeneous Clusters

The second algorithm splits on the diameter instead -- at each step the
cluster with the largest diameter is divided.

```{r cl4, box = "hcld, second algorithm"}
# the default stops after two splits; the manual's run goes on until
# the size bound is reached
hd <- tda_cluster(dm, method = "hierarchical", algorithm = "diameter",
                  max_splits = 5)
hd$clusters
hd$merges
```

`cl3p.cf` draws the merge table as a dendrogram: columns 1 and 2 of
`cl3.df` are the pair joined, column 5 the distance.

```{r cl3p, box = "cl3p.cf", class.source = "folded"}
edges <- data.frame(
    from = c(2, 3, 4, 5, 6, 7, 8, 9, 10, 11),
    to   = c(1, 1, 2, 2, 3, 3, 5, 5, 6, 6),
    v    = c(27.5, 27.5, 17.0, 17.0, 16.5, 16.5, 12.0, 12.0, 4.0, 4.0))
g <- tda_graph(edges, directed = FALSE)

plot(g, width = 90, height = 50, layout = "tree", rt = 1, pl = 2, nc = 11)
```

### 7.5.3.1 The Bond Energy Algorithm

`becl` is the bond energy algorithm on a graph; `cl5.cf` runs it on
`gd15.dat`, five nodes, with the four output options. Option 1 is the
node permutation, option 2 the edge list permuted (Box 3).

```{r cl5, box = "becl (cl5.cf)"}
gd15 <- read.table(ex("gd15.dat"), col.names = c("I", "J", "V"))
g15 <- tda_graph(gd15, directed = FALSE)
tda_g(g15, "becl", list(opt = 1, fmt = "3.0"))$table
tda_g(g15, "becl", list(opt = 2, fmt = "3.0"))$table

# options 3 and 4 take sc=, the value to fill a non-edge with; the
# default is -1 and cl5.cf asks for 0
tda_g(g15, "becl", list(opt = 3, fmt = "3.0", sc = 0))$table
tda_g(g15, "becl", list(opt = 4, fmt = "3.0", sc = 0))$table
```

### 7.6.1.1 Direct and Indirect Control

`tda_gfcf()` is `gfcf`, control in a directed valued graph -- `gd3.dat`
is a small ownership network. With `sc = 49` (control at more than 49
per cent) node 1 controls nodes 2 to 9, node 9 controls 1 to 8, node
3 controls 5; the pair table of the manual's Box 2 is the result's
`"pairs"` attribute.

```{r gfcf, box = "gfcf (gd13.cf)"}
gd3 <- read.table(ex("gd3.dat"), col.names = c("I", "J", "V"))
ctl <- tda_gfcf(gd3, threshold = 49)
ctl
head(attr(ctl, "pairs"), 8)
```

`gbcf` is the same question backwards -- who controls a given node --
on `gd3a.dat`. It writes two files: the pair table, and a summary with
one row per node (the manual's Box 4 shows both).

```{r gbcf, box = "gbcf (gd14.cf)"}
gd3a <- read.table(ex("gd3a.dat"), col.names = c("I", "J", "V"))
bwd <- tda_g_backward(tda_graph(gd3a, directed = TRUE), sc = 49)
bwd$table
bwd$summary

# gd14a.cf asks for the second layout, opt = 2
tda_g_backward(tda_graph(gd3a, directed = TRUE), sc = 49,
               opt = 2, fmt = "3.0")$table
```

In `gd12.cf` node 1 is drawn as a square, `shape =` for `plg`'s
`gt=2`, and several edges are curved or run back to an earlier node.

```{r gd12, box = "gd12.cf", fig.width = 3.94, fig.height = 2.27, out.width = "378px", class.source = "folded"}
p <- tda_ps(xlim = c(0, 11), ylim = c(2.0, 7.5), width = 100, height = 50)

nodes <- data.frame(
    id = c(9, 1, 3, 2, 4, 10, 7, 6, 5, 8, 11),
    x  = c(3, 3, 5, 5, 7, 9, 3, 5, 7, 9, 7),
    y  = c(7, 5, 5, 6, 5, 5, 3, 3, 4, 3, 3),
    grey = 0.8,
    shape = c(1, 2, 1, 1, 1, 1, 1, 1, 1, 1, 1))

edges <- data.frame(
    from  = c(9, 1, 1, 1, 2, 10, 3, 1, 3, 3, 11, 7, 5, 8, 7, 3, 4),
    to    = c(1, 9, 2, 3, 4, 4, 5, 7, 7, 6, 6, 8, 8, 4, 6, 4, 9),
    label = c(50, 10, 50, 50, 30, 15, 50, 20, 40, 30, 22, 30, 30, 20, 30, 10, 40),
    curve = c(5, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 4, 0, 0, 0, 0, 10),
    arrow = "1.2,1.0")

p <- tda_pl_graph(p, nodes, edges)

plot(p)
```

### 7.6.1.2 Control

`gio` computes ownership shares in a directed valued graph: for each
pair, the direct share `A(i,j)` and the share including holdings held
through other nodes, `Y(i,j)`. `gd11.cf` runs it on `gd2b.dat`.

```{r gio, box = "gio (gd11.cf)"}
gd2b <- read.table(ex("gd2b.dat"), col.names = c("I", "J", "V"))
tda_g_ownership(tda_graph(gd2b, directed = TRUE))$table
```

Two graphs on one page; in the second a reciprocal pair of edges is
drawn as two curves, and all edges carry arrowheads.

```{r gd9, box = "gd9.cf", fig.width = 3.15, fig.height = 1.87, out.width = "302px", class.source = "folded"}
p <- tda_ps(xlim = c(0, 11.5), ylim = c(0, 4), width = 80, height = 40)

nodesA <- data.frame(id = 1:4, x = c(0.5, 3, 5, 5), y = c(2, 2, 1, 3),
                     grey = 0.8)
edgesA <- data.frame(from = c(1, 2, 2, 4), to = c(2, 4, 3, 3),
                     label = c(0.5, 0.3, 0.1, 0.2), arrow = "1.5,1.0")
p <- tda_pl_graph(p, nodesA, edgesA)

nodesB <- data.frame(id = 1:4, x = c(6.5, 9, 11, 11), y = c(2, 2, 1, 3),
                     grey = 0.8)
edgesB <- data.frame(from = c(1, 2, 2, 4, 3), to = c(2, 4, 3, 3, 2),
                     label = c(0.5, 0.3, 0.1, 0.2, 0.1),
                     curve = c(0, 0, 0, 0, -3), arrow = "1.5,1.0")
p <- tda_pl_graph(p, nodesB, edgesB)

p <- tda_pl_text(p, "Graph A", at = c(0.5, 3))
p <- tda_pl_text(p, "Graph B", at = c(6.5, 3))

plot(p)
```

### 7.6.1.3 Measures of Flow Centrality

`gfc` (`tda_g(g, "gfc")`) is flow centrality on `gd14.dat`
(`gd41.cf`), one record per pair of nodes.

```{r gfc, box = "gfc (gd41.cf)"}
gd14 <- read.table(ex("gd14.dat"), col.names = c("I", "J", "V"))
tda_g(tda_graph(gd14, directed = TRUE), "gfc", list(fmt = "4.0"))$table
```

The section's Figure 1 is the directed graph the examples use, drawn by
`gd40.cf`.

```{r gd40, box = "gd40.cf", fig.width = 3.15, fig.height = 1.87, out.width = "302px", class.source = "folded"}
p <- tda_ps(xlim = c(0, 8), ylim = c(-1.5, 7), width = 80, height = 40)
nodes <- data.frame(id = 1:5, x = c(1, 3, 5, 3, 7), y = c(3, 6, 3, 0, 3),
                    grey = 0.8)
edges <- data.frame(
    from  = c(1, 1, 1, 2, 4, 3),
    to    = c(2, 3, 4, 3, 3, 5),
    label = c(3, 1, 2, 3, 2, 2),
    arrow = "1.5,1.0")
p <- tda_pl_graph(p, nodes, edges)

plot(p)
```

## 8 Set-valued Data

### 8.3.2 Evaluation of Inclusion Functions

Chapter 8 is about set-valued data: a case is not a number but an
interval, and the answer is an interval too. `tda_evalfi()` evaluates a
function over interval arguments.

The manual's box is `evalfi1`: the function's inclusion over the
interval and its derivative's, `derivative = TRUE`.

```{r evalfi, box = "An inclusion function and its derivative (evalfi1)"}
tda_evalfi("sin(x)", x = c(0, 1))
tda_evalfi("sin(x)", x = c(0, 1), derivative = TRUE)
```

### 8.4.1 The gmin Command

`tda_gmin()` finds the global minimum by subdivision. `start=` gives the
search interval, as `c(step, lower, upper)`; `tol_fe=` is the tolerance
on the function value within which a box is finally accepted. TDA 6.4p's
default is 1e-10, and with it the box containing the minimum of sin(x)
on [0, 10] is only temporarily accepted (its lower bound is -1 to
sixteen digits, but not to ten); 1e-6 accepts it, giving one box with
x in [4.71237, 4.71245].

The manual's box reports 39 iterations, 41 function evaluations and
77 inclusion evaluations. The program does not: running the manual's
command file unchanged through TDA gives 19, 20 and 35, and so does the
call here. The manual is out of date on this box.

```{r gmin, box = "gmin"}
g <- tda_gmin("sin(x)", start = list(c(1, 0, 10)), use_derivatives = TRUE,
              tol_fe = 1e-6)
g$value
g$boxes
g$estimates

# the counters TDA prints above the boxes
g$counts
```

### 8.4.2 The range Command

```{r range, box = "range"}
tda_range("sin(x)", start = list(c(1, 0, 10)), use_derivatives = TRUE)
```

### 8.5.1 Set-valued Discrete Variables

`self_consistent = TRUE` asks for the self-consistent estimate rather
than the bounds alone.

```{r sddf, box = "Distribution of a set-valued discrete variable"}
# the manual's box shows both data files as well as the results
id1 <- read.table(ex("id1.dat"))
id1
tda_sddf(id1, self_consistent = TRUE)

id1a <- read.table(ex("id1a.dat"))
id1a
tda_iddf(~iv(V1, V2), data = id1a, self_consistent = TRUE)
```

### 8.5.2 Interval-valued Variables

`iv(lower, upper)` is the interval-valued variable, the counterpart of
`Surv()` for this chapter.

```{r idf, box = "Distribution of an interval-valued variable"}
id2 <- read.table(ex("id2.dat"))
id2$XL <- id2$V1 - 30L
id2$XH <- id2$V1 + 30L
tda_idf(~iv(XL, XH), data = id2)
```

### 8.6.1 Mean Values

The mean of an interval-valued variable is itself an interval: every
value in it is the mean of some admissible data set.

```{r imean, box = "imean"}
tda_imean(~iv(XL, XH), data = id2)
```

### 8.6.2 Variances

```{r ivariance, box = "ivariance"}
tda_ivariance(~iv(V1, V2), data = id1a)
```
